
\chapter{The \texttt{utils} package}
\inputencoding{latin1}
\HeaderA{utils-package}{The R Utils Package}{utils.Rdash.package}
\aliasA{utils}{utils-package}{utils}
\keyword{package}{utils-package}
\keyword{programming}{utils-package}
%
\begin{Description}\relax
R utility functions
\end{Description}
%
\begin{Details}\relax
This package contains a collection of utility functions.

For a complete
list, use \code{library(help="utils")}.

\end{Details}
%
\begin{Author}\relax
R Development Core Team and contributors worldwide

Maintainer: R Core Team \email{R-core@r-project.org}
\end{Author}
\inputencoding{latin1}
\HeaderA{alarm}{Alert the user}{alarm}
\keyword{utilities}{alarm}
%
\begin{Description}\relax
Gives an audible or visual signal to the user.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
alarm()
\end{verbatim}
\end{Usage}
%
\begin{Details}\relax
\code{alarm()} works by sending a \code{"\bsl{}a"} character to the console.
On most platforms this will ring a bell, beep, or give some other signal
to the user (unless standard output has been redirected).
\end{Details}
%
\begin{Value}
No useful value is returned.
\end{Value}
%
\begin{Examples}
\begin{ExampleCode}
alarm()
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{apropos}{Find Objects by (Partial) Name}{apropos}
\aliasA{find}{apropos}{find}
\keyword{data}{apropos}
\keyword{documentation}{apropos}
\keyword{environment}{apropos}
%
\begin{Description}\relax
\code{apropos()} returns a character vector giving the names of
all objects in the search list matching \code{what}.

\code{find()} is a different user interface to the same task.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
apropos(what, where = FALSE, ignore.case = TRUE, mode = "any")

find(what, mode = "any", numeric = FALSE, simple.words = TRUE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{what}] character string with name of an object, or more generally
a \LinkA{regular expression}{regular expression} to match against.
\item[\code{where, numeric}] a logical indicating whether positions in the
search list should also be returned
\item[\code{ignore.case}] logical indicating if the search should be
case-insensitive, \code{TRUE} by default.  Note that in \R{} versions
prior to 2.5.0, the default was implicitly \code{ignore.case = FALSE}.
\item[\code{mode}] character; if not \code{"any"}, only objects whose
\code{\LinkA{mode}{mode}} equals \code{mode} are searched.
\item[\code{simple.words}] logical; if \code{TRUE}, the \code{what} argument is
only searched as whole word.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If \code{mode != "any"} only those objects which are of mode \code{mode}
are considered.
If \code{where} is \code{TRUE}, the positions in the search list are
returned as the names attribute.

\code{find} is a different user interface for the same task as
\code{apropos}. However, by default (\code{simple.words == TRUE}),
only full words are searched with \code{\LinkA{grep}{grep}(fixed = TRUE)}.
\end{Details}
%
\begin{Value}
For \code{apropos} character vector, sorted by name, possibly with
names giving the (numerical) positions on the search path.

For \code{find}, either a character vector of environment names, or for
\code{numeric = TRUE}, a numerical vector of positions on the search path,
with names giving the names of the corresponding environments.
\end{Value}
%
\begin{Author}\relax
Kurt Hornik and Martin Maechler (May 1997).
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{glob2rx}{glob2rx}} to convert wildcard patterns to regular expressions.

\code{\LinkA{objects}{objects}} for listing objects from one place,
\code{\LinkA{help.search}{help.search}} for searching the help system,
\code{\LinkA{search}{search}} for the search path.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(stats)


## Not run: apropos("lm")
apropos("GLM")                      # more than a dozen
## that may include internal objects starting '.__C__' if
## methods is attached
apropos("GLM", ignore.case = FALSE) # not one
apropos("lq")

cor <- 1:pi
find("cor")        #> ".GlobalEnv"   "package:stats"
find("cor", numeric=TRUE) # numbers with these names
find("cor", numeric=TRUE, mode="function")# only the second one
rm(cor)

## Not run: apropos(".", mode="list") # a long list

# need a DOUBLE backslash '\\' (in case you don't see it anymore)
apropos("\\[")

## Not run: # everything 
length(apropos("."))

# those starting with 'pr'
apropos("^pr")

# the 1-letter things
apropos("^.$")
# the 1-2-letter things
apropos("^..?$")
# the 2-to-4 letter things
apropos("^.{2,4}$")

# the 8-and-more letter things
apropos("^.{8,}$")
table(nchar(apropos("^.{8,}$")))

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{aspell}{Aspell Interface}{aspell}
\keyword{utilities}{aspell}
%
\begin{Description}\relax
Spell check given files via Aspell.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
aspell(files, filter, control = list(), encoding = "unknown")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{files}] a character vector with the names of files to be checked.
\item[\code{filter}] an optional filter for processing the files before spell
checking, given as either a function (with formals \code{ifile} and
\code{encoding}), or a character string specifying a built-in
filter, or a list with the name of a built-in filter and additional
arguments to be passed to it.  See \bold{Details} for available
filters.  If missing or \code{NULL}, no filtering is performed.
\item[\code{control}] a list or character vector of control options for
Aspell.
\item[\code{encoding}] the encoding of the files.  Recycled as needed.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
It is assumed that the Aspell executable \code{aspell} is available in
the system search path.  See \url{http://aspell.net} for information
on obtaining Aspell, and available dictionaries.

Currently the only available built-in filters are \code{"Rd"},
corresponding to \code{\LinkA{RdTextFilter}{RdTextFilter}}, and \code{"Sweave"},
corresponding to \code{\LinkA{SweaveTeXFilter}{SweaveTeXFilter}}.

The print method has for the objects returned by \code{aspell} has an
\code{indent} argument controlling the indentation of the positions of
possibly mis-spelled words.  The default is 2; Emacs users may find it
useful to use an indentation of 0 and visit output in grep-mode.
\end{Details}
%
\begin{Value}
A data frame inheriting from \code{aspell} (which has a useful print
method) with the information about possibly mis-spelled words.
\end{Value}
%
\begin{SeeAlso}\relax
Package \pkg{Aspell} on Omegahat
(\url{http://www.omegahat.org/Aspell}) for a fine-grained R interface
to the Aspell library.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
# To check all Rd files in a directory, skipping the \references sections
files <- Sys.glob("*.Rd")
aspell(files, filter=list("Rd", drop="\references"))

# To check all Sweave files 
files <- Sys.glob(c("*.Rnw", "*.Snw", "*.rnw", "*.snw"))
aspell(files, filter="Sweave", control="--mode=tex")

# To check all Texinfo files
files <- Sys.glob("*.texi")
aspell(files, control="--mode=texinfo")


## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{BATCH}{Batch Execution of R}{BATCH}
\keyword{utilities}{BATCH}
%
\begin{Description}\relax
Run \R{} non-interactively with input from \code{infile} and
send output (stdout/stderr) to another file.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
R CMD BATCH [options] infile [outfile]
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{infile}] the name of a file with \R{} code to be executed.
\item[\code{options}] a list of \R{} command line options, e.g., for setting the
amount of memory available and controlling the load/save process.
If \code{infile} starts with a \samp{-}, use \option{--} as the final
option.
The default options are \option{--restore --save --no-readline}.
\item[\code{outfile}] the name of a file to which to write output.  If not
given, the name used is that of \code{infile}, with a possible
\file{.R} extension stripped, and \file{.Rout} appended.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Use \command{R CMD BATCH --help} to be reminded of the usage.

By default, the input commands are printed along with the output.  To
suppress this behavior, add \code{options(echo = FALSE)} at the
beginning of \code{infile}, or use option \option{--slave}.

The \code{infile} can have end of line marked by LF or CRLF (but not
just CR), and files with an incomplete last line (missing end of line
(EOL) mark) are processed correctly.

A final expression \samp{proc.time()} will be executed after the input
script unless the latter calls \code{\LinkA{q}{q}(runLast=FALSE)} or is aborted.
This can be suppressed by the option \option{--no-timing}.

Additional options can be set by the environment variable
\env{R\_BATCH\_OPTIONS}: these come after
\option{--restore --save --no-readline} and before any options given
on the command line.
\end{Details}
%
\begin{Note}\relax
Unlike \command{Splus BATCH}, this does not run the \R{} process in the
background.  In most shells,
\command{R CMD BATCH [options] infile [outfile] \&}
will do so.

Report bugs to \email{r-bugs@r-project.org}.
\end{Note}
\inputencoding{latin1}
\HeaderA{browseEnv}{Browse Objects in Environment}{browseEnv}
\aliasA{wsbrowser}{browseEnv}{wsbrowser}
\keyword{interface}{browseEnv}
%
\begin{Description}\relax
The \code{browseEnv} function opens a browser with list of objects
currently in \code{sys.frame()} environment.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
browseEnv(envir = .GlobalEnv, pattern,
          excludepatt = "^last\\.warning",
          html = .Platform$OS.type != "mac",
          expanded = TRUE, properties = NULL,
          main = NULL, debugMe = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{envir}] an \code{\LinkA{environment}{environment}} the objects of which are to
be browsed.
\item[\code{pattern}] a \LinkA{regular expression}{regular expression} for object subselection
is passed to the internal \code{\LinkA{ls}{ls}()} call.
\item[\code{excludepatt}] a regular expression for \emph{dropping} objects
with matching names.
\item[\code{html}] is used on non Macintosh machines to display the workspace
on a HTML page in your favorite browser.
\item[\code{expanded}] whether to show one level of recursion.  It can be useful
to switch it to \code{FALSE} if your workspace is large.  This
option is ignored if \code{html} is set to \code{FALSE}.
\item[\code{properties}] a named list of global properties (of the objects chosen)
to be showed in the browser;  when \code{NULL} (as per default),
user, date, and machine information is used.
\item[\code{main}] a title string to be used in the browser; when \code{NULL}
(as per default) a title is constructed.
\item[\code{debugMe}] logical switch; if true, some diagnostic output is produced.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Very experimental code.  Only allows one level of recursion into
object structures. The HTML version is not dynamic.

It can be generalized.  See sources
(\file{..../library/base/R/databrowser.R}) for details.

\code{wsbrowser()} is currently just an internally used function;
its argument list will certainly change.

Most probably, this should rather work through using the \file{tkWidget}
package (from \url{www.Bioconductor.org}).
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{str}{str}}, \code{\LinkA{ls}{ls}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
if(interactive()) {
   ## create some interesting objects :
   ofa <- ordered(4:1)
   ex1 <- expression(1+ 0:9)
   ex3 <- expression(u,v, 1+ 0:9)
   example(factor, echo = FALSE)
   example(table, echo = FALSE)
   example(ftable, echo = FALSE)
   example(lm, echo = FALSE, ask = FALSE)
   example(str, echo = FALSE)

   ## and browse them:
   browseEnv()

   ## a (simple) function's environment:
   af12 <- approxfun(1:2, 1:2, method = "const")
   browseEnv(envir = environment(af12))
 }
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{browseURL}{Load URL into a WWW Browser}{browseURL}
\keyword{file}{browseURL}
%
\begin{Description}\relax
Load a given URL into a WWW browser.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
browseURL(url, browser = getOption("browser"), encodeIfNeeded = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{url}] a non-empty character string giving the URL to be loaded.
\item[\code{browser}] a non-empty character string giving the name of the
program to be used as hypertext browser.  It should be in the PATH,
or a full path specified.  Alternatively, an \R{} function to be
called to invoke the browser.

\item[\code{encodeIfNeeded}] Should the URL be encoded by
\code{\LinkA{URLencode}{URLencode}} before passing to the browser?  This is not
needed (and might be harmful) if the \code{browser} program/function
itself does encoding, and can be harmful for \samp{file://} URLs on some
systems and for \samp{http://} URLs passed to some CGI applications.
Fortunately, most URLs do not need encoding.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The default browser is set by option \code{"browser"}, in turn set by
the environment variable \env{R\_BROWSER} which is by default set in
file \file{\var{\LinkA{R\_HOME}{R.Rul.HOME}}/etc/Renviron} to a choice
made manually or automatically when \R{} was configured.  (See
\code{\LinkA{Startup}{Startup}} for where to override that default value.)

If \code{browser} supports remote control and \R{} knows how to perform
it, the URL is opened in any already running browser or a new one if
necessary.  This mechanism currently is available for browsers which
support the \code{"-remote openURL(...)"} interface (which includes
Mozilla >= 0.9.5 and Mozilla Firefox), Galeon, KDE konqueror
(\emph{via} kfmclient) and the GNOME interface to Mozilla.  Note that
the type of browser is determined from its name, so this mechanism
will only be used if the browser is installed under its canonical
name.

Because \code{"-remote"} will use any browser displaying on the X
server (whatever machine it is running on), the remote control
mechanism is only used if \code{DISPLAY} points to the local host.
This may not allow displaying more than one URL at a time from a
remote host.

It is the caller's responsibility to encode \code{url} if necessary
(see \code{\LinkA{URLencode}{URLencode}}).  This can be tricky for file URLs,
where the format accepted can depend on the browser and OS.
\end{Details}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: ## for KDE users who want to open files in a new tab
options(browser="kfmclient newTab") 
browseURL("http://www.r-project.org")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{browseVignettes}{List Vignettes in an HTML Browser}{browseVignettes}
\aliasA{print.browseVignettes}{browseVignettes}{print.browseVignettes}
\keyword{documentation}{browseVignettes}
%
\begin{Description}\relax
List available vignettes in an HTML browser with links to PDF,
LaTeX/noweb source, and (tangled) R code (if available).
\end{Description}
%
\begin{Usage}
\begin{verbatim}
browseVignettes(package = NULL, lib.loc = NULL, all = TRUE)

## S3 method for class 'browseVignettes':
print(x, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{package}] a character vector with the names of packages to
search through, or \code{NULL} in which "all" packages (as defined
by argument \code{all}) are searched.
\item[\code{lib.loc}] a character vector of directory names of \R{} libraries,
or \code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.
\item[\code{all}] logical; if \code{TRUE} search
all available packages in the library trees specified by \code{lib.loc}, 
and if \code{FALSE}, search only attached packages.
\item[\code{x}] Object of class \code{browseVignettes}.
\item[\code{...}] Further arguments, ignored by the \code{print} method. 
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Function \code{browseVignettes} returns an object of the same class;
the print method displays it as an HTML page in a browser (using
\code{\LinkA{browseURL}{browseURL}}).
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{browseURL}{browseURL}}, \code{\LinkA{vignette}{vignette}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
## List vignettes from all *attached* packages
browseVignettes(all = FALSE)

## List vignettes from a specific package
browseVignettes("grid")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{bug.report}{Send a Bug Report}{bug.report}
\keyword{utilities}{bug.report}
\keyword{error}{bug.report}
%
\begin{Description}\relax
Invokes an editor to write a bug report and optionally mail it to the
automated r-bugs repository at \email{r-bugs@r-project.org}.  Some standard
information on the current version and configuration of \R{} are
included automatically.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
bug.report(subject = "",
           ccaddress = Sys.getenv("USER"),
           method = getOption("mailer"),
           address = "r-bugs@r-project.org",
           file = "R.bug.report")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{subject}] Subject of the email. Please do not use single quotes
(\samp{'}) in the subject! File separate bug reports for multiple bugs
\item[\code{ccaddress}] Optional email address for copies (default is current
user).  Use \code{ccaddress = FALSE} for no copies.
\item[\code{method}] Submission method, one of \code{"mailx"},
\code{"gnudoit"}, \code{"none"}, or \code{"ess"}.
\item[\code{address}] Recipient's email address.
\item[\code{file}] File to use for setting up the email (or storing it when
method is \code{"none"} or sending mail fails).
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Currently direct submission of bug reports works only on Unix systems.
If the submission method is \code{"mailx"}, then the default editor is
used to write the bug report. Which editor is used can be controlled
using \code{\LinkA{options}{options}}, type \code{getOption("editor")} to see what
editor is currently defined. Please use the help pages of the
respective editor for details of usage. After saving the bug report
(in the temporary file opened) and exiting the editor
the report is mailed using a Unix command line mail utility such as
\code{mailx}.  A copy of the mail is sent to the current user.

If method is \code{"gnudoit"}, then an emacs mail buffer is opened
and used for sending the email.

If method is \code{"none"} or \code{NULL} (and in every case on
Windows systems), then only an editor is opened to help writing the
bug report.  The report can then be copied to your favorite email
program and be sent to the r-bugs list.

If method is \code{"ess"} the body of the mail is simply sent to
stdout.
\end{Details}
%
\begin{Value}
Nothing useful.
\end{Value}
%
\begin{Section}{When is there a bug?}
If \R{} executes an illegal instruction, or dies with an operating
system error message that indicates a problem in the program (as
opposed to something like ``disk full''), then it is certainly a
bug.

Taking forever to complete a command can be a bug, but you must make
certain that it was really \R{}'s fault.  Some commands simply take a
long time.  If the input was such that you KNOW it should have been
processed quickly, report a bug.  If you don't know whether the
command should take a long time, find out by looking in the manual or
by asking for assistance.

If a command you are familiar with causes an \R{} error message in a
case where its usual definition ought to be reasonable, it is probably
a bug.  If a command does the wrong thing, that is a bug.  But be sure
you know for certain what it ought to have done.  If you aren't
familiar with the command, or don't know for certain how the command
is supposed to work, then it might actually be working right.  Rather
than jumping to conclusions, show the problem to someone who knows for
certain.

Finally, a command's intended definition may not be best for
statistical analysis.  This is a very important sort of problem, but
it is also a matter of judgement.  Also, it is easy to come to such a
conclusion out of ignorance of some of the existing features.  It is
probably best not to complain about such a problem until you have
checked the documentation in the usual ways, feel confident that you
understand it, and know for certain that what you want is not
available. The mailing list \code{r-devel@r-project.org} is a better
place for discussions of this sort than the bug list.

If you are not sure what the command is supposed to do
after a careful reading of the manual this indicates a bug in the
manual.  The manual's job is to make everything clear.  It is just as
important to report documentation bugs as program bugs.

If the online argument list of a function disagrees with the manual,
one of them must be wrong, so report the bug.
\end{Section}
%
\begin{Section}{How to report a bug}
When you decide that there is a bug, it is important to report it and
to report it in a way which is useful.  What is most useful is an
exact description of what commands you type, from when you start \R{}
until the problem happens.  Always include the version of \R{}, machine,
and operating system that you are using; type \kbd{version} in \R{} to
print this.  To help us keep track of which bugs have been fixed and
which are still open please send a separate report for each bug.

The most important principle in reporting a bug is to report FACTS,
not hypotheses or categorizations.  It is always easier to report the
facts, but people seem to prefer to strain to posit explanations and
report them instead.  If the explanations are based on guesses about
how \R{} is implemented, they will be useless; we will have to try to
figure out what the facts must have been to lead to such
speculations.  Sometimes this is impossible.  But in any case, it is
unnecessary work for us.

For example, suppose that on a data set which you know to be quite
large the command \code{data.frame(x, y, z, monday, tuesday)} never
returns.  Do not report that \code{data.frame()} fails for large data
sets.  Perhaps it fails when a variable name is a day of the week.  If
this is so then when we got your report we would try out the
\code{data.frame()} command on a large data set, probably with no day
of the week variable name, and not see any problem. There is no way in
the world that we could guess that we should try a day of the week
variable name.

Or perhaps the command fails because the last command you used was a
\code{[} method that had a bug causing \R{}'s internal data structures
to be corrupted and making the \code{data.frame()} command fail from
then on.  This is why we need to know what other commands you have
typed (or read from your startup file).

It is very useful to try and find simple examples that produce
apparently the same bug, and somewhat useful to find simple examples
that might be expected to produce the bug but actually do not.  If you
want to debug the problem and find exactly what caused it, that is
wonderful.  You should still report the facts as well as any
explanations or solutions.

Invoking \R{} with the \option{--vanilla} option may help in isolating a
bug.  This ensures that the site profile and saved data files are not
read.

A bug report can be generated using the
\code{bug.report()} function.  This automatically includes the version
information and sends the bug to the correct address.  Alternatively
the bug report can be emailed to \email{r-bugs@r-project.org} or
submitted to the Web page at \url{http://bugs.r-project.org}.

Bug reports on \strong{contributed packages} should be sent to the
package maintainer rather than to r-bugs.
\end{Section}
%
\begin{Author}\relax
This help page is adapted from the Emacs manual and the R FAQ
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{help.request}{help.request}} which you possibly should try
\emph{before} \code{bug.report}.
The R FAQ, also \code{\LinkA{sessionInfo}{sessionInfo}()} from which you may add
to the bug report.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{capture.output}{Send output to a character string or file}{capture.output}
\keyword{utilities}{capture.output}
%
\begin{Description}\relax
Evaluates its arguments with the output being returned as a character
string or sent to a file.  Related to \code{\LinkA{sink}{sink}} in the same
way that \code{\LinkA{with}{with}} is related to \code{\LinkA{attach}{attach}}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
capture.output(..., file = NULL, append = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{...}] Expressions to be evaluated.
\item[\code{file}] A file name or a connection, or \code{NULL} to return
the output as a character vector.  If the connection is not open,
it will be opened initially and closed on exit.
\item[\code{append}] logical.  If \code{file} a file name or unopened
connection, append or overwrite?
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
An attempt is made to write output as far as possible to \code{file}
if there is an error in evaluating the expressions, but for
\code{file = NULL} all output will be lost.
\end{Details}
%
\begin{Value}
A character string (if \code{file=NULL}), or invisible \code{NULL}.
\end{Value}
%
\begin{SeeAlso}\relax
 \code{\LinkA{sink}{sink}}, \code{\LinkA{textConnection}{textConnection}} 
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(stats)
glmout <- capture.output(example(glm))
glmout[1:5]
capture.output(1+1, 2+2)
capture.output({1+1; 2+2})
## Not run: 
## on Unix with enscript available
ps <- pipe("enscript -o tempout.ps","w")
capture.output(example(glm), file=ps)
close(ps)

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{chooseCRANmirror}{Select a CRAN Mirror}{chooseCRANmirror}
\aliasA{getCRANmirrors}{chooseCRANmirror}{getCRANmirrors}
\keyword{utilities}{chooseCRANmirror}
%
\begin{Description}\relax
Interact with the user to choose a CRAN mirror.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
chooseCRANmirror(graphics = getOption("menu.graphics"))

getCRANmirrors(all = FALSE, local.only = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{graphics}] Logical.
If true and \pkg{tcltk} and an X server are available, use a Tk
widget, or if under the AQUA interface use a Mac OS X widget,
otherwise use \code{\LinkA{menu}{menu}}.
\item[\code{all}] Logical, get all known mirrors or only the ones flagged as
OK.
\item[\code{local.only}] Logical, try to get most recent list from CRAN or
use file on local disk only.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
A list of mirrors is stored in file
\file{\var{\LinkA{R\_HOME}{R.Rul.HOME}}/doc/CRAN\_mirrors.csv}, but first an on-line list of
current mirrors is consulted, and the file copy used only if the
on-line list is inaccessible.

This function was originally written to support a Windows GUI menu
item, but is also called by \code{\LinkA{contrib.url}{contrib.url}} if it finds the
initial dummy value of \code{\LinkA{options}{options}("repos")}.
\end{Details}
%
\begin{Value}
None for \code{chooseCRANmirror()}, this function is invoked for its
side effect of updating \code{options("repos")}.

\code{getCRANmirrors()} returns a data frame with mirror information.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{setRepositories}{setRepositories}}, \code{\LinkA{contrib.url}{contrib.url}}.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{citation}{Citing R and R Packages in Publications}{citation}
\aliasA{toBibtex.citation}{citation}{toBibtex.citation}
\aliasA{toBibtex.citationList}{citation}{toBibtex.citationList}
\keyword{misc}{citation}
%
\begin{Description}\relax
How to cite R and R packages in publications.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
citation(package = "base", lib.loc = NULL)
## S3 method for class 'citation':
toBibtex(object, ...)
## S3 method for class 'citationList':
toBibtex(object, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{package}] a character string with the name of a single package.
An error occurs if more than one package name is given.
\item[\code{lib.loc}] a character vector with path names of \R{} libraries, or
\code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.  If the default is used, the loaded
packages are searched before the libraries.
\item[\code{object}] return object of \code{citation}.
\item[\code{...}] currently not used.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The R core development team and the very active community of package
authors have invested a lot of time and effort in creating R as it is
today. Please give credit where credit is due and cite R and R
packages when you use them for data analysis.

Execute function \code{citation()} for information on how to cite the
base R system in publications. If the name of a non-base package is
given, the function
either returns the information contained in the \code{\LinkA{CITATION}{CITATION}}
file of the package or auto-generates citation information. In the
latter case the package \file{DESCRIPTION} file is parsed, the
resulting citation object may be arbitrarily bad, but is quite useful
(at least as a starting point) in most cases.

If only one reference is given, the print method shows both a text
version and a BibTeX entry for it, if a package has more than one
reference then only the text versions are shown. The BibTeX versions
can be obtained using function \code{toBibtex} (see the examples below).
\end{Details}
%
\begin{Value}
An object of class \code{"citationList"}.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{citEntry}{citEntry}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## the basic R reference
citation()

## references for a package -- might not have these installed
if(nchar(system.file(package="lattice"))) citation("lattice")
if(nchar(system.file(package="foreign"))) citation("foreign")

## extract the bibtex entry from the return value
x <- citation()
toBibtex(x)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{citEntry}{Writing Package CITATION Files}{citEntry}
\aliasA{CITATION}{citEntry}{CITATION}
\aliasA{citFooter}{citEntry}{citFooter}
\aliasA{citHeader}{citEntry}{citHeader}
\aliasA{readCitationFile}{citEntry}{readCitationFile}
\keyword{misc}{citEntry}
%
\begin{Description}\relax
The \file{CITATION} file of R packages contains an annotated list of
references that should be used for citing the packages.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
citEntry(entry, textVersion, header = NULL, footer = NULL, ...)
citHeader(...)
citFooter(...)
readCitationFile(file, meta = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{entry}] a character string with a BibTeX entry type
\item[\code{textVersion}] a character string with a text representation of
the reference
\item[\code{header}] a character string with optional header text
\item[\code{footer}] a character string with optional footer text
\item[\code{file}] a file name
\item[\code{...}] see details below
\item[\code{meta}] a list of package metadata as obtained by
\code{\LinkA{packageDescription}{packageDescription}}, or \code{NULL} (default).
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The \file{CITATION} file of an R package should be placed in the
\file{inst} subdirectory of the package source. The file is an R
source file and may contain arbitrary R commands including
conditionals and computations. The file is \code{source()}ed by the R
parser in a temporary environment and all resulting objects of class
\code{"citation"} (the return value of \code{citEntry}) are collected.

Typically the file will contain zero or more calls to \code{citHeader},
then one or more calls to \code{citEntry}, and  finally zero or more
calls to  \code{citFooter}. \code{citHeader} and \code{citFooter} are
simply wrappers to \code{\LinkA{paste}{paste}}, and their \code{...} argument
is passed on to \code{\LinkA{paste}{paste}} as is.
\end{Details}
%
\begin{Value}
\code{citEntry} returns an object of class \code{"citation"},
\code{readCitationFile} returns an object of class \code{"citationList"}.
\end{Value}
%
\begin{Section}{Entry Types}
\code{citEntry} creates \code{"citation"} objects, which are modeled
after BibTeX entries. The entry should be a valid BibTeX entry type,
e.g.,
\begin{description}

\item[article:] An article from a journal or magazine.
\item[book:] A book with an explicit publisher.
\item[inbook:] A part of a book, which may be a chapter (or section
or whatever) and/or a range of pages. 
\item[incollection:] A part of a book having its own title.
\item[inproceedings:] An article in a conference proceedings.
\item[manual:] Technical documentation like a software manual.
\item[mastersthesis:] A Master's thesis.
\item[misc:] Use this type when nothing else fits.
\item[phdthesis:] A PhD thesis.
\item[proceedings:] The proceedings of a conference.
\item[techreport:] A report published by a school or other
institution, usually numbered within a series.
\item[unpublished:] A document having an author and title, but not
formally published.

\end{description}

\end{Section}
%
\begin{Section}{Entry Fields}
The \code{...} argument of \code{citEntry} can be any number of
BibTeX fields, including
\begin{description}

\item[address:] The address of the publisher or other type of
institution.

\item[author:] The name(s) of the author(s), either 
as a character string in the format described in the LaTeX book,
or a \code{\LinkA{personList}{personList}} object.

\item[booktitle:] Title of a book, part of which is being cited.
\item[chapter:] A chapter (or section or whatever) number.

\item[editor:] Name(s) of editor(s), same format as \code{author}.

\item[institution:] The publishing institution of a technical report.

\item[journal:] A journal name.

\item[note:] Any additional information that can help the reader.
The first word should be capitalized.

\item[number:] The number of a journal, magazine, technical report,
or of a work in a series.

\item[pages:] One or more page numbers or range of numbers.

\item[publisher:] The publisher's name.

\item[school:] The name of the school where a thesis was written.

\item[series:] The name of a series or set of books.

\item[title:] The work's title.

\item[volume:] The volume of a journal or multi-volume book.

\item[year:] The year of publication.

\end{description}

\end{Section}
%
\begin{Examples}
\begin{ExampleCode}
basecit <- system.file("CITATION", package="base")
source(basecit, echo=TRUE)
readCitationFile(basecit)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{close.socket}{Close a Socket}{close.socket}
\keyword{misc}{close.socket}
%
\begin{Description}\relax
Closes the socket and frees the space in the file descriptor table.  The
port may not be freed immediately.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
close.socket(socket, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{socket}] A \code{socket} object
\item[\code{...}] further arguments passed to or from other methods.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
logical indicating success or failure
\end{Value}
%
\begin{Author}\relax
Thomas Lumley
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{make.socket}{make.socket}}, \code{\LinkA{read.socket}{read.socket}}
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{combn}{Generate All Combinations of n Elements, Taken m at a Time}{combn}
\keyword{utilities}{combn}
\keyword{iteration}{combn}
%
\begin{Description}\relax
Generate all combinations of the elements of \code{x} taken \code{m}
at a time.  If \code{x} is a positive integer, returns all
combinations of the elements of \code{seq(x)} taken \code{m} at a
time.  If argument \code{FUN} is not \code{NULL}, applies a function given
by the argument to each point.  If simplify is FALSE,  returns
a list; otherwise returns an \code{\LinkA{array}{array}}, typically a
\code{\LinkA{matrix}{matrix}}.  \code{...} are passed unchanged to the
\code{FUN} function, if specified.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
combn(x, m, FUN = NULL, simplify = TRUE, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] vector source for combinations, or integer \code{n} for
\code{x <- \LinkA{seq}{seq}(n)}.
\item[\code{m}] number of elements to choose.
\item[\code{FUN}] function to be applied to each combination; default
\code{NULL} means the identity, i.e., to return the combination
(vector of length \code{m}).
\item[\code{simplify}] logical indicating if the result should be simplified
to an \code{\LinkA{array}{array}} (typically a \code{\LinkA{matrix}{matrix}}); if
FALSE, the function returns a \code{\LinkA{list}{list}}.  Note that when
\code{simplify = TRUE} as by default, the dimension of the result is
simply determined from \code{FUN(\var{1st combination})} (for
efficiency reasons).  This will badly fail if \code{FUN(u)} is not of
constant length.
\item[\code{...}] optionally, further arguments to \code{FUN}.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
a \code{\LinkA{list}{list}} or \code{\LinkA{array}{array}} (in nondegenerate cases),
see the \code{simplify} argument above.
\end{Value}
%
\begin{Author}\relax
Scott Chasalow wrote the original in 1994 for S;
R package \pkg{combinat} and documentation by Vince Carey
\email{stvjc@channing.harvard.edu};
small changes by the R core team, notably to return an array in all
cases of \code{simplify = TRUE}, e.g., for \code{combn(5,5)}.
\end{Author}
%
\begin{References}\relax
Nijenhuis, A. and Wilf, H.S. (1978)
\emph{Combinatorial Algorithms for Computers and Calculators};
Academic Press, NY.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{choose}{choose}} for fast computation of the \emph{number} of
combinations. \code{\LinkA{expand.grid}{expand.grid}} for creating a data frame from
all combinations of factors or vectors.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
combn(letters[1:4], 2)
(m <- combn(10, 5, min))   # minimum value in each combination
mm <- combn(15, 6, function(x) matrix(x, 2,3))
stopifnot(round(choose(10,5)) == length(m),
          c(2,3, round(choose(15,6))) == dim(mm))

## Different way of encoding points:
combn(c(1,1,1,1,2,2,2,3,3,4), 3, tabulate, nbins = 4)

## Compute support points and (scaled) probabilities for a
## Multivariate-Hypergeometric(n = 3, N = c(4,3,2,1)) p.f.:
# table.mat(t(combn(c(1,1,1,1,2,2,2,3,3,4), 3, tabulate,nbins=4)))
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{compareVersion}{Compare Two Package Version Numbers}{compareVersion}
\keyword{utilities}{compareVersion}
%
\begin{Description}\relax
Compare two package version numbers to see which is later.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
compareVersion(a, b)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{a, b}] Character strings representing package version numbers.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\R{} package version numbers are of the form \code{x.y-z} for integers
\code{x}, \code{y} and \code{z}, with components after \code{x}
optionally missing (in which case the version number is older than
those with the components present). 
\end{Details}
%
\begin{Value}
\code{0} if the numbers are equal, \code{-1} if \code{b} is later
and \code{1} if \code{a} is later (analogous to the C function
\code{strcmp}).
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{package\_version}{package.Rul.version}},
\code{\LinkA{library}{library}}, \code{\LinkA{packageStatus}{packageStatus}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
compareVersion("1.0", "1.0-1")
compareVersion("7.2-0","7.1-12")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{COMPILE}{Compile Files for Use with R}{COMPILE}
\keyword{utilities}{COMPILE}
%
\begin{Description}\relax
Compile given source files so that they can subsequently be collected
into a shared library using \command{R CMD SHLIB} and be loaded into R
using \code{dyn.load()}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
R CMD COMPILE [options] srcfiles
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{srcfiles}] A list of the names of source files to be compiled.
Currently, C, C++, Objective C, Objective C++ and FORTRAN are
supported; the corresponding files should have the extensions
\file{.c}, \file{.cc} (or \file{.cpp} or \file{.C}), \file{.m},
\file{.mm} (or \file{.M}) and \file{.f}, respectively.
\item[\code{options}] A list of compile-relevant settings, such as special
values for \code{CFLAGS} or \code{FFLAGS}, or for obtaining
information about usage and version of the utility.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Note that Ratfor is not supported.  If you have Ratfor source code,
you need to convert it to FORTRAN.  On many Solaris systems mixing
Ratfor and FORTRAN code will work.

Objective C and Objective C++ support is optional and will work only
if the corresponding compilers were available at R configure time.
\end{Details}
%
\begin{Note}\relax
Some binary distributions of \R{} have \code{COMPILE} in a separate
bundle, e.g. an \code{R-devel} RPM.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{SHLIB}{SHLIB}},
\code{\LinkA{dyn.load}{dyn.load}};
the section on ``Customizing compilation under Unix'' in
``R Administration and Installation''
(see the \file{doc/manual} subdirectory of the \R{} source tree).  
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{count.fields}{Count the Number of Fields per Line}{count.fields}
\keyword{file}{count.fields}
%
\begin{Description}\relax
\code{count.fields} counts the number of fields, as separated by
\code{sep}, in each of the lines of \code{file} read.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
count.fields(file, sep = "", quote = "\"'", skip = 0,
             blank.lines.skip = TRUE, comment.char = "#")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] a character string naming an ASCII data file, or a
\code{\LinkA{connection}{connection}}, which will be opened if necessary,
and if so closed at the end of the function call.

\item[\code{sep}] the field separator character.  Values on each line of the
file are separated by this character.  By default, arbitrary amounts
of whitespace can separate fields.

\item[\code{quote}] the set of quoting characters

\item[\code{skip}] the number of lines of the data file to skip before
beginning to read data.

\item[\code{blank.lines.skip}] logical: if \code{TRUE} blank lines in the
input are ignored.

\item[\code{comment.char}] character: a character vector of length one
containing a single character or an empty string.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This used to be used by \code{\LinkA{read.table}{read.table}} and can still be
useful in discovering problems in reading a file by that function.

For the handling of comments, see \code{\LinkA{scan}{scan}}.
\end{Details}
%
\begin{Value}
A vector with the numbers of fields found.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{read.table}{read.table}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
cat("NAME", "1:John", "2:Paul", file = "foo", sep = "\n")
count.fields("foo", sep = ":")
unlink("foo")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{data}{Data Sets}{data}
\aliasA{print.packageIQR}{data}{print.packageIQR}
\keyword{documentation}{data}
\keyword{datasets}{data}
%
\begin{Description}\relax
Loads specified data sets, or list the available data sets.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
data(..., list = character(0), package = NULL, lib.loc = NULL,
     verbose = getOption("verbose"), envir = .GlobalEnv)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{...}] a sequence of names or literal character strings.
\item[\code{list}] a character vector.
\item[\code{package}] 
a character vector giving the package(s) to look
in for data sets, or \code{NULL}.

By default, all packages in the search path are used, then
the \file{data} subdirectory (if present) of the current working
directory.

\item[\code{lib.loc}] a character vector of directory names of \R{} libraries,
or \code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.
\item[\code{verbose}] a logical.  If \code{TRUE}, additional diagnostics are
printed.
\item[\code{envir}] the \LinkA{environment}{environment} where the data should be loaded.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Currently, four formats of data files are supported:

\begin{enumerate}

\item files ending \file{.R} or \file{.r} are
\code{\LinkA{source}{source}()}d in, with the \R{} working directory changed
temporarily to the directory containing the respective file.
(\code{data} ensures that the \pkg{utils} package is attached, in
case it had been run \emph{via} \code{utils::data}.)

\item files ending \file{.RData} or \file{.rda} are
\code{\LinkA{load}{load}()}ed.

\item files ending \file{.tab}, \file{.txt} or \file{.TXT} are read
using \code{\LinkA{read.table}{read.table}(..., header = TRUE)}, and hence
result in a data frame.

\item files ending \file{.csv} or \file{.CSV} are read using
\code{\LinkA{read.table}{read.table}(..., header = TRUE, sep = ";")},
and also result in a data frame.

\end{enumerate}

If more than one matching file name is found, the first on this list
is used.  (Files with extensions \file{.txt}, \file{.tab} or
\file{.csv} can be compressed, with or without further extension
\file{.gz}, \file{.bz2} or \file{.xz}.)

The data sets to be loaded can be specified as a sequence of names or
character strings, or as the character vector \code{list}, or as both.

For each given data set, the first two types (\file{.R} or \file{.r},
and \file{.RData} or \file{.rda} files) can create several variables
in the load environment, which might all be named differently from the
data set.  The third and fourth types will always result in the
creation of a single variable with the same name (without extension)
as the data set.

If no data sets are specified, \code{data} lists the available data
sets.  It looks for a new-style data index in the \file{Meta} or, if
this is not found, an old-style \file{00Index} file in the \file{data}
directory of each specified package, and uses these files to prepare a
listing.  If there is a \file{data} area but no index, available data
files for loading are computed and included in the listing, and a
warning is given: such packages are incomplete.  The information about
available data sets is returned in an object of class
\code{"packageIQR"}.  The structure of this class is experimental.
Where the datasets have a different name from the argument that should
be used to retrieve them the index will have an entry like
\code{beaver1 (beavers)} which tells us that dataset \code{beaver1}
can be retrieved by the call \code{data(beaver)}.

If \code{lib.loc} and \code{package} are both \code{NULL} (the
default), the data sets are searched for in all the currently loaded
packages then in the \file{data} directory (if any) of the current
working directory.

If \code{lib.loc = NULL} but \code{package} is specified as a
character vector, the specified package(s) are searched for first
amongst loaded packages and then in the default library/ies
(see \code{\LinkA{.libPaths}{.libPaths}}).

If \code{lib.loc} \emph{is} specified (and not \code{NULL}), packages
are searched for in the specified library/ies, even if they are
already loaded from another library.

To just look in the \file{data} directory of the current working
directory, set \code{package = character(0)} (and \code{lib.loc =
    NULL}, the default).
\end{Details}
%
\begin{Value}
A character vector of all data sets specified, or information about
all available data sets in an object of class \code{"packageIQR"} if
none were specified.
\end{Value}
%
\begin{Note}\relax
The data files can be many small files.  On some file systems it is
desirable to save space, and the files in the \file{data} directory of
an installed package can be zipped up as a zip archive
\file{Rdata.zip}.  You will need to provide a single-column file
\file{filelist} of file names in that directory.

One can take advantage of the search order and the fact that a
\file{.R} file will change directory.  If raw data are stored in
\file{mydata.txt} then one can set up \file{mydata.R} to read
\file{mydata.txt} and pre-process it, e.g., using \code{transform}.
For instance one can convert numeric vectors to factors with the
appropriate labels.  Thus, the \file{.R} file can effectively contain
a metadata specification for the plaintext formats.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{help}{help}} for obtaining documentation on data sets,
\code{\LinkA{save}{save}} for \emph{creating} the second (\file{.rda}) kind
of data, typically the most efficient one.

The `Writing R Extensions' for considerations in preparing the
\file{data} directory of a package.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(utils)
data()                       # list all available data sets
try(data(package = "rpart") )# list the data sets in the rpart package
data(USArrests, "VADeaths")  # load the data sets 'USArrests' and 'VADeaths'
help(USArrests)              # give information on data set 'USArrests'
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{dataentry}{Spreadsheet Interface for Entering Data}{dataentry}
\aliasA{data.entry}{dataentry}{data.entry}
\aliasA{de}{dataentry}{de}
\methaliasA{de.ncols}{dataentry}{de.ncols}
\methaliasA{de.restore}{dataentry}{de.restore}
\methaliasA{de.setup}{dataentry}{de.setup}
\keyword{utilities}{dataentry}
\keyword{file}{dataentry}
%
\begin{Description}\relax
A spreadsheet-like editor for entering or editing data.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
data.entry(..., Modes = NULL, Names = NULL)
dataentry(data, modes)
de(..., Modes = list(), Names = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{...}] A list of variables: currently these should be numeric or
character vectors or list containing such vectors.
\item[\code{Modes}] The modes to be used for the variables.
\item[\code{Names}] The names to be used for the variables.
\item[\code{data}] A list of numeric and/or character vectors.
\item[\code{modes}] A list of length up to that of \code{data} giving the
modes of (some of) the variables. \code{list()} is allowed.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The data entry editor is only available on some platforms and GUIs.
Where available it provides a means to visually edit a matrix or
a collection of variables (including a data frame) as described in the
Notes section.

\code{data.entry} has side effects, any changes made in the
spreadsheet are reflected in the variables.  The functions \code{de},
\code{de.ncols}, \code{de.setup} and \code{de.restore} are designed to
help achieve these side effects. If the user passes in a matrix,
\code{X} say, then the matrix is broken into columns before
\code{dataentry} is called. Then on return the columns are collected
and glued back together and the result assigned to the variable
\code{X}.  If you don't want this behaviour use dataentry directly.

The primitive function is \code{dataentry}. It takes a list of
vectors of possibly different lengths and modes (the second argument)
and opens a spreadsheet with these variables being the columns.
The columns of the dataentry window are returned as vectors in a
list when the spreadsheet is closed.

\code{de.ncols} counts the number of columns which are supplied as arguments
to \code{data.entry}. It attempts to count columns in lists, matrices
and vectors.  \code{de.setup} sets things up so that on return the
columns can be regrouped and reassigned to the correct name. This
is handled by \code{de.restore}.
\end{Details}
%
\begin{Value}
\code{de} and \code{dataentry} return the edited value of their
arguments. \code{data.entry} invisibly returns a vector of variable
names but its main value is its side effect of assigning new version
of those variables in the user's workspace.
\end{Value}
%
\begin{Section}{Resources}
The data entry window responds to X resources of class
\code{R\_dataentry}.  Resources \code{foreground}, \code{background} and
\code{geometry} are utilized.
\end{Section}
%
\begin{Note}\relax
The details of interface to the data grid may differ by platform and
GUI.  The following description applies to
the X11-based implementation under Unix.

You can navigate around the grid using the cursor keys or by clicking
with the (left) mouse button on any cell.  The active cell is
highlighted by thickening the surrounding rectangle.  Moving to the
right or down will scroll the grid as needed: there is no constraint
to the rows or columns currently in use.

There are alternative ways to navigate using the keys.  Return and
(keypad) Enter and LineFeed all move down. Tab moves right and
Shift-Tab move left.  Home moves to the top left.

PageDown or Control-F moves down a page, and PageUp or
Control-B up by a page.  End will show the last used column and the
last few rows used (in any column).

Using any other key starts an editing process on the currently
selected cell: moving away from that cell enters the edited value
whereas Esc cancels the edit and restores the previous value.  When
the editing process starts the cell is cleared.
In numerical columns
(the default) only letters making up a valid number (including
\code{-.eE}) are accepted, and entering an invalid edited value (such
as blank) enters \code{NA} in that cell.  The last entered value can
be deleted using the  BackSpace or Del(ete) key.  Only a limited
number of characters (currently 29) can be entered in a cell, and if
necessary only the start or end of the string will be displayed, with the
omissions indicated by \code{>} or \code{<}.  (The start is shown
except when editing.)


Entering a value in a cell further down a column than the last used
cell extends the variable and fills the gap (if any) by \code{NA}s (not
shown on screen).

The column names can only be selected by clicking in them.  This gives
a popup menu to select the column type (currently Real (numeric) or
Character) or to change the name.  Changing the type converts the
current contents of the column (and converting from Character to Real
may generate \code{NA}s.)
If changing the name is selected the
header cell becomes editable (and is cleared).  As with all cells, the
value is entered by moving away from the cell by clicking elsewhere or
by any of the keys for moving down (only).

New columns are created by entering values in them (and not by just
assigning a new name).  The mode of the column is auto-detected from
the first value entered: if this is a valid number it gives a numeric
column.  Unused columns are ignored, so
adding data in \code{var5} to a three-column grid adds one extra
variable, not two.

The \code{Copy} button copies the currently selected cell:
\code{paste} copies the last copied value to the current cell, and
right-clicking selects a cell \emph{and} copies in the value.
Initially the value is blank, and attempts to paste a blank value will
have no effect.

Control-L will refresh the display, recalculating field widths to fit
the current entries.

In the default mode the column widths are chosen to fit the contents
of each column, with a default of 10 characters for empty columns.
you can specify fixed column widths by setting option
\code{de.cellwidth} to the required fixed width (in characters).
(set it to zero to return to variable widths).  The displayed
width of any field is limited to
600 pixels (and by the window width).
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{vi}{vi}}, \code{\LinkA{edit}{edit}}: \code{edit} uses
\code{dataentry} to edit data frames.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
# call data entry with variables x and y
## Not run: data.entry(x,y)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{debugger}{Post-Mortem Debugging}{debugger}
\aliasA{dump.frames}{debugger}{dump.frames}
\keyword{utilities}{debugger}
\keyword{error}{debugger}
%
\begin{Description}\relax
Functions to dump the evaluation environments (frames) and to examine
dumped frames.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
dump.frames(dumpto = "last.dump", to.file = FALSE)
debugger(dump = last.dump)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{dumpto}] a character string. The name of the object or file to
dump to.
\item[\code{to.file}] logical. Should the dump be to an \R{} object or to a
file?
\item[\code{dump}] An \R{} dump object created by \code{dump.frames}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
To use post-mortem debugging, set the option \code{error} to be a call
to \code{dump.frames}.  By default this dumps to an \R{} object
\code{"last.dump"} in the workspace, but it can be set to dump to a
file (a dump of the object produced by a call to \code{\LinkA{save}{save}}).
The dumped object contain the call stack, the active environments and
the last error message as returned by \code{\LinkA{geterrmessage}{geterrmessage}}.

When dumping to file, \code{dumpto} gives the name of the dumped
object and the file name has \file{.rda} appended.

A dump object of class \code{"dump.frames"} can be examined
by calling \code{debugger}. This will give the error message and a
list of environments from which to select repeatedly. When an
environment is selected, it is copied and the \code{browser} called
from within the copy.

If \code{dump.frames} is installed as the error handler, execution
will continue even in non-interactive sessions. See the examples for
how to dump and then quit.
\end{Details}
%
\begin{Value}
Invisible \code{NULL}.
\end{Value}
%
\begin{Note}\relax
Functions such as \code{\LinkA{sys.parent}{sys.parent}} and
\code{\LinkA{environment}{environment}} applied to closures will not work correctly
inside \code{debugger}.

If the error occurred when computing the default value of a formal argument
the debugger will report "recursive default argument reference" when
trying to examine that environment. 

Of course post-mortem debugging will not work if \R{} is too damaged to
produce and save the dump, for example if it has run out of workspace.
\end{Note}
%
\begin{References}\relax
Becker, R. A., Chambers, J. M. and Wilks, A. R. (1988)
\emph{The New S Language}.
Wadsworth \& Brooks/Cole.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{options}{options}} for setting \code{error} options;
\code{\LinkA{recover}{recover}} is an interactive debugger working similarly to
\code{debugger} but directly after the error occurs.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
options(error=quote(dump.frames("testdump", TRUE)))

f <- function() {
    g <- function() stop("test dump.frames")
    g()
}
f()   # will generate a dump on file "testdump.rda"
options(error=NULL)

## possibly in another R session
load("testdump.rda")
debugger(testdump)
Available environments had calls:
1: f()
2: g()
3: stop("test dump.frames")

Enter an environment number, or 0 to exit
Selection: 1
Browsing in the environment with call:
f()
Called from: debugger.look(ind)
Browse[1]> ls()
[1] "g"
Browse[1]> g
function() stop("test dump.frames")
<environment: 759818>
Browse[1]> 
Available environments had calls:
1: f()
2: g()
3: stop("test dump.frames")

Enter an environment number, or 0 to exit
Selection: 0

## A possible setting for non-interactive sessions
options(error=quote({dump.frames(to.file=TRUE); q()}))

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{demo}{Demonstrations of R Functionality}{demo}
\keyword{documentation}{demo}
\keyword{utilities}{demo}
%
\begin{Description}\relax
\code{demo} is a user-friendly interface to running some demonstration
\R{} scripts.  \code{demo()} gives the list of available topics.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
demo(topic, package = NULL, lib.loc = NULL,
     character.only = FALSE, verbose = getOption("verbose"),
     echo = TRUE, ask = getOption("demo.ask"))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{topic}] the topic which should be demonstrated, given as a
\LinkA{name}{name} or literal character string, or a character string,
depending on whether \code{character.only} is \code{FALSE} (default)
or \code{TRUE}.  If omitted, the list of available topics is
displayed.
\item[\code{package}] a character vector giving the packages to look into for
demos, or \code{NULL}.  By default, all packages in the search path
are used.
\item[\code{lib.loc}] a character vector of directory names of \R{} libraries,
or \code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.  If the default is used, the loaded
packages are searched before the libraries.
\item[\code{character.only}] logical; if \code{TRUE}, use \code{topic} as
character string.
\item[\code{verbose}] a logical.  If \code{TRUE}, additional diagnostics are
printed.
\item[\code{echo}] a logical.  If \code{TRUE}, show the \R{} input when sourcing.
\item[\code{ask}] a logical (or \code{"default"}) indicating if
\code{\LinkA{devAskNewPage}{devAskNewPage}(ask=TRUE)} should be called before
graphical output happens from the demo code.  The value
\code{"default"} (the factory-fresh default) means to ask if
\code{echo == TRUE} and the graphics device appears to be
interactive.  This parameter applies both to any currently opened
device and to any devices opened by the demo code.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If no topics are given, \code{demo} lists the available demos.  The
corresponding information is returned in an object of class
\code{"packageIQR"}.  The structure of this class is experimental.  In
earlier versions of R, an empty character vector was returned along
with listing available demos.
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{source}{source}} and \code{\LinkA{devAskNewPage}{devAskNewPage}} which 
are called by \code{demo}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
demo() # for attached packages

## All available demos:
demo(package = .packages(all.available = TRUE))

## Display a demo, pausing between pages
demo(lm.glm, package="stats", ask=TRUE)

## Display it without pausing
demo(lm.glm, package="stats", ask=FALSE)

## Not run: 
 ch <- "scoping"
 demo(ch, character = TRUE)

## End(Not run)

## Find the location of a demo
system.file("demo", "lm.glm.R", package="stats")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{download.file}{Download File from the Internet}{download.file}
\keyword{utilities}{download.file}
%
\begin{Description}\relax
This function can be used to download a file from the Internet.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
download.file(url, destfile, method, quiet = FALSE, mode = "w",
              cacheOK = TRUE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{url}] A character string naming the URL of a resource to be
downloaded.

\item[\code{destfile}] A character string with the name where the downloaded
file is saved.  Tilde-expansion is performed.

\item[\code{method}] Method to be used for downloading files.  Currently
download methods \code{"internal"}, \code{"wget"} and \code{"lynx"}
are available, and there is a value \code{"auto"}: see
`Details'.  The method can also be set through the option
\code{"download.file.method"}: see \code{\LinkA{options}{options}()}.

\item[\code{quiet}] If \code{TRUE}, suppress status messages (if any), and
the progress bar.

\item[\code{mode}] character.  The mode with which to write the file. Useful
values are \code{"w"}, \code{"wb"} (binary), \code{"a"} (append) and
\code{"ab"}.  Only used for the \code{"internal"} method.


\item[\code{cacheOK}] logical.  Is a server-side cached value acceptable?
Implemented for the \code{"internal"} and \code{"wget"} methods.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The function \code{download.file} can be used to download a single
file as described by \code{url} from the internet and store it in
\code{destfile}.
The \code{url} must start with a scheme such as
\samp{http://}, \samp{ftp://} or \samp{file://}.

If \code{method = "auto"} is chosen (the default), the internal method
is chosen for \samp{file://} URLs, and for the others provided
\code{\LinkA{capabilities}{capabilities}("http/ftp")} is true (which it almost always
is).  Otherwise methods \code{"wget"} and \code{"lynx"} are tried in turn.

\code{cacheOK = FALSE} is useful for \samp{http://} URLs, and will
attempt to get a copy directly from the site rather than from an
intermediate cache.  (Not all platforms support it.)
It is used by \code{\LinkA{available.packages}{available.packages}}.

The remaining details apply to method \code{"internal"} only.

Note that \samp{https://} connections are
not supported.

See \code{\LinkA{url}{url}} for how \samp{file://} URLs are interpreted,
especially on Windows.  This function does decode encoded URLs.

The timeout for many parts of the transfer can be set by the option
\code{timeout} which defaults to 60 seconds.

The level of detail provided during transfer can be set by the
\code{quiet} argument and the \code{internet.info} option.  The
details depend on the platform and scheme, but setting
\code{internet.info} to 0 gives all available details, including
all server responses. Using 2 (the default) gives only serious
messages, and 3 or more suppresses all messages.

A progress bar tracks the transfer. If the file length is known, an
equals represents 2\% of the transfer completed: otherwise a dot
represents 10Kb.

Method \code{"wget"} can be used with proxy firewalls which require
user/password authentication if proper values are stored in the
configuration file for \code{wget}.
\end{Details}
%
\begin{Value}
An (invisible) integer code, \code{0} for success and non-zero for
failure.  For the \code{"wget"} and \code{"lynx"} methods this is the
status code returned by the external program.  The \code{"internal"}
method can return \code{1}, but will in most cases throw an error.
\end{Value}
%
\begin{Section}{Setting Proxies}
This applies to the internal code only.

Proxies can be specified via environment variables.
Setting \env{"no\_proxy"} to \code{"*"} stops any proxy being tried.
Otherwise the setting of \env{"http\_proxy"} or \env{"ftp\_proxy"}
(or failing that, the all upper-case version) is consulted and if
non-empty used as a proxy site.  For FTP transfers, the username
and password on the proxy can be specified by \env{"ftp\_proxy\_user"}
and \env{"ftp\_proxy\_password"}.  The form of \env{"http\_proxy"}
should be \code{"http://proxy.dom.com/"} or
\code{"http://proxy.dom.com:8080/"} where the port defaults to
\code{80} and the trailing slash may be omitted. For
\env{"ftp\_proxy"} use the form \code{"ftp://proxy.dom.com:3128/"}
where the default port is \code{21}.  These environment variables
must be set before the download code is first used: they cannot be
altered later by calling \code{\LinkA{Sys.setenv}{Sys.setenv}}.

Usernames and passwords can be set for HTTP proxy transfers via
environment variable \env{http\_proxy\_user} in the form
\code{user:passwd}.  Alternatively, \env{http\_proxy} can be of the
form \code{"http://user:pass@proxy.dom.com:8080/"} for compatibility
with \code{wget}.  Only the HTTP/1.0 basic authentication scheme is
supported.
\end{Section}
%
\begin{Note}\relax
Methods \code{"wget"} and \code{"lynx"} are for historical
compatibility.  They will block all other activity on the \R{} process.

For methods \code{"wget"} and \code{"lynx"} a system call is made to
the tool given by \code{method}, and the respective program must be
installed on your system and be in the search path for executables.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{options}{options}} to set the \code{HTTPUserAgent}, \code{timeout}
and \code{internet.info} options.

\code{\LinkA{url}{url}} for a finer-grained way to read data from URLs.

\code{\LinkA{url.show}{url.show}}, \code{\LinkA{available.packages}{available.packages}},
\code{\LinkA{download.packages}{download.packages}} for applications.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{edit}{Invoke a Text Editor}{edit}
\methaliasA{edit.default}{edit}{edit.default}
\aliasA{emacs}{edit}{emacs}
\aliasA{pico}{edit}{pico}
\aliasA{vi}{edit}{vi}
\aliasA{xedit}{edit}{xedit}
\aliasA{xemacs}{edit}{xemacs}
\keyword{utilities}{edit}
%
\begin{Description}\relax
Invoke a text editor on an \R{} object.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
## Default S3 method:
edit(name = NULL, file = "", title = NULL,
     editor = getOption("editor"), ...)

vi(name = NULL, file = "")
emacs(name = NULL, file = "")
pico(name = NULL, file = "")
xemacs(name = NULL, file = "")
xedit(name = NULL, file = "")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{name}] a named object that you want to edit. If name is missing
then the file specified by \code{file} is opened for editing.
\item[\code{file}] a string naming the file to write the edited version to.
\item[\code{title}] a display name for the object being edited.
\item[\code{editor}] a string naming the text editor you want to use.  On
Unix the default is set from the environment variables \env{EDITOR}
or \env{VISUAL} if either is set, otherwise \code{vi} is used.  On
Windows it defaults to \code{notepad}.
\item[\code{...}] further arguments to be passed to or from methods.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\code{edit} invokes the text editor specified by \code{editor} with
the object \code{name} to be edited.  It is a generic function,
currently with a default method and one for data frames and matrices.

\code{data.entry} can be used to edit data, and is used by \code{edit}
to edit matrices and data frames on systems for which
\code{data.entry} is available.

It is important to realize that \code{edit} does not change the object
called \code{name}. Instead, a copy of name is made and it is that
copy which is changed.  Should you want the changes to apply to the
object \code{name} you must assign the result of \code{edit} to
\code{name}.  (Try \code{\LinkA{fix}{fix}} if you want to make permanent
changes to an object.)

In the form \code{edit(name)},
\code{edit} deparses \code{name} into a temporary file and invokes the
editor \code{editor} on this file. Quitting from the editor causes
\code{file} to be parsed and that value returned.
Should an error occur in parsing, possibly due to incorrect syntax, no
value is returned. Calling \code{edit()}, with no arguments, will
result in the temporary file being reopened for further editing.

Note that deparsing is not perfect, and the object recreated after
editing can differ in subtle ways from that deparsed: see
\code{\LinkA{dput}{dput}} and \code{\LinkA{.deparseOpts}{.deparseOpts}}. (The deparse options
used are the same as the defaults for \code{dump}.)  Editing a
function will preserve its environment.  See
\code{\LinkA{edit.data.frame}{edit.data.frame}} for further changes that can occur when
editing a data frame or matrix.

Currently only the internal editor in Windows makes use of the 
\code{title} option; it displays the given name in the window 
header.
\end{Details}
%
\begin{Note}\relax
The functions \code{vi}, \code{emacs}, \code{pico}, \code{xemacs},
\code{xedit} rely on the corresponding editor being available and
being on the path. This is system-dependent.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{edit.data.frame}{edit.data.frame}},
\code{\LinkA{data.entry}{data.entry}},
\code{\LinkA{fix}{fix}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
# use xedit on the function mean and assign the changes
mean <- edit(mean, editor = "xedit")

# use vi on mean and write the result to file mean.out
vi(mean, file = "mean.out")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{edit.data.frame}{Edit Data Frames and Matrices}{edit.data.frame}
\aliasA{edit.matrix}{edit.data.frame}{edit.matrix}
\keyword{utilities}{edit.data.frame}
%
\begin{Description}\relax
Use data editor on data frame or matrix contents.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
## S3 method for class 'data.frame':
edit(name, factor.mode = c("character", "numeric"),
     edit.row.names = any(row.names(name) != 1:nrow(name)), ...)

## S3 method for class 'matrix':
edit(name, edit.row.names = !is.null(dn[[1]]), ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{name}] A data frame or (numeric, logical or character) matrix.
\item[\code{factor.mode}] How to handle factors (as integers or using
character levels) in a data frame.
\item[\code{edit.row.names}] logical. Show the row names (if they exist) be
displayed as a separate editable column?  It is an error to ask for
this on a matrix with \code{NULL} row names.
\item[\code{...}] further arguments passed to or from other methods.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
At present, this only works on simple data frames containing numeric,
logical or character vectors and factors, and numeric, logical or
character matrices.  Any other mode of matrix will give an error, and
a warning is given when the matrix has a class (which will be discarded).

Data frame columns are coerced on input to \emph{character} unless
numeric (in the sense of \code{is.numeric}), logical or factor.  A
warning is given when classes are discarded.  Special characters
(tabs, non-printing ASCII, etc.) will be displayed as escape sequences.

Factors columns are represented in the spreadsheet as either numeric
vectors (which are more suitable for data entry) or character vectors
(better for browsing). After editing, vectors are padded with
\code{NA} to have the same length and factor attributes are restored.
The set of factor levels can not be changed by editing in numeric
mode; invalid levels are changed to \code{NA} and a warning is issued.
If new factor levels are introduced in character mode, they are added
at the end of the list of levels in the order in which they
encountered.

It is possible to use the data-editor's facilities to select the mode
of columns to swap between numerical and factor columns in a data
frame.  Changing any column in a numerical matrix to character will
cause the result to be coerced to a character matrix.  Changing 
the mode of logical columns is not supported.

For a data frame, the row names will be taken from the original object
if \code{edit.row.names = FALSE} and the number of rows is unchanged,
and from the edited output if \code{edit.row.names = TRUE} and there
are no duplicates.  (If the \code{row.names} column is incomplete, it
is extended by entries like \code{row223}.)  In all other cases the
row names are replaced by \code{seq(length=nrows)}.

For a matrix, colnames will be added (of the form \code{col7}) if
needed.  The rownames will be taken from the original object if
\code{edit.row.names = FALSE} and the number of rows is unchanged
(otherwise \code{NULL}), and from the edited output if
\code{edit.row.names = TRUE}.  (If the \code{row.names} column is
incomplete, it is extended by entries like \code{row223}.)

Editing a matrix or data frame will lose all attributes apart from the
row and column names.
\end{Details}
%
\begin{Value}
The edited data frame or matrix.
\end{Value}
%
\begin{Note}\relax
\code{fix(dataframe)} works for in-place editing by calling this
function.

If the data editor is not available, a dump of the object is presented
for editing using the default method of \code{edit}.

At present the data editor is limited to 65535 rows.
\end{Note}
%
\begin{Author}\relax
 Peter Dalgaard 
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{data.entry}{data.entry}}, \code{\LinkA{edit}{edit}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
edit(InsectSprays)
edit(InsectSprays, factor.mode="numeric")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{example}{Run an Examples Section from the Online Help}{example}
\keyword{documentation}{example}
\keyword{utilities}{example}
%
\begin{Description}\relax
Run all the \R{} code from the \bold{Examples} part of \R{}'s online help
topic \code{topic} with two possible exceptions, \code{dontrun} and
\code{dontshow}, see `Details' below.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
example(topic, package = NULL, lib.loc = NULL,
        local = FALSE, echo = TRUE,
        verbose = getOption("verbose"),
        setRNG = FALSE, ask = getOption("example.ask"),
        prompt.prefix = abbreviate(topic, 6))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{topic}] name or literal character string: the online
\code{\LinkA{help}{help}} topic the examples of which should be run.
\item[\code{package}] a character vector giving the package names to look
into for the topic, or \code{NULL} (the default), when all packages on
the \LinkA{search}{search} path are used.
\item[\code{lib.loc}] a character vector of directory names of \R{} libraries,
or \code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.  If the default is used, the loaded
packages are searched before the libraries.
\item[\code{local}] logical: if \code{TRUE} evaluate locally, if \code{FALSE}
evaluate in the workspace.
\item[\code{echo}] logical;  if \code{TRUE}, show the \R{} input when sourcing.
\item[\code{verbose}] logical;  if \code{TRUE}, show even more when running
example code.
\item[\code{setRNG}] logical or expression;  if not \code{FALSE}, the random
number generator state is saved, then initialized to a specified state,
the example is run and the (saved) state is restored.
\code{setRNG = TRUE} sets the same state as
\code{R CMD \LinkA{check}{check}} does for 
running a package's examples.  This is currently equivalent to
\code{setRNG = \{RNGkind("default", "default"); set.seed(1)\}}.
\item[\code{ask}] logical (or \code{"default"}) indicating if
\code{\LinkA{devAskNewPage}{devAskNewPage}(ask=TRUE)} should be called
before graphical output happens from the example code.  The value
\code{"default"} (the factory-fresh default) means to ask if
\code{echo == TRUE} and the graphics device appears to be
interactive.  This parameter applies both to any currently opened
device and to any devices opened by the example code.   
\item[\code{prompt.prefix}] character; prefixes the prompt to be used if
\code{echo = TRUE}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If \code{lib.loc} is not specified, the packages are searched for
amongst those already loaded, then in the libraries given by
\code{\LinkA{.libPaths}{.libPaths}()}.  If \code{lib.loc} is specified, packages
are searched for only in the specified libraries, even if they are
already loaded from another library.  The search stops at the first
package found that has help on the topic.

An attempt is made to load the package before running the examples,
but this will not replace a package loaded from another location.

If \code{local = TRUE} objects are not created in the workspace and so
not available for examination after \code{example} completes: on the
other hand they cannot overwrite objects of the same name in the
workspace.

As detailed in the manual \emph{Writing \R{} Extensions}, the author of
the help page can markup parts of the examples for two exception rules
\begin{description}

\item[\code{dontrun}] encloses code that should not be run.
\item[\code{dontshow}] encloses code that is invisible on help
pages, but will be run both by the package checking tools,
and the \code{example()} function.  This was previously
\code{testonly}, and that form is still accepted.

\end{description}


If the examples file contains non-ASCII characters the encoding used
will matter.  If in a UTF-8 locale \code{example} first tries UTF-8
and then Latin-1.  (This can be overridden by setting the encoding in
the \file{.Rd} file.)
\end{Details}
%
\begin{Value}
The value of the last evaluated expression.
\end{Value}
%
\begin{Note}\relax
The examples can be many small files.  On some file systems it is
desirable to save space, and the files in the \file{R-ex} directory
of an installed package can be zipped up as a zip archive
\file{Rex.zip}.
\end{Note}
%
\begin{Author}\relax
Martin Maechler and others
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{demo}{demo}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
example(InsectSprays)
## force use of the standard package 'stats':
example("smooth", package="stats", lib.loc=.Library)

## set RNG *before* example as when R CMD check is run:

r1 <- example(quantile, setRNG = TRUE)
x1 <- rnorm(1)
u <- runif(1)
## identical random numbers
r2 <- example(quantile, setRNG = TRUE)
x2 <- rnorm(1)
stopifnot(identical(r1, r2))
## but x1 and x2 differ since the RNG state from before example()
## differs and is restored!
x1; x2
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{file.edit}{Edit One or More Files}{file.edit}
\keyword{utilities}{file.edit}
%
\begin{Description}\relax
Edit one or more files in a text editor.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
file.edit(..., title = file, editor = getOption("editor"))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{...}] one or more character vectors containing the names of the
files to be edited.
\item[\code{title}] the title to use in the editor; defaults to the filename. 
\item[\code{editor}] the text editor to be used.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Path expansion (see \code{\LinkA{path.expand}{path.expand}} will be done on
\code{names}.

The behaviour of this function is very system dependent.  Currently
files can be opened only one at a time on Unix; on Windows, the
internal editor allows multiple files to be opened, but has a limit of
50 simultaneous edit windows.

The \code{title} argument is used for the window caption in Windows,
and is ignored on other platforms.
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{files}{files}},
\code{\LinkA{file.show}{file.show}},
\code{\LinkA{edit}{edit}},
\code{\LinkA{fix}{fix}},
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
# open two R scripts for editing
file.edit("script1.R", "script2.R")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{file\_test}{Shell-style Tests on Files}{file.Rul.test}
\keyword{file}{file\_test}
%
\begin{Description}\relax
Utility for shell-style file tests.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
file_test(op, x, y)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{op}] a character string specifying the test to be performed.
Unary tests (only \code{x} is used) are \code{"-f"} (existence and
not being a directory), \code{"-d"} (existence and directory) and
\code{"-x"} (executable as a file or searchable as a directory). 
Binary tests are \code{"-nt"} (strictly newer than, using the modification
dates) and \code{"-ot"} (strictly older than): in both cases the
test is false unless both files exist.
\item[\code{x,y}] character vectors giving file paths.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
`Existence' here means being on the file system and accessible
by the \code{stat} system call (or a 64-bit extension) -- on a
Unix-alike this requires execute permission on all of the directories in
the path that leads to the file, but no permissions on the file
itself.
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{file.exists}{file.exists}} which only tests for existence
(\code{test -e} on some systems) but not for not being a directory.

\code{\LinkA{file.path}{file.path}}, \code{\LinkA{file.info}{file.info}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
dir <- file.path(R.home(), "library", "stats")
file_test("-d", dir)
file_test("-nt", file.path(dir, "R"), file.path(dir, "demo"))
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{findLineNum}{Find the Location of a Line of Source Code, or Set a Breakpoint There.}{findLineNum}
\aliasA{setBreakpoint}{findLineNum}{setBreakpoint}
\keyword{debugging}{findLineNum}
%
\begin{Description}\relax
These functions locate objects containing particular lines of source
code, using the information saved when the code was parsed with
\code{options(keep.source = TRUE)}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
findLineNum(srcfile, line, nameonly = TRUE, envir = parent.frame(),
            lastenv)

setBreakpoint(srcfile, line, nameonly = TRUE, envir = parent.frame(),
              lastenv, verbose = TRUE, tracer, print = FALSE, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{srcfile}] The name of the file containing the source code.
\item[\code{line}] The line number within the file.  See Details for an
alternate way to specify this.
\item[\code{nameonly}] If \code{TRUE} (the default), we require only a match
to \code{basename(srcfile)}, not to the full path.
\item[\code{envir}] Where do we start looking for function objects?
\item[\code{lastenv}] Where do we stop?  See the Details.
\item[\code{verbose}] Should we print information on where breakpoints were set?
\item[\code{tracer}] An optional \code{tracer} function to pass to
\code{\LinkA{trace}{trace}}.  By default, a call to \code{\LinkA{browser}{browser}}
is inserted.
\item[\code{print}] The \code{print} argument to pass to \code{\LinkA{trace}{trace}}.
\item[\code{...}] Additional arguments to pass to \code{\LinkA{trace}{trace}}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The \code{findLineNum} function searches through all objects in
environment \code{envir}, it's parent, grandparent, etc., all the way
back to \code{lastenv}.

\code{lastenv} defaults to the global environment if
\code{envir} is not specified, and to the
root environment \code{\LinkA{emptyenv}{emptyenv}()} if \code{envir} is
specified.  (The first default tends to be quite fast, and will
usually find all user code other than S4 methods; the second one is
quite slow, as it will typically search all attached system
libraries.)

\code{setBreakpoint} is a simple wrapper function for
\code{\LinkA{trace}{trace}}.  It will set breakpoints at the locations found
by \code{findLineNum}.

The \code{srcfile} is normally a filename entered as a character
string, but it may be a \code{"\LinkA{srcfile}{srcfile}"} object, or it may
include a suffix like \code{"filename.R\#nn"}, in which case the number
\code{nn} will be used as a default value for \code{line}.

As described in the description of the \code{where} argument on the
man page for \code{\LinkA{trace}{trace}}, the \R{} package system uses a
complicated scheme that may include more than one copy of a function
in a package.  The user will typically see the public one on the
search path, while code in the package will see a private one in the
package NAMESPACE.  If you set \code{envir} to the environment of a
function in the package, by default \code{findLineNum} will find both
versions, and \code{setBreakpoint} will set the breakpoint in both.
(This can be controlled using \code{lastenv}; e.g. 
\code{envir=environment(foo)}, \code{lastenv=globalenv()} will find only the
private copy, as the search is stopped before seeing the public
copy.

S version 4 methods are also somewhat tricky to find.  They are stored
with the generic function, which may be in the \pkg{base} or other
package, so it is usually necessary to have \code{lastenv=emptyenv()}
in order to find them.  In some cases transformations are done by \R{}
when storing them and \code{findLineNum} may not be able to find the
original code.  Many special cases, e.g. methods on primitive
generics, are not yet supported.
\end{Details}
%
\begin{Value}
\code{fineLineNum} returns a list of objects containing location
information.  A \code{print} method is defined for them.

\code{setBreakpoint} has no useful return value; it is called for the
side effect of calling \code{\LinkA{trace}{trace}}. 
\end{Value}
%
\begin{Author}\relax
Duncan Murdoch
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{trace}{trace}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
# Find what function was defined in the file mysource.R at line 100:
findLineNum("mysource.R#100")

# Set a breakpoint in both copies of that function, assuming one is in the
# same namespace as myfunction and the other is on the search path
setBreakpoint("mysource.R#100", envir=environment(myfunction))

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{fix}{Fix an Object}{fix}
\keyword{utilities}{fix}
%
\begin{Description}\relax
\code{fix} invokes \code{\LinkA{edit}{edit}} on \code{x} and then assigns the new
(edited) version of \code{x} in the user's workspace.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
fix(x, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] the name of an \R{} object, as a name or a character string.
\item[\code{...}] arguments to pass to editor: see \code{\LinkA{edit}{edit}}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The name supplied as \code{x} need not exist as an \R{} object, in 
which case a function with no arguments and an empty body is supplied 
for editing.

Editing an \R{} object may change it in ways other than are obvious: see
the comment under \code{\LinkA{edit}{edit}}.  See
\code{\LinkA{edit.data.frame}{edit.data.frame}} for changes that can occur when editing
a data frame or matrix.
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{edit}{edit}},
\code{\LinkA{edit.data.frame}{edit.data.frame}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
 ## Assume 'my.fun' is a user defined function :
 fix(my.fun)
 ## now my.fun is changed
 ## Also,
 fix(my.data.frame) # calls up data editor
 fix(my.data.frame, factor.mode="char") # use of ...

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{flush.console}{Flush Output to A Console}{flush.console}
\keyword{utilities}{flush.console}
%
\begin{Description}\relax
This does nothing except on console-based versions of \R{}.
On the Mac OS X and Windows GUIs, it ensures that the display of
output in the console is current, even if output buffering is on.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
flush.console()
\end{verbatim}
\end{Usage}
\inputencoding{latin1}
\HeaderA{format}{Format Unordered and Ordered Lists}{format}
\aliasA{formatOL}{format}{formatOL}
\aliasA{formatUL}{format}{formatUL}
\keyword{print}{format}
%
\begin{Description}\relax
Format unordered (itemize) and ordered (enumerate) lists.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
formatUL(x, label = "*", offset = 0,
         width = 0.9 * getOption("width"))
formatOL(x, type = "arabic", offset = 0, start = 1,
         width = 0.9 * getOption("width"))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] a character vector of list items.
\item[\code{label}] a character string used for labelling the items.
\item[\code{offset}] a non-negative integer giving the offset (indentation)
of the list.
\item[\code{width}] a positive integer giving the target column for wrapping
lines in the output.
\item[\code{type}] a character string specifying the `type' of the
labels in the ordered list.  If \code{"arabic"} (default), arabic
numerals are used.  For \code{"Alph"} or \code{"alph"}, single upper
or lower case letters are employed (in this case, the number of the
last item must not exceed 26.  Finally, for \code{"Roman"} or
\code{"roman"}, the labels are given as upper or lower case roman
numerals (with the number of the last item maximally 3899).
\code{type} can be given as a unique abbreviation of the above, or
as one of the \acronym{HTML} style tokens \code{"1"} (arabic),
\code{"A"}/\code{"a"} (alphabetic), or \code{"I"}/\code{"i"}
(roman), respectively.
\item[\code{start}] a positive integer specifying the starting number of the
first item in an ordered list.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
A character vector with the formatted entries.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{formatDL}{formatDL}} for formatting description lists.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## A simpler recipe.
x <- c("Mix dry ingredients thoroughly.",
       "Pour in wet ingredients.",
       "Mix for 10 minutes.",
       "Bake for one hour at 300 degrees.")
## Format and output as an unordered list.
writeLines(formatUL(x))
## Format and output as an ordered list.
writeLines(formatOL(x))
## Ordered list using lower case roman numerals.
writeLines(formatOL(x, type = "i"))
## Ordered list using upper case letters and some offset.
writeLines(formatOL(x, type = "A", offset = 5))
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{getAnywhere}{Retrieve an R Object, Including from a Name Space}{getAnywhere}
\aliasA{argsAnywhere}{getAnywhere}{argsAnywhere}
\aliasA{print.getAnywhere}{getAnywhere}{print.getAnywhere}
\aliasA{[.getAnywhere}{getAnywhere}{[.getAnywhere}
\keyword{data}{getAnywhere}
%
\begin{Description}\relax
These functions locate all objects with name matching their argument,
whether visible on the search path, registered as an S3 method or in a
name space but not exported. \code{getAnywhere()} returns the objects
and \code{argsAnywhere()} returns the arguments of any objects that
are functions.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
getAnywhere(x)
argsAnywhere(x)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] a character string or name.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The functions look at all loaded name spaces, whether or not they are
associated with a package on the search list.  

The functions do not search literally ``anywhere'': for
example, local evaluation frames and namespaces that are not loaded
will not be searched.

Where functions are found as an S3 method, an attempt is made to find
which name space registered them.  This may not be correct, especially
if a name space is unloaded.
\end{Details}
%
\begin{Value}
For \code{getAnywhere()} an object of class \code{"getAnywhere"}.  This is a list with components
\begin{ldescription}
\item[\code{name}] the name searched for.
\item[\code{objs}] a list of objects found
\item[\code{where}] a character vector explaining where the object(s) were found
\item[\code{visible}] logical: is the object visible
\item[\code{dups}] logical: is the object identical to one earlier in the
list.

\end{ldescription}
Normally the structure will be hidden by the \code{print} method.
There is a \code{[} method to extract one or more of the objects
found.

For \code{argsAnywhere()} one or more argument lists as returned by \code{\LinkA{args}{args}}.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{get}{get}},  \code{\LinkA{getFromNamespace}{getFromNamespace}}, \code{\LinkA{args}{args}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
getAnywhere("format.dist")
getAnywhere("simpleLoess") # not exported from stats
argsAnywhere(format.dist)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{getFromNamespace}{Utility functions for Developing Namespaces}{getFromNamespace}
\aliasA{assignInNamespace}{getFromNamespace}{assignInNamespace}
\aliasA{fixInNamespace}{getFromNamespace}{fixInNamespace}
\keyword{data}{getFromNamespace}
%
\begin{Description}\relax
Utility functions to access and replace the non-exported functions in
a name space, for use in developing packages with name spaces.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
getFromNamespace(x, ns, pos = -1, envir = as.environment(pos))

assignInNamespace(x, value, ns, pos = -1,
                  envir = as.environment(pos))

fixInNamespace(x, ns, pos = -1, envir = as.environment(pos), ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] an object name (given as a character string).
\item[\code{value}] an \R{} object.
\item[\code{ns}] a name space, or character string giving the name space.
\item[\code{pos}] where to look for the object: see \code{\LinkA{get}{get}}.
\item[\code{envir}] an alternative way to specify an environment to look in.
\item[\code{...}] arguments to pass to the editor: see \code{\LinkA{edit}{edit}}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The name space can be specified in several ways.  Using, for example,
\code{ns = "stats"} is the most direct, but a loaded package with a
name space can be specified via any of the methods used for
\code{\LinkA{get}{get}}: \code{ns} can also be the environment printed as
\code{<namespace:foo>}.

\code{getFromNamespace} is similar to (but predates) the
\code{\LinkA{:::}{:::}} operator, but is more flexible in how the name space
is specified.

\code{fixInNamespace} invokes \code{\LinkA{edit}{edit}} on the object named
\code{x} and assigns the revised object in place of the original
object.  For compatibility with \code{fix}, \code{x} can be unquoted.
\end{Details}
%
\begin{Value}
\code{getFromNamespace} returns the object found (or gives an error).

\code{assignInNamespace} and \code{fixInNamespace} are invoked for
their side effect of changing the object in the name space.
\end{Value}
%
\begin{Note}\relax
\code{assignInNamespace} and \code{fixInNamespace} change the copy in
the name space, but not any copies already exported from the name space,
in particular an object of that name in the package (if already
attached) and any copies already imported into other name spaces.
They are really intended to be used \emph{only} for objects which are
not exported from the name space.  They do attempt to alter a copy
registered as an S3 method if one is found.

They can only be used to change the values of objects in the
name space, not to create new objects.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{get}{get}}, \code{\LinkA{fix}{fix}}, \code{\LinkA{getS3method}{getS3method}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
getFromNamespace("findGeneric", "utils")
## Not run: 
fixInNamespace("predict.ppr", "stats")
stats:::predict.ppr
getS3method("predict", "ppr")
## alternatively
fixInNamespace("predict.ppr", pos = 3)
fixInNamespace("predict.ppr", pos = "package:stats")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{getS3method}{Get An S3 Method}{getS3method}
\keyword{data}{getS3method}
%
\begin{Description}\relax
Get a method for an S3 generic, possibly from a name space. 
\end{Description}
%
\begin{Usage}
\begin{verbatim}
getS3method(f, class, optional = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{f}] character: name of the generic.
\item[\code{class}] character: name of the class.
\item[\code{optional}] logical: should failure to find the generic or a
method be allowed?
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
S3 methods may be hidden in packages with name spaces, and will not
then be found by \code{\LinkA{get}{get}}: this function can retrieve
such functions, primarily for debugging purposes.
\end{Details}
%
\begin{Value}
The function found, or \code{NULL} if no function is found and
\code{optional = TRUE}.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{methods}{methods}}, \code{\LinkA{get}{get}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(stats)
exists("predict.ppr") # false
getS3method("predict", "ppr")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{glob2rx}{Change Wildcard or Globbing Pattern into Regular Expression}{glob2rx}
\keyword{file}{glob2rx}
\keyword{character}{glob2rx}
\keyword{utilities}{glob2rx}
%
\begin{Description}\relax
Change \emph{wildcard} aka \emph{globbing} patterns into the
corresponding regular expressions (\code{\LinkA{regexp}{regexp}}).
\end{Description}
%
\begin{Usage}
\begin{verbatim}
glob2rx(pattern, trim.head = FALSE, trim.tail = TRUE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{pattern}] character vector
\item[\code{trim.head}] logical specifying if leading \code{"\textasciicircum{}.*"} should be
trimmed from the result.
\item[\code{trim.tail}] logical specifying if trailing \code{".*\$"} should be
trimmed from the result.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This takes a wildcard as used by most shells and returns an equivalent
regular expression.  \code{?} is mapped to \code{.} (match a single
character), \code{*} to \code{.*} (match any string, including an
empty one), and the pattern is anchored (it must start at the
beginning and end at the end).  Optionally, the resulting regexp is
simplified.

Note that now even \code{(}, \code{[} and \code{\{} can be used
in \code{pattern}, but \code{glob2rx()} may not work correctly with
arbitrary characters in \code{pattern}.
\end{Details}
%
\begin{Value}
A character vector of the same length as the input \code{pattern}
where each wildcard is translated to the corresponding
regular expression.
\end{Value}
%
\begin{Author}\relax
Martin Maechler, Unix/sed based version, 1991; current: 2004
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{regexp}{regexp}} about regular expression,
\code{\LinkA{sub}{sub}}, etc about substitutions using regexps.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
stopifnot(glob2rx("abc.*") == "^abc\\.",
          glob2rx("a?b.*") == "^a.b\\.",
          glob2rx("a?b.*", trim.tail=FALSE) == "^a.b\\..*$",
          glob2rx("*.doc") == "^.*\\.doc$",
          glob2rx("*.doc", trim.head=TRUE) == "\\.doc$",
          glob2rx("*.t*")  == "^.*\\.t",
          glob2rx("*.t??") == "^.*\\.t..$",
          glob2rx("*[*")  == "^.*\\["
)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{head}{Return the First or Last Part of an Object}{head}
\methaliasA{head.data.frame}{head}{head.data.frame}
\methaliasA{head.default}{head}{head.default}
\methaliasA{head.ftable}{head}{head.ftable}
\methaliasA{head.function}{head}{head.function}
\methaliasA{head.matrix}{head}{head.matrix}
\methaliasA{head.table}{head}{head.table}
\aliasA{tail}{head}{tail}
\methaliasA{tail.data.frame}{head}{tail.data.frame}
\methaliasA{tail.default}{head}{tail.default}
\methaliasA{tail.ftable}{head}{tail.ftable}
\methaliasA{tail.function}{head}{tail.function}
\methaliasA{tail.matrix}{head}{tail.matrix}
\methaliasA{tail.table}{head}{tail.table}
\keyword{manip}{head}
%
\begin{Description}\relax
Returns the first or last parts of a vector, matrix, table, data frame
or function.  Since \code{head()} and \code{tail()} are generic
functions, they may also have been extended to other classes.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
head(x, ...)
## Default S3 method:
head(x, n = 6L, ...)
## S3 method for class 'data.frame':
head(x, n = 6L, ...)
## S3 method for class 'matrix':
head(x, n = 6L, ...)
## S3 method for class 'ftable':
head(x, n = 6L, ...)
## S3 method for class 'table':
head(x, n = 6L, ...)
## S3 method for class 'function':
head(x, n = 6L, ...)

tail(x, ...)
## Default S3 method:
tail(x, n = 6L, ...)
## S3 method for class 'data.frame':
tail(x, n = 6L, ...)
## S3 method for class 'matrix':
tail(x, n = 6L, addrownums = TRUE, ...)
## S3 method for class 'ftable':
tail(x, n = 6L, addrownums = FALSE, ...)
## S3 method for class 'table':
tail(x, n = 6L, addrownums = TRUE, ...)
## S3 method for class 'function':
tail(x, n = 6L, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] an object
\item[\code{n}] a single integer. If positive, size for the resulting
object: number of elements for a vector (including lists), rows for
a matrix or data frame or lines for a function. If negative, all but
the \code{n} last/first number of elements of \code{x}.
\item[\code{addrownums}] if there are no row names, create them from the row
numbers.
\item[\code{...}] arguments to be passed to or from other methods.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
For matrices, 2-dim tables and data frames, \code{head()} (\code{tail()}) returns
the first (last) \code{n} rows when \code{n > 0} or all but the
last (first) \code{n} rows when \code{n < 0}.  \code{head.matrix()} and
\code{tail.matrix()} are exported.  For functions, the
lines of the deparsed function are returned as character strings.

If a matrix has no row names, then \code{tail()} will add row names of
the form \code{"[n,]"} to the result, so that it looks similar to the
last lines of \code{x} when printed.  Setting \code{addrownums =
    FALSE} suppresses this behaviour.
\end{Details}
%
\begin{Value}
An object (usually) like \code{x} but generally smaller.  For
\code{\LinkA{ftable}{ftable}} objects \code{x}, a transformed \code{format(x)}.
\end{Value}
%
\begin{Author}\relax
Patrick Burns, improved and corrected by R-Core. Negative argument
added by Vincent Goulet.
\end{Author}
%
\begin{Examples}
\begin{ExampleCode}
head(letters)
head(letters, n = -6L)

head(freeny.x, n = 10L)
head(freeny.y)

tail(letters)
tail(letters, n = -6L)

tail(freeny.x)
tail(freeny.y)

tail(library)

head(stats::ftable(Titanic))
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{help}{Documentation}{help}
\keyword{documentation}{help}
%
\begin{Description}\relax
\code{help} is the primary interface to the help systems.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
help(topic, package = NULL, lib.loc = NULL,
     verbose = getOption("verbose"),
     try.all.packages = getOption("help.try.all.packages"),
     help_type = getOption("help_type"),
     chmhelp = getOption("chmhelp"),
     htmlhelp = getOption("htmlhelp"),
     offline = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{topic}] usually, a \LinkA{name}{name} or character string specifying the
topic for which help is sought.  A character string (enclosed in
explicit single or double quotes) is always taken as naming a topic.

If the value of \code{topic} is a length-one
character vector the topic is taken to be the value of the only
element.  Otherwise \code{topic} must be a name or a \LinkA{reserved}{reserved}
word (if syntactically valid) or character string.

See `Details' for what happens if this is omitted.

\item[\code{package}] a name or character vector giving the packages to look
into for documentation, or \code{NULL}. By default, all packages in
the search path are used. To avoid a name being deparsed use e.g.
\code{(pkg\_ref)}.
\item[\code{lib.loc}] a character vector of directory names of \R{} libraries,
or \code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.  If the default is used, the loaded
packages are searched before the libraries.
\item[\code{verbose}] logical; if \code{TRUE}, the file name is reported.
\item[\code{try.all.packages}] logical; see \code{Note}.
\item[\code{help\_type}] character string: the type of help required.
Possible values are \code{"text"}, \code{"html"},
\code{"postscript"}, \code{"ps"} and \code{"pdf"}.  Case is ignored,
and partial matching is allowed.

\item[\code{chmhelp}] a deprecated way to specify \code{help\_type = "text"}.
\item[\code{htmlhelp}] a deprecated way to specify \code{help\_type = "html"}.
\item[\code{offline}] a deprecated way to specify \code{help\_type = "postscript"}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The following types of help are available:
\begin{itemize}

\item Plain text help
\item HTML help pages with hyperlinks to other topics, shown in a
browser by \code{\LinkA{browseURL}{browseURL}}.
(Where possible an existing browser window is re-used: the Mac OS X
GUI uses its own browser window.)
If for some reason HTML help is unavailable (see
\code{\LinkA{startDynamicHelp}{startDynamicHelp}}), plain text help will be used
instead.
\item For \code{help} only, typeset as a PostScript or PDF file --
see the section on `Offline help'.

\end{itemize}

The `factory-fresh' default is text help except from the Mac OS
GUI, which uses HTML help displayed in its own browser window.

With packages installed under \R{} 2.10.0 or later help pages are
generated when needed from parsed Rd files.  Packages installed under
earlier versions of \R{} do not have parsed Rd files, but do (by
default) have prebuilt text, html and latex help pages, and these are
used if parsed Rd files are not available.

The rendering of text help will use directional quotes in suitable
locales (UTF-8 and single-byte Windows locales): sometimes the fonts
used do not support these quotes so this can be turned off by setting
\code{\LinkA{options}{options}(useFancyQuotes = FALSE)}.

\code{topic} is not optional: if it is omitted \R{} will give (text)
information on the package (including hints to suitable help topics)
if a package is specified, a (text) list of available packages if
\code{lib.loc} only is specified, and help on \code{help} itself if
none of the first three arguments is specified.

Some topics need to be quoted (by \LinkA{backtick}{backtick}s) or given as a
character string.  There include those which cannot syntactically
appear on their own such as unary and binary operators,
\code{function} and control-flow \LinkA{reserved}{reserved} words (including
\code{if}, \code{else} \code{for}, \code{in}, \code{repeat},
\code{while}, \code{break} and \code{next}.  The other \code{reserved}
words can be used as if they were names, for example \code{TRUE},
\code{NA} and \code{Inf}.

If multiple help files matching \code{topic} are found, in interactive
use a menu is presented for the user to choose one: in batch use the
first on the search path is used.  (For HTML help the menu will be an
HTML page, otherwise a graphical menu if possible if
\code{\LinkA{getOption}{getOption}("menu.graphics")} is true, the default.)
\end{Details}
%
\begin{Section}{Offline help}
Typeset documentation is produced by running the LaTeX version of the
help page through \command{latex} and \command{dvips} or, if
\code{help\_type = "PDF"}, \command{pdflatex}.  This will either
produce a PostScript or PDF file or (depending on the configuration of
\command{dvips}) send a PostScript file to a printer.

The appearance of the output can be customized through a file
\file{Rhelp.cfg} somewhere in your LaTeX search path: this will be
input as a LaTeX style file after \code{Rd.sty}.  Some 
\LinkA{environment variables}{environment variables} are consulted, notably \env{R\_PAPERSIZE}
(\emph{via} \code{getOption("papersize")}) and \env{R\_RD4DVI} /
\env{R\_RD4PDF} (see `Making manuals' in the
`R Installation and Administration Manual').

If there is a function \code{offline\_help\_helper} in the workspace or
further down the search path it is used to do the typesetting,
otherwise the function of that name in the \code{utils} name space
(to which the first paragraph applies).  It should have two 
arguments, the name of the LaTeX file to be typeset and the type.
\end{Section}
%
\begin{Note}\relax
Unless \code{lib.loc} is specified explicitly, the loaded packages are
searched before those in the specified libraries.  This ensures that
if a library is loaded from a library not in the known library trees,
then the help from the loaded library is used.  If \code{lib.loc} is
specified explicitly, the loaded packages are \emph{not} searched.

If this search fails and argument \code{try.all.packages} is
\code{TRUE} and neither \code{packages} nor \code{lib.loc} is
specified, then all the packages in the known library trees are
searched for help on \code{topic} and a list of (any) packages where
help may be found is displayed (with hyperlinks for \code{help\_type =
  "html"}).  \strong{NB:} searching all packages can be slow.
\end{Note}
%
\begin{References}\relax
Becker, R. A., Chambers, J. M. and Wilks, A. R. (1988)
\emph{The New S Language}.
Wadsworth \& Brooks/Cole.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{?}{Question}} for shortcuts to help topics.

\code{\LinkA{help.search}{help.search}()} or \code{\LinkA{??}{??}} for finding help pages
on a vague topic;
\code{\LinkA{help.start}{help.start}()} which opens the HTML version of the \R{}
help pages;
\code{\LinkA{library}{library}()} for listing available packages and the
help objects they contain;
\code{\LinkA{data}{data}()} for listing available data sets;
\code{\LinkA{methods}{methods}()}.

Use \code{\LinkA{prompt}{prompt}()} to get a prototype for writing \code{help}
pages of your own package.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
help()
help(help)              # the same

help(lapply)

help("for")             # or ?"for", but quotes/backticks are needed

help(package="splines") # get help even when package is not loaded

topi <- "women"
help(topi)

try(help("bs", try.all.packages=FALSE)) # reports not found (an error)
help("bs", try.all.packages=TRUE)       # reports can be found
                                        # in package 'splines'
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{help.request}{Send a Post to R-help}{help.request}
\keyword{utilities}{help.request}
\keyword{error}{help.request}
%
\begin{Description}\relax
Prompts the user to check they have done all that is expected of them
before sending a post to the R-help mailing list, provides a template
for the post with session information included and optionally sends
the email (on Unix systems).
\end{Description}
%
\begin{Usage}
\begin{verbatim}
help.request(subject = "",
             ccaddress = Sys.getenv("USER"),
             method = getOption("mailer"),
             address = "r-help@R-project.org",
             file = "R.help.request")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{subject}] subject of the email.  Please do not use single quotes
(\kbd{'}) in the subject!  Post separate help requests for multiple
queries.
\item[\code{ccaddress}] optional email address for copies (default is current
user).  Use \code{ccaddress = FALSE} for no copies.
\item[\code{method}] submission method: for Unix one of \code{"mailx"},
\code{"gnudoit"}, \code{"none"} or \code{"ess"}; for Windows either
\code{"none"} (default) or \code{"mailto"}.
\item[\code{address}] recipient's email address.
\item[\code{file}] file to use for setting up the email (or storing it when
method is \code{"none"} or sending mail fails).
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This function is not intended to replace the posting
guide. Please read the guide before posting to R-help or using this
function (see \url{http://www.r-project.org/posting-guide.html}).

The \code{help.request} function:
\begin{itemize}

\item asks whether the user has consulted relevant resources,
stopping and opening the relevant url if a negative response if
given.
\item checks whether the current version of R is being used and
whether the add-on packages are up-to-date, giving the option of
updating where necessary.
\item asks whether the user has prepared appropriate (minimal,
reproducible, self-contained, commented) example code ready to
paste into the post.

\end{itemize}

Once this checklist has been completed a template post is prepared
including current session information.

If method is \code{"none"} or \code{NULL}, then the default text
editor is opened for the user to complete the post.  Which editor is
used can be controlled using \code{\LinkA{options}{options}}, type
\code{getOption("editor")} to see what editor is currently
defined.  Please use the help pages of the respective editor for
details of usage.  The report can then be copied to your favorite email
program and sent to the r-help list.

On Windows systems there is an experimental \code{"mailto"} option,
which sends the template post to the system's default email program for
the user to edit and send.

On Unix systems there are three options for direct submission of the
post. If the submission method is \code{"mailx"}, then the default
editor is used to write the help request. After saving the help request
(in the temporary file opened) and exiting the editor the report is
mailed using a Unix command line mail utility such as \code{mailx}.  A
copy of the mail is sent to the current user. If method is
\code{"gnudoit"}, then an emacs mail buffer is opened and used for
sending the email. If method is \code{"ess"} the body of the mail is
simply sent to stdout.
\end{Details}
%
\begin{Value}
Nothing useful.
\end{Value}
%
\begin{Author}\relax
Heather Turner, based on code and help page of
\code{\LinkA{bug.report}{bug.report}()}.
\end{Author}
%
\begin{SeeAlso}\relax
The posting guide
(\url{http://www.r-project.org/posting-guide.html}),
also \code{\LinkA{sessionInfo}{sessionInfo}()} from which you may add
to the help request.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{help.search}{Search the Help System}{help.search}
\aliasA{??}{help.search}{??}
\aliasA{print.hsearch}{help.search}{print.hsearch}
\keyword{documentation}{help.search}
%
\begin{Description}\relax
Allows for searching the help system for documentation matching a
given character string in the (file) name, alias, title, concept or
keyword entries (or any combination thereof), using either
\LinkA{fuzzy matching}{fuzzy matching} or \LinkA{regular expression}{regular expression} matching.  Names
and titles of the matched help entries are displayed nicely formatted.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
help.search(pattern, fields = c("alias", "concept", "title"),
            apropos, keyword, whatis, ignore.case = TRUE,
            package = NULL, lib.loc = NULL,
            help.db = getOption("help.db"),
            verbose = getOption("verbose"),
            rebuild = FALSE, agrep = NULL, use_UTF8 = FALSE)
??pattern
field??pattern
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{pattern}] a character string to be matched in the specified
fields.  If this is given, the arguments \code{apropos},
\code{keyword}, and \code{whatis} are ignored.
\item[\code{fields}] a character vector specifying the fields of the help
database to be searched.  The entries must be abbreviations of
\code{"name"}, \code{"title"}, \code{"alias"}, \code{"concept"}, and
\code{"keyword"}, corresponding to the help page's (file) name, its
title, the topics and concepts it provides documentation for, and
the keywords it can be classified to.
\item[\code{apropos}] a character string to be matched in the help page
topics and title.
\item[\code{keyword}] a character string to be matched in the help page
`keywords'. `Keywords' are really categories: the
standard categories are listed in file \file{R.home("doc")/KEYWORDS}
(see also the example) and some package writers have defined their
own.  If \code{keyword} is specified, \code{agrep} defaults to
\code{FALSE}. \item[\code{whatis}] a character string to be matched in
the help page topics.
\item[\code{ignore.case}] a logical.  If \code{TRUE}, case is ignored during
matching; if \code{FALSE}, pattern matching is case sensitive.
\item[\code{package}] a character vector with the names of packages to
search through, or \code{NULL} in which case \emph{all} available
packages in the library trees specified by \code{lib.loc} are
searched.
\item[\code{lib.loc}] a character vector describing the location of \R{}
library trees to search through, or \code{NULL}.  The default value
of \code{NULL} corresponds to all libraries currently known.
\item[\code{help.db}] a character string giving the file path to a previously
built and saved help database, or \code{NULL}.
\item[\code{verbose}] logical; if \code{TRUE}, the search process is traced.
Integer values are also accepted, with \code{TRUE} being equivalent
to \code{2}, and \code{1} being less verbose.  On Windows a progress
bar is shown during rebuilding, and on Unix a heartbeat is shown for
\code{verbose = 1} and a package-by-package list for
\code{verbose >= 2}.
\item[\code{rebuild}] a logical indicating whether the help database should
be rebuilt.  This will be done automatically if \code{lib.loc} or
the search path is changed, or if \code{package} is used and a value
is not found.
\item[\code{agrep}] if \code{NULL} (the default unless \code{keyword} is
used) and the character string to
be matched consists of alphanumeric characters, whitespace or a dash
only, approximate (fuzzy) matching via \code{\LinkA{agrep}{agrep}} is used
unless the string has fewer than 5 characters; otherwise, it is
taken to contain a \LinkA{regular expression}{regular expression} to be matched via
\code{\LinkA{grep}{grep}}.  If \code{FALSE}, approximate matching is not
used.  Otherwise, one can give a numeric or a list specifying the
maximal distance for the approximate match, see argument
\code{max.distance} in the documentation for \code{\LinkA{agrep}{agrep}}.
\item[\code{use\_UTF8}] logical: should be results be given in UTF-8 encoding?
Also changes the meaning of regexps in \code{agrep} to be Perl regexps.
\item[\code{field}] a single value of \code{fields} to search.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Upon installation of a package, a pre-built help.search index is
serialized as \file{hsearch.rds} in the \file{Meta} directory
(provided the package has any help pages).  These files are used to
create the database.

The arguments \code{apropos} and \code{whatis} play a role similar to
the Unix commands with the same names.

Searching with \code{agrep = FALSE} will be several times faster
than the default (once the database is built).  However, as from
\R{} 2.10.0 approximate searches should be fast enough (around a second
with 2000 packages installed).

If possible, the help database is saved in memory or (if memory
limits have been set: see \code{\LinkA{mem.limits}{mem.limits}}) to a file in the
session temporary directory for use by subsequent calls in the
session.

Note that currently the aliases in the matching help files are not
displayed.

As with \code{\LinkA{?}{?}}, in \code{??} the pattern may be prefixed with a 
package name followed by \code{::} or \code{:::} to limit the search
to that package.
\end{Details}
%
\begin{Value}
The results are returned in a list object of class \code{"hsearch"},
which has a print method for nicely formatting the results of the
query.  This mechanism is experimental, and may change in future
versions of \R{}.

In R.app on Mac OS X, this will show up a browser with selectable items. On
exiting this browser, the help pages for the selected items will be
shown in separate help windows.

The internal format of the class is undocumented and subject to change.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{help}{help}};
\code{\LinkA{help.start}{help.start}} for starting the hypertext (currently HTML)
version of \R{}'s online documentation, which offers a similar search
mechanism.

\code{\LinkA{RSiteSearch}{RSiteSearch}} to access an on-line search of \R{} resources.

\code{\LinkA{apropos}{apropos}} uses regexps and has nice examples.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
help.search("linear models")    # In case you forgot how to fit linear
                                # models
help.search("non-existent topic")

??utils::help  # All the topics matching "help" in the utils package

## Not run: 
help.search("print")            # All help pages with topics or title
                                # matching 'print'
help.search(apropos = "print")  # The same

help.search(keyword = "hplot")  # All help pages documenting high-level
                                # plots.
file.show(file.path(R.home("doc"), "KEYWORDS"))  # show all keywords

## Help pages with documented topics starting with 'try'.
help.search("\\btry", fields = "alias")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{help.start}{Hypertext Documentation}{help.start}
\keyword{documentation}{help.start}
%
\begin{Description}\relax
Start the hypertext (currently HTML) version of \R{}'s online
documentation.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
help.start(update = FALSE, gui = "irrelevant",
           browser = getOption("browser"), remote = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{update}] logical: should this attempt to update the package index to
reflect the currently available packages.  (Not attempted if
\code{remote} is non-\code{NULL}.)
\item[\code{gui}] just for compatibility with S-PLUS.
\item[\code{browser}] the name of the program to be used as hypertext
browser.  It should be in the \env{PATH}, or a full path specified.
Alternatively, it can be an \R{} function which will be called with a
URL as its only argument.

\item[\code{remote}] A character string giving a valid URL for the
\file{\var{\LinkA{R\_HOME}{R.Rul.HOME}}} directory on a remote location.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Unless \code{remote} is specified this requires the HTTP server to be
available (it will be started if possible: see
\code{\LinkA{startDynamicHelp}{startDynamicHelp}}).

One of the links on the index page is the HTML package index,
\file{R.home("docs")/html/packages.html}, which can be remade by
\code{\LinkA{make.packages.html}{make.packages.html}(temp = FALSE)}.  For local operation,
the HTTP server will remake a temporary version of this list when the
link is first clicked, and each time thereafter check if updating is
needed (if \code{\LinkA{.libPaths}{.libPaths}} has changed or any of the
directories has been changed).  This can be slow, and using
\code{update = TRUE} will ensure that the packages list is updated
before launching the index page.

Argument \code{remote} can be used to point to HTML help published by
another \R{} installation: it will typically only show packages from the
main library of that installation.
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{help}{help}()} for on- and off-line help in other formats.

\code{\LinkA{browseURL}{browseURL}} for how the help file is displayed.

\code{\LinkA{RSiteSearch}{RSiteSearch}} to access an on-line search of \R{} resources.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
help.start()
## Not run: 
## the 'remote' arg can be tested by
help.start(remote=paste("file://", R.home(), sep=""))

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{index.search}{Search Indices for Help Files}{index.search}
\keyword{utilities}{index.search}
%
\begin{Description}\relax
Used to search the indices for help files, possibly under aliases.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
index.search(topic, path, file="AnIndex", type = "help")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{topic}] The keyword to be searched for in the indices.
\item[\code{path}] The path(s) to the packages to be searched.
\item[\code{file}] The index file to be searched.  Normally
\file{"AnIndex"}.
\item[\code{type}] The type of file required.  Deprecated in \R{} 2.10.0.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
For each package in \code{path}, examine the file \code{file} in
directory \file{type}, and look up the matching file stem for topic
\code{topic}, if any.  
\end{Details}
%
\begin{Value}
A character vector of matching files, as if they are in directory
\code{type} of the corresponding package.  In the special cases of
\code{type = "html"}, \code{"R-ex"} and \code{"latex"} the file
extensions \code{".html"}, \code{".R"} and \code{".tex"} are added.
Note that as from \R{} 2.10.0 no such files exist.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{help}{help}},
\code{\LinkA{example}{example}}
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{INSTALL}{Install Add-on Packages}{INSTALL}
\keyword{utilities}{INSTALL}
%
\begin{Description}\relax
Utility for installing add-on packages.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
R CMD INSTALL [options] [-l lib] pkgs
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{pkgs}] a space-separated list with the path names of the packages to be
installed.
\item[\code{lib}] the path name of the \R{} library tree to install to.  Also
accepted in the form \samp{--library=lib}.
\item[\code{options}] a space-separated list of options through which in
particular the process for building the help files can be controlled.
Most options should only be given once, and paths including spaces should
be quoted.  Use \command{R CMD INSTALL --help} for the full current
list of options.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This will stop at the first error, so if you want all the \code{pkgs}
to be tried, call this via a shell loop.

If used as \command{R CMD INSTALL pkgs} without explicitly specifying
\code{lib}, packages are installed into the library tree rooted at the
first directory in the library path which would be used by \R{} run in
the current environment.

To install into the library tree \code{\var{lib}}, use
\command{R CMD INSTALL -l \var{lib} \var{pkgs}}.
This prepends \code{lib} to the library path for
duration of the install, so required packages in the installation
directory will be found (and used in preference to those in other
libraries).

Both \code{lib} and the elements of \code{pkgs} may be absolute or
relative path names of directories.  \code{pkgs} may also contain
names of package/bundle archive files: these are then
extracted to a temporary directory.  Precisely what forms are accepted
depends on the system (see \code{\LinkA{untar}{untar}}): this will certainly
include filenames of the form \file{pkg\_version.tar.gz} as obtained
from CRAN, \file{pkg.tgz} and \file{pkg.tar.gz}, usually
\file{pkg\_version.tar.bz2} and \file{pkg.tar.bz2}, and on systems with
recent toolsets, \file{pkg\_version.ext} for any supported extension
such as \file{tar.lzma} and \file{tar.xz}.
Finally, binary package/bundle archive files (as created by
\command{R CMD build --binary}) can be supplied.

The package sources can be cleaned up prior to installation by
\option{--preclean} or after by \option{--clean}: cleaning is
essential if the sources are to be used with more than one
architecture or platform.

Some package sources contain a \file{configure} script that can be
passed arguments or variables via the option \option{--configure-args}
and \option{--configure-vars}, respectively, if necessary.  The latter
is useful in particular if libraries or header files needed for the
package are in non-system directories.  In this case, one can use the
configure variables \code{LIBS} and \code{CPPFLAGS} to specify these
locations (and set these via \option{--configure-vars}), see section
``Configuration variables'' in ``R Installation and
Administration'' for more information.  (If these are used more than
once on the command line they are concatenated.)  The configure
mechanism can be bypassed using the option \option{--no-configure}.

If the attempt to install the package fails, leftovers are removed.
If the package was already installed, the old version is restored.
This happens either if a command encounters an error or if the
install is interrupted from the keyboard: after cleaning up the script
terminates.

By default the library directory is `locked' by creating a
directory \file{00LOCK} within it.  This has two purposes: it
prevents any other process installing into that library concurrently,
and is used to store any previous version of the package/bundle to
restore on error.  A finer-grained locking is provided by the option
\option{--pkglock} which creates a separate lock for each
package/bundle: this allows enough freedom for careful parallel
installation as done by \code{\LinkA{install.packages}{install.packages}(Ncpus =
  \var{n})} with \code{\var{n} > 1}.  Finally locking (and restoration
on error) can be suppressed by \option{--no-lock} or \option{--unsafe}
(two names for the same option).

Some platforms (notably Mac OS X) support sub-architectures in which
binaries for different CPUs are installed within the same library
tree. For such installations, the default behaviour is to try to build
packages for all installed sub-architectures unless the package has a
configure script or a \file{src/Makefile}, when only the
sub-architecture running \command{R CMD INSTALL} is used.  To use only
that sub-architecture, use \option{--no-multiarch}.  To install just
the compiled code for another sub-architecture, use
\option{--libs-only}.

Use \command{R CMD INSTALL --help} for concise usage information,
including all the available options
\end{Details}
%
\begin{Section}{Packages using the methods package}
Packages that require the methods package and make use functions such
as \code{\LinkA{setMethod}{setMethod}} or \code{\LinkA{setClass}{setClass}}, should be
installed using lazy-loading: use the field \code{LazyLoad} in the
\file{DESCRIPTION} file to ensure this.
\end{Section}
%
\begin{Note}\relax
Some parts of the operation of \code{INSTALL} depend on the \R{}
temporary directory (see \code{\LinkA{tempdir}{tempdir}}, usually under
\file{/tmp}) having both write and execution access to the account
running \R{}.  This is usually the case, but if \file{/tmp} has been
mounted as \code{noexec}, environment variable \env{TMPDIR} may need
to be set to a directory from which execution is allowed.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{REMOVE}{REMOVE}} and \code{\LinkA{library}{library}} for information on
using several library trees;
\code{\LinkA{update.packages}{update.packages}} for automatic update of packages using
the internet (or other \R{} level installation of packages, such as by
\code{install.packages}).

The section on ``Add-on packages'' in ``R Installation and
Administration'' and the chapter on ``Creating R packages'' in
``Writing \R{} Extensions''
\code{\LinkA{RShowDoc}{RShowDoc}} and the \file{doc/manual} subdirectory of the
\R{} source tree).
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{installed.packages}{Find Installed Packages}{installed.packages}
\keyword{utilities}{installed.packages}
%
\begin{Description}\relax
Find (or retrieve) details of all packages installed in the specified
libraries.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
installed.packages(lib.loc = NULL, priority = NULL,
                   noCache = FALSE, fields = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{lib.loc}] 
character vector describing the location of \R{} library trees to
search through, or \code{NULL} for all known trees
(see \code{\LinkA{.libPaths}{.libPaths}}).

\item[\code{priority}] 
character vector or \code{NULL} (default).  If non-null, used to
select packages; \code{"high"} is equivalent to
\code{c("base", "recommended")}.  To select all packages without an
assigned priority use \code{priority = "NA"}.

\item[\code{noCache}] Do not use cached information.

\item[\code{fields}] a character vector giving the fields to extract from
each package's \code{DESCRIPTION} file in addition to the default
ones, or \code{NULL} (default).  Unavailable fields result in
\code{NA} values.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\code{installed.packages} scans the \file{DESCRIPTION} files of each
package found along \code{lib.loc} and returns a matrix of package
names, library paths and version numbers.

\strong{Note:} this works with package names, not bundle names.

The information found is cached (by library) for the \R{} session and
specified \code{fields} argument, and updated only if the top-level
library directory has been altered, for example by installing or
removing a package.  If the cached information becomes confused, it
can be refreshed by running \code{installed.packages(noCache =
  TRUE)}.
\end{Details}
%
\begin{Value}
A matrix with one row per package, row names the package names and
column names
\code{"Package"}, \code{"LibPath"},
\code{"Version"}, \code{"Priority"},
\code{"Bundle"}, \code{"Contains"},
\code{"Depends"}, \code{"Imports"}, \code{"LinkingTo"},
\code{"Suggests"}, \code{"Enhances"},
\code{"OS\_type"}, \code{"License"} and
\code{"Built"} (the \R{} version the package was built under).
Additional columns can be specified using the \code{fields}
argument.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{update.packages}{update.packages}}, \code{\LinkA{INSTALL}{INSTALL}}, \code{\LinkA{REMOVE}{REMOVE}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
str(ip <- installed.packages(priority = "high"))
ip[, c(1,3:5)]
plic <- installed.packages(priority = "high", fields="License")
## what licenses are there:
table( plic[,"License"] )
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{LINK}{Create Executable Programs}{LINK}
\keyword{utilities}{LINK}
%
\begin{Description}\relax
Front-end for creating executable programs.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
R CMD LINK [options] linkcmd
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{linkcmd}] a list of commands to link together suitable object
files (include library objects) to create the executable program.
\item[\code{options}] further options to control the linking, or for
obtaining information about usage and version.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The linker front-end is useful in particular when linking against the
R shared library, in which case \code{linkcmd} must contain \code{-lR}
but need not specify its library path.

Currently only works if the C compiler is used for linking, and no C++
code is used.

Use \command{R CMD LINK --help} for more usage information.
\end{Details}
%
\begin{Note}\relax
Some binary distributions of \R{} have \code{LINK} in a separate
bundle, e.g. an \code{R-devel} RPM.
\end{Note}
\inputencoding{latin1}
\HeaderA{localeToCharset}{Select a Suitable Encoding Name from a Locale Name}{localeToCharset}
\keyword{utilities}{localeToCharset}
%
\begin{Description}\relax
This functions aims to find a suitable coding for the locale named, by
default the current locale, and if it is a UTF-8 locale a suitable
single-byte encoding.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
localeToCharset(locale = Sys.getlocale("LC_CTYPE"))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{locale}] character string naming a locale.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The operation differs by OS.
Locale names are normally like \code{es\_MX.iso88591}.  If final
component indicates an encoding and it is not \code{utf8} we just need
to look up the equivalent encoding name.  Otherwise, the language
(here \code{es}) is used to choose a primary or fallback encoding.

In the \code{C} locale the answer will be \code{"ASCII"}.
\end{Details}
%
\begin{Value}
A character vector naming an encoding and possibly a fallback
single-encoding,  \code{NA} if unknown.
\end{Value}
%
\begin{Note}\relax
The encoding names are those used by \code{libiconv}, and ought also
to work with \code{glibc} but maybe not with commercial Unixen.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{Sys.getlocale}{Sys.getlocale}}, \code{\LinkA{iconv}{iconv}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
localeToCharset()
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{ls.str}{List Objects and their Structure}{ls.str}
\aliasA{lsf.str}{ls.str}{lsf.str}
\aliasA{print.ls\_str}{ls.str}{print.ls.Rul.str}
\keyword{print}{ls.str}
\keyword{utilities}{ls.str}
%
\begin{Description}\relax
\code{ls.str} and \code{lsf.str} are variations of \code{\LinkA{ls}{ls}}
applying \code{\LinkA{str}{str}()} to each matched name: see section Value.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
ls.str(pos = -1, name, envir, all.names = FALSE,
       pattern, mode = "any")

lsf.str(pos = -1, envir, ...)

## S3 method for class 'ls\_str':
print(x, max.level = 1, give.attr = FALSE, ...,
      digits = max(1, getOption("str")$digits.d))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{pos}] integer indicating \code{\LinkA{search}{search}} path position.
\item[\code{name}] optional name indicating \code{\LinkA{search}{search}} path
position, see \code{\LinkA{ls}{ls}}.
\item[\code{envir}] environment to use, see \code{\LinkA{ls}{ls}}.
\item[\code{all.names}] logical indicating if names which begin with a
\code{.} are omitted; see \code{\LinkA{ls}{ls}}.
\item[\code{pattern}] a \LinkA{regular expression}{regular expression} passed to \code{\LinkA{ls}{ls}}.
Only names matching \code{pattern} are considered.
\item[\code{max.level}] maximal level of nesting which is applied for
displaying nested structures, e.g., a list containing sub lists.
Default 1: Display only the first nested level.
\item[\code{give.attr}] logical; if \code{TRUE} (default), show attributes
as sub structures.
\item[\code{mode}] character specifying the \code{\LinkA{mode}{mode}} of objects to
consider.  Passed to \code{\LinkA{exists}{exists}} and \code{\LinkA{get}{get}}.
\item[\code{x}] an object of class \code{"ls\_str"}.
\item[\code{...}] further arguments to pass.  \code{lsf.str} passes them to
\code{ls.str} which passes them on to \code{\LinkA{ls}{ls}}.  The
(non-exported) print method \code{print.ls\_str} passes them to
\code{\LinkA{str}{str}}.
\item[\code{digits}] the number of significant digits to use for printing.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
\code{ls.str} and \code{lsf.str} return an object of class
\code{"ls\_str"}, basically the character vector of matching names
(functions only for \code{lsf.str}), similarly to
\code{\LinkA{ls}{ls}}, with a \code{print()} method that calls \code{\LinkA{str}{str}()}
on each object.
\end{Value}
%
\begin{Author}\relax
Martin Maechler
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{str}{str}}, \code{\LinkA{summary}{summary}}, \code{\LinkA{args}{args}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(stats)

lsf.str()#- how do the functions look like which I am using?
ls.str(mode = "list") #- what are the structured objects I have defined?

## create a few objects
example(glm, echo = FALSE)
ll <- as.list(LETTERS)
print(ls.str(), max.level = 0)# don't show details

## which base functions have "file" in their name ?
lsf.str(pos = length(search()), pattern = "file")

## demonstrating that  ls.str() works inside functions
## ["browser/debug mode"]:
tt <- function(x, y=1) { aa <- 7; r <- x + y; ls.str() }
(nms <- sapply(strsplit(capture.output(tt(2))," *: *"), `[`, 1))
stopifnot(nms == c("aa", "r","x","y"))
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{make.packages.html}{Update HTML Package List}{make.packages.html}
\keyword{utilities}{make.packages.html}
%
\begin{Description}\relax
Re-create the HTML documentation files to reflect all available packages.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
make.packages.html(lib.loc = .libPaths(), temp = TRUE, verbose = TRUE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{lib.loc}] character vector. List of libraries to be included.
\item[\code{temp}] logical: should the package indices be created in a
temporary location for use by the HTTP server?
\item[\code{verbose}] ogical: should messages and a heartbeat be shown?
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This creates the \file{packages.html} file, either a temporary copy
for use by \code{\LinkA{help.start}{help.start}}, or the copy in
\file{R.home("doc")/html} (for which you will need write permission).

It can be very slow, as all the package \file{DESCRIPTION} files in
all the library trees are read.

For \code{temp = TRUE} there is some caching of information, so the
file will only be re-created if \code{lib.loc} or any of the
directories it lists have been changed.
\end{Details}
%
\begin{Value}
Invisible logical, with \code{FALSE} indicating a failure to create
the file, probably due to lack of suitable permissions.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{help.start}{help.start}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
# to prefer HTML help, put in your .Rprofile
options(help_type = "html")
make.packages.html(temp = FALSE)
# this can be slow for large numbers of installed packages.

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{make.socket}{Create a Socket Connection}{make.socket}
\aliasA{print.socket}{make.socket}{print.socket}
\keyword{misc}{make.socket}
%
\begin{Description}\relax
With \code{server = FALSE} attempts to open a client socket to the
specified port and host. With \code{server = TRUE} listens on the
specified port for a connection and then returns a server socket. It is
a good idea to use \code{\LinkA{on.exit}{on.exit}} to ensure that a socket is
closed, as you only get 64 of them.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
make.socket(host = "localhost", port, fail = TRUE, server = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{host}] name of remote host
\item[\code{port}] port to connect to/listen on
\item[\code{fail}] failure to connect is an error?
\item[\code{server}] a server socket?
\end{ldescription}
\end{Arguments}
%
\begin{Value}
An object of class \code{"socket"}.
\begin{ldescription}
\item[\code{socket}] socket number. This is for internal use
\item[\code{port}] port number of the connection
\item[\code{host}] name of remote computer
\end{ldescription}
\end{Value}
%
\begin{Section}{Warning}
I don't know if the connecting host name returned
when \code{server = TRUE} can be trusted. I suspect not.
\end{Section}
%
\begin{Author}\relax
Thomas Lumley
\end{Author}
%
\begin{References}\relax
Adapted from Luke Tierney's code for \code{XLISP-Stat}, in turn
based on code from Robbins and Robbins "Practical UNIX Programming"
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{close.socket}{close.socket}}, \code{\LinkA{read.socket}{read.socket}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
daytime <- function(host = "localhost"){
    a <- make.socket(host, 13)
    on.exit(close.socket(a))
    read.socket(a)
}
## Official time (UTC) from US Naval Observatory
## Not run: daytime("tick.usno.navy.mil")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{memory.size}{Report on Memory Allocation}{memory.size}
\aliasA{memory.limit}{memory.size}{memory.limit}
\keyword{utilities}{memory.size}
%
\begin{Description}\relax
\code{memory.size} and \code{memory.limit} are used to manage the
total memory allocation on Windows.  On other platforms these are
stubs which report infinity with a warning.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
memory.size(max = FALSE)

memory.limit(size = NA)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{max}] logical. If true the maximum amount of memory obtained from
the OS is reported, otherwise the amount currently in use.
\item[\code{size}] numeric. If \code{NA} report the memory size, otherwise
request a new limit, in Mb.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
To restrict memory usage on a Unix-alike use the facilities of the
shell used to launch \R{}, e.g. \code{limit} or \code{ulimit}.
\end{Details}
%
\begin{Value}
Size in bytes: always \code{Inf}.
\end{Value}
%
\begin{SeeAlso}\relax
\LinkA{Memory-limits}{Memory.Rdash.limits} for other limits.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{menu}{Menu Interaction Function}{menu}
\keyword{utilities}{menu}
\keyword{programming}{menu}
%
\begin{Description}\relax
\code{menu} presents the user with a menu of choices labelled from 1
to the number of choices.  To exit without choosing an item one can
select \samp{0}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
menu(choices, graphics = FALSE, title = "")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{choices}] a character vector of choices
\item[\code{graphics}] a logical indicating whether a graphics menu should be
used if available.
\item[\code{title}] a character string to be used as the title of the menu.
\code{NULL} is also accepted.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If \code{graphics = TRUE} and a windowing system is available
(Windows, Mac OS X or X11 \emph{via} Tcl/Tk) a listbox widget is
used, otherwise a text menu.  It is an error to use \code{menu} in a
non-interactive session.

Ten or fewer items will be displayed in a single column, more in
multiple columns if possible within the current display width.

No title is displayed if \code{title} is \code{NULL} or \code{""}.
\end{Details}
%
\begin{Value}
The number corresponding to the selected item, or 0 if no choice was
made.
\end{Value}
%
\begin{References}\relax
Becker, R. A., Chambers, J. M. and Wilks, A. R. (1988)
\emph{The New S Language}.
Wadsworth \& Brooks/Cole.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{select.list}{select.list}}, which is used to implement the graphical
menu, and allows multiple selections.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
switch(menu(c("List letters", "List LETTERS")) + 1,
       cat("Nothing done\n"), letters, LETTERS)

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{methods}{List Methods for S3 Generic Functions or Classes}{methods}
\aliasA{print.MethodsFunction}{methods}{print.MethodsFunction}
\keyword{methods}{methods}
%
\begin{Description}\relax
List all available methods for an S3 generic function, or all
methods for a class.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
methods(generic.function, class)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{generic.function}] a generic function, or a character string naming a
generic function.
\item[\code{class}] a symbol or character string naming a class: only used if
\code{generic.function} is not supplied.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Function \code{methods} can be used to find out about the methods for
a particular generic function or class.  The functions listed are those
which \emph{are named like methods} and may not actually be methods
(known exceptions are discarded in the code).  Note that the listed
methods may not be user-visible objects, but often help will be
available for them.

If \code{class} is used, we check that a matching generic can be found
for each user-visible object named.  If \code{generic.function} is
given, there is a warning if it appears not to be a generic function.
(The check for being generic used can be fooled.)
\end{Details}
%
\begin{Value}
An object of class \code{"MethodsFunction"}, a
character vector of function names with an \code{"info"} attribute.
There is a \code{print} method which marks with an asterisk any
methods which are not visible: such functions can be examined by
\code{\LinkA{getS3method}{getS3method}} or \code{\LinkA{getAnywhere}{getAnywhere}}.

The \code{"info"} attribute is a data frame, currently with a
logical column, \code{visible} and a factor column \code{from}
(indicating where the methods were found).
\end{Value}
%
\begin{Note}\relax
This scheme is called \emph{S3} (S version 3).  For new projects,
it is recommended to use the more flexible and robust \emph{S4} scheme
provided in the \pkg{methods} package.  Functions can have both S3
and S4 methods, and function \code{\LinkA{showMethods}{showMethods}} will
list the S4 methods (possibly none).

The original \code{methods} function was written by Martin Maechler.
\end{Note}
%
\begin{References}\relax
Chambers, J. M. (1992)
\emph{Classes and methods: object-oriented programming in S.}
Appendix A of \emph{Statistical Models in S}
eds J. M. Chambers and T. J. Hastie, Wadsworth \& Brooks/Cole.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{S3Methods}{S3Methods}}, \code{\LinkA{class}{class}}, \code{\LinkA{getS3method}{getS3method}}.

For S4, \code{\LinkA{showMethods}{showMethods}}, \code{\LinkA{Methods}{Methods}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(stats)

methods(summary)
methods(class = "aov")
methods("[[")    # uses C-internal dispatching
methods("$")
methods("$<-")   # replacement function
methods("+")     # binary operator
methods("Math")  # group generic
require(graphics)
methods("axis")  # looks like it has methods, but not generic
## Not run: 
methods(print)   # over 100

## End(Not run)
## --> help(showMethods) for related examples
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{mirrorAdmin}{Managing Repository Mirrors}{mirrorAdmin}
\aliasA{checkCRAN}{mirrorAdmin}{checkCRAN}
\aliasA{mirror2html}{mirrorAdmin}{mirror2html}
\keyword{misc}{mirrorAdmin}
%
\begin{Description}\relax
Functions helping to maintain CRAN, some of them may also be useful
for administrators of other repository networks.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
mirror2html(mirrors = NULL, file = "mirrors.html",
  head = "mirrors-head.html", foot = "mirrors-foot.html")
checkCRAN(method)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{mirrors}] A data frame, by default the CRAN list of mirrors is used.
\item[\code{file}] A connection object or a character string.
\item[\code{head}] Name of optional header file.
\item[\code{foot}] Name of optional footer file.
\item[\code{method}] Download method, see \code{download.file}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\code{mirror2html} creates the HTML file for the CRAN list of mirrors
and invisibly returns the HTML text.

\code{checkCRAN} performs a sanity checks on all CRAN mirrors.
\end{Details}
\inputencoding{latin1}
\HeaderA{modifyList}{Recursively Modify Elements of a List}{modifyList}
\keyword{utilities}{modifyList}
%
\begin{Description}\relax
Modifies a possibly nested list recursively by changing a subset of
elements at each level to match a second list.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
modifyList(x, val)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] a named \code{\LinkA{list}{list}}, possibly empty.
\item[\code{val}] a named list with components to replace corresponding
components in \code{x}.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
A modified version of \code{x}, with the modifications determined as
follows (here, list elements are identified by their names).  Elements
in \code{val} which are missing from \code{x} are added to \code{x}.
For elements that are common to both but are not both lists
themselves, the component in \code{x} is replaced by the one in
\code{val}.  For common elements that are both lists, \code{x[[name]]}
is replaced by \code{modifyList(x[[name]], val[[name]])}.
\end{Value}
%
\begin{Author}\relax
 Deepayan Sarkar \email{Deepayan.Sarkar@R-project.org}
\end{Author}
%
\begin{Examples}
\begin{ExampleCode}
foo <- list(a = 1, b = list(c = "a", d = FALSE))
bar <- modifyList(foo, list(e = 2, b = list(d = TRUE)))
str(foo)
str(bar)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{news}{Build and Query R or Package News Information}{news}
%
\begin{Description}\relax
Build and query the news for R or add-on packages.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
news(query, package = "R", lib.loc = NULL, format = NULL, 
     reader = NULL, db = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{query}] an expression for selecting news entries
\item[\code{package}] a character string giving the name of an installed
add-on package, or \code{"R"}.
\item[\code{lib.loc}] a character vector of directory names of R libraries,
or \code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.
\item[\code{format}] Not yet used.
\item[\code{reader}] Not yet used.
\item[\code{db}] a news db obtained from \code{news()}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If \code{package} is \code{"R"} (default), \code{\LinkA{readNEWS}{readNEWS}} in
package \pkg{tools} is used to build a news db from the R \file{NEWS}
file.  Otherwise, if the given add-on package can be found in the
given libraries and has a \file{NEWS} file, it is attempted to read
the package news in structured form.  The \file{NEWS} files in add-on
packages use a variety of different formats; the default news reader
should be capable to extract individual news entries from a majority
of packages from the standard repositories, which use (slight
variations of) the following format:

\begin{itemize}

\item Entries are grouped according to version, with version header
\samp{Changes in version} at the beginning of a line, followed by a
version number, optionally followed by an ISO 8601 (\%Y-\%m-\%d, see
\code{\LinkA{strptime}{strptime}}) format date, possibly parenthesized.
\item Entries may be grouped according to category, with a category
header (different from a version header) starting at the beginning
of a line.
\item Entries are written as itemize-type lists, using one of
\samp{o}, \samp{*}, \samp{-} or \samp{+} as item tag.  Entries must
be indented, and ideally use a common indentation for the item
texts.

\end{itemize}


Additional formats and readers may be supported in the future.

The news db built is a character data frame inheriting from
\code{"news\_db"} with variables \code{Version}, \code{Category},
\code{Date} and \code{Text}, where the last contains the entry texts
read, and the other variables may be \code{NA} if they were missing or
could not be determined.

Using \code{query}, one can select news entries from the db.  If
missing or \code{NULL}, the complete db is returned.  Otherwise,
\code{query} should be an expression involving (a subset of) the
variables \code{Version}, \code{Category}, \code{Date} and
\code{Text}, and when evaluated within the db returning a logical
vector with length the number of entries in the db.  The entries for
which evaluation gave \code{TRUE} are selected.  When evaluating,
\code{Version} and \code{Date} are coerced to
\code{\LinkA{numeric\_version}{numeric.Rul.version}} and \code{\LinkA{Date}{Date}} objects,
respectively, so that the comparison operators for these classes can
be employed.
\end{Details}
%
\begin{Value}
An data frame inheriting from class \code{"news\_db"}.
\end{Value}
%
\begin{Examples}
\begin{ExampleCode}
## Build a db of all R news entries.
db <- news()
## Bug fixes with PR number in 2.9.0.
news(Version == "2.9.0" & grepl("^BUG", Category) & grepl("PR#", Text),
     db = db)
## Entries with version >= 2.8.1 (including "2.8.1 patched"):
table(news(Version >= "2.8.1", db = db)$Version)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{normalizePath}{Express File Paths in Canonical Form}{normalizePath}
\keyword{utilities}{normalizePath}
%
\begin{Description}\relax
Convert file paths to canonical form for the platform, to display them
in a user-understandable form.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
normalizePath(path)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{path}] character vector of file paths.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Where the platform supports it this turns paths into absolute paths
in their canonical form (no \samp{./}, \samp{../} nor symbolic links).

If the path is not a real path the result is undefined.
On Unix-alikes, this will likely be the corresponding input element.
On Windows, it will likely result in an error being signalled.
\end{Details}
%
\begin{Value}
A character vector.
\end{Value}
%
\begin{Examples}
\begin{ExampleCode}
cat(normalizePath(c(R.home(), tempdir())), sep = "\n")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{nsl}{Look up the IP Address by Hostname}{nsl}
\keyword{utilities}{nsl}
%
\begin{Description}\relax
Interface to \code{gethostbyname}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
nsl(hostname)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{hostname}] the name of the host.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
The IP address, as a character string, or \code{NULL} if the call fails.
\end{Value}
%
\begin{Note}\relax
This was included as a test of internet connectivity, to fail if
the node running R is not connected.  It will also return \code{NULL}
if BSD networking is not supported, including the header file
\file{arpa/inet.h}.
\end{Note}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: nsl("www.r-project.org")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{object.size}{Report the Space Allocated for an Object}{object.size}
\aliasA{print.object\_size}{object.size}{print.object.Rul.size}
\keyword{utilities}{object.size}
%
\begin{Description}\relax
Provides an estimate of the memory that is being used to store an \R{} object.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
object.size(x)

## S3 method for class 'object\_size':
print(x, quote = FALSE,
      units = c("b", "auto", "Kb", "Mb", "Gb"), ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] An \R{} object.
\item[\code{quote}] logical, indicating whether or not the result should be
printed with surrounding quotes.
\item[\code{units}] The units to be used in printing the size.
\item[\code{...}] Arguments to be passed to or from other methods.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Exactly which parts of the memory allocation should be attributed to
which object is not clear-cut.  This function merely provides a rough
indication: it should be reasonably accurate for atomic vectors, but
does not detect if elements of a list are shared, for example.
(Sharing amongst elements of a character vector is taken into account,
but not that between character vectors in a single object.)

The calculation is of the size of the object, and excludes the space
needed to store its name in the symbol table.

Associated space (e.g. the environment of a function and what the
pointer in a \code{EXTPTRSXP} points to) is not included in the
calculation.

Object sizes are larger on 64-bit platforms than 32-bit ones, but will
very likely be the same on different platforms with the same word
length and pointer size.
\end{Details}
%
\begin{Value}
An object of class \code{"object.size"} with a length-one double value,
an estimate of the memory allocation attributable to the object in bytes.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{Memory-limits}{Memory.Rdash.limits}} for the design limitations on object size.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
object.size(letters)
object.size(ls)
print(object.size(library), units = "auto")
## find the 10 largest objects in the base package
z <- sapply(ls("package:base"), function(x)
            object.size(get(x, envir = baseenv())))
as.matrix(rev(sort(z))[1:10])
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{package.skeleton}{Create a Skeleton for a New Source Package}{package.skeleton}
\keyword{file}{package.skeleton}
\keyword{utilities}{package.skeleton}
%
\begin{Description}\relax
\code{package.skeleton} automates some of the setup for a new source
package.  It creates directories, saves functions, data, and R code files to
appropriate places, and creates skeleton help files and a
\file{Read-and-delete-me} file describing further steps in packaging. 
\end{Description}
%
\begin{Usage}
\begin{verbatim}
package.skeleton(name = "anRpackage", list,
                 environment = .GlobalEnv,
                 path = ".", force = FALSE, namespace = FALSE,
                 code_files = character())
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{name}] character string: the package name and directory name for
your package.
\item[\code{list}] character vector naming the \R{} objects to put in the
package.  Usually, at most one of \code{list}, \code{environment},
or \code{code\_files} will be supplied.  See `Details'.
\item[\code{environment}] an environment where objects are looked for.  See
`Details'.
\item[\code{path}] path to put the package directory in.
\item[\code{force}] If \code{FALSE} will not overwrite an existing directory.
\item[\code{namespace}] a logical indicating whether to add a name space for
the package.  If \code{TRUE}, a \code{NAMESPACE} file is created
to export all objects whose names begin with a letter, plus all S4
methods and classes.
\item[\code{code\_files}] a character vector with the paths to R code files to
build the package around.  See `Details'.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The arguments \code{list}, \code{environment}, and \code{code\_files}
provide alternative ways to initialize the package.  If
\code{code\_files} is supplied, the files so named will be sourced to
form the environment, then used to generate the package skeleton.
Otherwise \code{list} defaults to the non-hidden files in
\code{environment} (those whose name does not start with \code{.}),
but can be supplied to select a subset of the objects in that
environment.

Stubs of help files are generated for functions, data objects, and
S4 classes and methods, using the \code{\LinkA{prompt}{prompt}},
\code{\LinkA{promptClass}{promptClass}},  and \code{\LinkA{promptMethods}{promptMethods}} functions.

The package sources are placed in subdirectory \code{name} of
\code{path}.  If \code{code\_files} is supplied, these files are
copied; otherwise, objects will be dumped into individual source
files.
The file names in \code{code\_files} should  have suffix \code{".R"} and
be in the current working directory.

The  filenames created for source and documentation try to be valid
for all OSes known to run R.  Invalid characters are replaced by
\samp{\_}, invalid names are preceded by \samp{zz}, and finally the
converted names are made unique by \code{\LinkA{make.unique}{make.unique}(sep =
    "\_")}.  This can be done for code and help files but not data files
(which are looked for by name). Also, the code and help files should
have names starting with an ASCII letter or digit, and this is
checked and if necessary \code{z} prepended.

When you are done, delete the \file{Read-and-delete-me} file, as it
should not be distributed.
\end{Details}
%
\begin{Value}
Used for its side-effects.
\end{Value}
%
\begin{References}\relax
Read the \emph{Writing R Extensions} manual for more details.

Once you have created a \emph{source} package you need to install it:
see the \emph{R Installation and Administration} manual,
\code{\LinkA{INSTALL}{INSTALL}} and \code{\LinkA{install.packages}{install.packages}}.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{prompt}{prompt}}, \code{\LinkA{promptClass}{promptClass}}, and
\code{\LinkA{promptMethods}{promptMethods}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(stats)
## two functions and two "data sets" :
f <- function(x,y) x+y
g <- function(x,y) x-y
d <- data.frame(a=1, b=2)
e <- rnorm(1000)

package.skeleton(list=c("f","g","d","e"), name="mypkg")

\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{packageDescription}{Package Description}{packageDescription}
\aliasA{print.packageDescription}{packageDescription}{print.packageDescription}
\keyword{utilities}{packageDescription}
%
\begin{Description}\relax
Parses and returns the \file{DESCRIPTION} file of a package.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
packageDescription(pkg, lib.loc = NULL, fields = NULL,
                   drop = TRUE, encoding = "")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{pkg}] a character string with the package name.
\item[\code{lib.loc}] a character vector of directory names of \R{} libraries,
or \code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.  If the default is used, the loaded
packages are searched before the libraries.
\item[\code{fields}] a character vector giving the tags of fields to return
(if other fields occur in the file they are ignored).
\item[\code{drop}] If \code{TRUE} and the length of \code{fields} is 1, then
a single character string with the value of the respective field is
returned instead of an object of class \code{"packageDescription"}.
\item[\code{encoding}] If there is an \code{Encoding} field, to what encoding
should re-encoding be attempted?  If \code{NA}, no re-encoding.  The
other values are as used by \code{\LinkA{iconv}{iconv}}, so the default
\code{""} indicates the encoding of the current locale.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
A package will not be `found' unless it has a \file{DESCRIPTION} file
which contains a valid \code{Version} field.  Different warnings are
given when no package directory is found and when there is a suitable
directory but no valid \file{DESCRIPTION} file.

An \LinkA{attach}{attach}ed environment named to look like a package
(e.g. \code{package:utils2}) will be ignored.
\end{Details}
%
\begin{Value}
If a \file{DESCRIPTION} file for the given package is found and can
successfully be read, \code{packageDescription} returns an object of
class \code{"packageDescription"}, which is a named list with the
values of the (given) fields as elements and the tags as names, unless
\code{drop = TRUE}.

If parsing the \file{DESCRIPTION} file was not successful, it returns
a named list of \code{NA}s with the field tags as names if \code{fields}
is not null, and \code{NA} otherwise.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{read.dcf}{read.dcf}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
packageDescription("stats")
packageDescription("stats", fields = c("Package", "Version"))

packageDescription("stats", fields = "Version")
packageDescription("stats", fields = "Version", drop = FALSE)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{packageStatus}{Package Management Tools}{packageStatus}
\aliasA{print.packageStatus}{packageStatus}{print.packageStatus}
\aliasA{summary.packageStatus}{packageStatus}{summary.packageStatus}
\aliasA{update.packageStatus}{packageStatus}{update.packageStatus}
\aliasA{upgrade}{packageStatus}{upgrade}
\methaliasA{upgrade.packageStatus}{packageStatus}{upgrade.packageStatus}
\keyword{utilities}{packageStatus}
%
\begin{Description}\relax
Summarize information about installed packages and packages
available at various repositories, and automatically upgrade outdated
packages.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
packageStatus(lib.loc = NULL, repositories = NULL, method,
              type = getOption("pkgType"))

## S3 method for class 'packageStatus':
summary(object, ...)

## S3 method for class 'packageStatus':
update(object, lib.loc = levels(object$inst$LibPath),
       repositories = levels(object$avail$Repository), ...)

## S3 method for class 'packageStatus':
upgrade(object, ask = TRUE, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{lib.loc}] a character vector describing the location of \R{}
library trees to search through, or \code{NULL}.  The default value
of \code{NULL} corresponds to all libraries currently known.
\item[\code{repositories}] a character vector of URLs describing the location of \R{}
package repositories on the Internet or on the local machine.
\item[\code{method}] Download method, see \code{\LinkA{download.file}{download.file}}.
\item[\code{type}] type of package distribution:
see \code{\LinkA{install.packages}{install.packages}}.
\item[\code{object}] an object of class \code{"packageStatus"} as returned by
\code{packageStatus}.
\item[\code{ask}] if \code{TRUE}, the user is prompted which packages should
be upgraded and which not.
\item[\code{...}] currently not used.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The URLs in \code{repositories} should be full paths to the
appropriate contrib sections of the repositories.  The default is
\code{contrib.url(getOption("repos"))}.

There are \code{print} and \code{summary} methods for the
\code{"packageStatus"} objects: the \code{print} method gives a brief
tabular summary and the \code{summary} method prints the results.

The \code{update} method updates the \code{"packageStatus"} object.
The \code{upgrade} method is similar to \code{\LinkA{update.packages}{update.packages}}:
it offers to install the current versions of those packages which are not
currently up-to-date.
\end{Details}
%
\begin{Value}
An object of class \code{"packageStatus"}.  This is a list with two
components

\begin{ldescription}
\item[\code{inst}] a data frame with columns as the \emph{matrix} returned by
\code{\LinkA{installed.packages}{installed.packages}} plus \code{"Status"}, a factor with
levels \code{c("ok", "upgrade")}.  Only the newest version of each
package is reported, in the first repository in which it appears.


\item[\code{avail}] a data frame with columns as the \emph{matrix} returned by
\code{\LinkA{available.packages}{available.packages}} plus \code{"Status"}, a factor with
levels \code{c("installed", "not installed", "unavailable")}..

\end{ldescription}
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{installed.packages}{installed.packages}}, \code{\LinkA{available.packages}{available.packages}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
x <- packageStatus()
print(x)
summary(x)
upgrade(x)
x <- update(x)
print(x)

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{page}{Invoke a Pager on an R Object}{page}
\keyword{utilities}{page}
%
\begin{Description}\relax
Displays a representation of the object named by \code{x} in a pager
\emph{via} \code{\LinkA{file.show}{file.show}}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
page(x, method = c("dput", "print"), ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] An \R{} object, or a character string naming an object.
\item[\code{method}] The default method is to dump the object \emph{via}
\code{\LinkA{dput}{dput}}.  An alternative is to use \code{print} and
capture the output to be shown in the pager.
\item[\code{...}] additional arguments for \code{\LinkA{dput}{dput}},
\code{\LinkA{print}{print}} or \code{\LinkA{file.show}{file.show}} (such as \code{title}).
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If \code{x} is a length-one character vector, it is used as the name
of an object to look up in the environment from which \code{page} is
called.   All other objects are displayed directly.

A default value of \code{title} is passed to \code{file.show} if one
is not supplied in \code{...}.
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{file.show}{file.show}}, \code{\LinkA{edit}{edit}}, \code{\LinkA{fix}{fix}}.

To go to a new page when graphing, see \code{\LinkA{frame}{frame}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: ## four ways to look at the code of 'page'
page(page)             # as an object
page("page")           # a character string
v <- "page"; page(v)   # a length-one character vector
page(utils::page)      # a call

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{person}{Person Names and Contact Information}{person}
\aliasA{as.character.person}{person}{as.character.person}
\aliasA{as.character.personList}{person}{as.character.personList}
\aliasA{as.person}{person}{as.person}
\methaliasA{as.person.default}{person}{as.person.default}
\aliasA{as.personList}{person}{as.personList}
\methaliasA{as.personList.default}{person}{as.personList.default}
\methaliasA{as.personList.person}{person}{as.personList.person}
\aliasA{personList}{person}{personList}
\aliasA{toBibtex.person}{person}{toBibtex.person}
\aliasA{toBibtex.personList}{person}{toBibtex.personList}
\keyword{misc}{person}
%
\begin{Description}\relax
A class and utility methods for holding information about persons
like name and email address.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
person(first = "", last = "", middle = "", email = "")
personList(...)
as.person(x)
as.personList(x)

## S3 method for class 'person':
as.character(x, ...)
## S3 method for class 'personList':
as.character(x, ...)

## S3 method for class 'person':
toBibtex(object, ...)
## S3 method for class 'personList':
toBibtex(object, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{first}] character string, first name
\item[\code{middle}] character string, middle name(s)
\item[\code{last}] character string, last name
\item[\code{email}] character string, email address
\item[\code{...}] for \code{personList} an arbitrary number of \code{person}
objects
\item[\code{x}] a character string or an object of class \code{person} or
\code{personList}
\item[\code{object}] an object of class \code{person} or
\code{personList}
\end{ldescription}
\end{Arguments}
%
\begin{Examples}
\begin{ExampleCode}
## create a person object directly
p1 <- person("Karl", "Pearson", email = "pearson@stats.heaven")
p1

## convert a string
p2 <- as.person("Ronald Aylmer Fisher")
p2

## create one object holding both
p <- personList(p1, p2)
ps <- as.character(p)
ps
as.personList(ps)

## convert to BibTeX author field
toBibtex(p)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{PkgUtils}{Utilities for Building and Checking Add-on Packages}{PkgUtils}
\aliasA{build}{PkgUtils}{build}
\aliasA{check}{PkgUtils}{check}
\keyword{utilities}{PkgUtils}
%
\begin{Description}\relax
Utilities for checking whether the sources of an \R{} add-on package
work correctly, and for building a source or binary package from
them.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
R CMD build [options] pkgdirs
R CMD check [options] pkgdirs
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{pkgdirs}] a list of names of directories with sources of \R{}
add-on packages.  For \code{check} this can also be the filename of
a compressed \command{tar} archive with extension \file{.tar.gz} or
\file{.tgz} or \file{.tar.bz2}.
\item[\code{options}] further options to control the processing, or for
obtaining information about usage and version of the utility.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\command{R CMD check}
checks \R{} add-on packages from their sources, performing a wide
variety of diagnostic checks.

\command{R CMD build}
builds \R{} source or binary packages from their sources.  The name(s)
of the packages are taken from the \file{DESCRIPTION} files and not
from the directory names.

Use \command{R CMD \var{foo} --help}
to obtain usage information on utility \code{\var{foo}}.

Several of the options to \code{build --binary} are passed to
\code{\LinkA{INSTALL}{INSTALL}} so consult its help for the details.
\end{Details}
%
\begin{SeeAlso}\relax
The sections on ``Checking and building packages'' and
``Processing Rd format'' in ``Writing \R{} Extensions''
(see the \file{doc/manual} subdirectory of the \R{} source tree).

\code{\LinkA{INSTALL}{INSTALL}} is called by \code{build --binary}.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{prompt}{Produce Prototype of an R Documentation File}{prompt}
\methaliasA{prompt.data.frame}{prompt}{prompt.data.frame}
\methaliasA{prompt.default}{prompt}{prompt.default}
\keyword{documentation}{prompt}
%
\begin{Description}\relax
Facilitate the constructing of files documenting \R{} objects.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
prompt(object, filename = NULL, name = NULL, ...)

## Default S3 method:
prompt(object, filename = NULL, name = NULL,
       force.function = FALSE, ...)

## S3 method for class 'data.frame':
prompt(object, filename = NULL, name = NULL, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{object}] an \R{} object, typically a function for the default
method.  Can be \code{\LinkA{missing}{missing}} when \code{name} is specified.
\item[\code{filename}] usually, a connection or a character string giving the
name of the file to which the documentation shell should be written.
The default corresponds to a file whose name is \code{name} followed
by \code{".Rd"}.  Can also be \code{NA} (see below).
\item[\code{name}] a character string specifying the name of the object.
\item[\code{force.function}] a logical.  If \code{TRUE}, treat \code{object}
as function in any case.
\item[\code{...}] further arguments passed to or from other methods.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Unless \code{filename} is \code{NA}, a documentation shell for
\code{object} is written to the file specified by \code{filename}, and
a message about this is given.  For function objects, this shell
contains the proper function and argument names.  R documentation
files thus created still need to be edited and moved into the
\file{man} subdirectory of the package containing the object to be
documented.

If \code{filename} is \code{NA}, a list-style representation of the
documentation shell is created and returned.  Writing the shell to a
file amounts to \code{cat(unlist(x), file = filename, sep = "\bsl{}n")},
where \code{x} is the list-style representation.

When \code{prompt} is used in \code{\LinkA{for}{for}} loops or scripts, the
explicit \code{name} specification will be useful.
\end{Details}
%
\begin{Value}
If \code{filename} is \code{NA}, a list-style representation of the
documentation shell.  Otherwise, the name of the file written to is
returned invisibly.
\end{Value}
%
\begin{Section}{Warning}
The default filename may not be a valid filename under limited file
systems (e.g. those on Windows).

Currently, calling \code{prompt} on a non-function object assumes that
the object is in fact a data set and hence documents it as such.  This
may change in future versions of \R{}.  Use \code{\LinkA{promptData}{promptData}} to
create documentation skeletons for data sets.
\end{Section}
%
\begin{Note}\relax
The documentation file produced by \code{prompt.data.frame} does not
have the same format as many of the data frame documentation files in
the \pkg{base} package.  We are trying to settle on a preferred
format for the documentation.
\end{Note}
%
\begin{Author}\relax
Douglas Bates for \code{prompt.data.frame}
\end{Author}
%
\begin{References}\relax
Becker, R. A., Chambers, J. M. and Wilks, A. R. (1988)
\emph{The New S Language}.
Wadsworth \& Brooks/Cole.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{promptData}{promptData}}, \code{\LinkA{help}{help}} and the chapter on
``Writing \R{} documentation'' in ``Writing \R{} Extensions''
(see the \file{doc/manual} subdirectory of the \R{} source tree).

For creation of many help pages (for a package),
see \code{\LinkA{package.skeleton}{package.skeleton}}.

To prompt the user for input, see \code{\LinkA{readline}{readline}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(graphics)
prompt(plot.default)
prompt(interactive, force.function = TRUE)
unlink("plot.default.Rd")
unlink("interactive.Rd")

prompt(women) # data.frame
unlink("women.Rd")

prompt(sunspots) # non-data.frame data
unlink("sunspots.Rd")

## Not run: 
## Create a help file for each function in the .GlobalEnv:
for(f in ls()) if(is.function(get(f))) prompt(name = f)

## End(Not run)

\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{promptData}{Generate a Shell for Documentation of Data Sets}{promptData}
\keyword{documentation}{promptData}
%
\begin{Description}\relax
Generates a shell of documentation for a data set.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
promptData(object, filename = NULL, name = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{object}] an \R{} object to be documented as a data set.
\item[\code{filename}] usually, a connection or a character string giving the
name of the file to which the documentation shell should be written.
The default corresponds to a file whose name is \code{name} followed
by \code{".Rd"}.  Can also be \code{NA} (see below).
\item[\code{name}] a character string specifying the name of the object.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Unless \code{filename} is \code{NA}, a documentation shell for
\code{object} is written to the file specified by \code{filename}, and
a message about this is given.

If \code{filename} is \code{NA}, a list-style representation of the
documentation shell is created and returned.  Writing the shell to a
file amounts to \code{cat(unlist(x), file = filename, sep = "\bsl{}n")},
where \code{x} is the list-style representation.

Currently, only data frames are handled explicitly by the code.
\end{Details}
%
\begin{Value}
If \code{filename} is \code{NA}, a list-style representation of the
documentation shell.  Otherwise, the name of the file written to is
returned invisibly.
\end{Value}
%
\begin{Section}{Warning}
This function is still experimental.  Both interface and value might
change in future versions.  In particular, it may be preferable to use
a character string naming the data set and optionally a specification
of where to look for it instead of using \code{object}/\code{name} as
we currently do.  This would be different from \code{\LinkA{prompt}{prompt}},
but consistent with other prompt-style functions in package
\pkg{methods}, and also allow prompting for data set documentation
without explicitly having to load the data set.
\end{Section}
%
\begin{SeeAlso}\relax
\code{\LinkA{prompt}{prompt}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
promptData(sunspots)
unlink("sunspots.Rd")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{promptPackage}{Generate a Shell for Documentation of a Package}{promptPackage}
\keyword{documentation}{promptPackage}
%
\begin{Description}\relax
Generates a shell of documentation for an installed or source package.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
promptPackage(package, lib.loc = NULL, filename = NULL,
              name = NULL, final = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{package}] the name of an \emph{installed} or \emph{source} package
to be documented.
\item[\code{lib.loc}] a character vector describing the location of \R{}
library trees to search through, or \code{NULL}.  The default value
of \code{NULL} corresponds to all libraries currently known.  For a
source package this should specify the parent directory of the
package's sources.
\item[\code{filename}] usually, a connection or a character string giving the
name of the file to which the documentation shell should be written.
The default corresponds to a file whose name is \code{name} followed
by \code{".Rd"}.  Can also be \code{NA} (see below).
\item[\code{name}] a character string specifying the name of the help topic,
typically of the form \samp{<pkg>-package}.
\item[\code{final}] a logical value indicating whether to attempt to
create a usable version of the help topic, rather than just a shell.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Unless \code{filename} is \code{NA}, a documentation shell for
\code{package} is written to the file specified by \code{filename}, and
a message about this is given.  

If \code{filename} is \code{NA}, a list-style representation of the
documentation shell is created and returned.  Writing the shell to a
file amounts to \code{cat(unlist(x), file = filename, sep = "\bsl{}n")},
where \code{x} is the list-style representation.

If \code{final} is \code{TRUE}, the generated documentation will not
include the place-holder slots for manual editing, it will be usable
as-is.  In most cases a manually edited file is preferable (but
\code{final = TRUE} is certainly less work).
\end{Details}
%
\begin{Value}
If \code{filename} is \code{NA}, a list-style representation of the
documentation shell.  Otherwise, the name of the file written to is
returned invisibly.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{prompt}{prompt}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
filename <- tempfile()
promptPackage("utils", filename = filename)
file.show(filename)
unlink(filename)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{Question}{Documentation Shortcuts}{Question}
\aliasA{?}{Question}{?}
\keyword{documentation}{Question}
%
\begin{Description}\relax
These functions provide access to documentation.
Documentation on a topic with name \code{name} (typically, an \R{}
object or a data set) can be displayed by either \code{help("name")} or
\code{?name}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
?topic

type?topic
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{topic}] Usually, a \LinkA{name}{name} or character string specifying the
topic for which help is sought.

Alternatively, a function call to ask for documentation on a
corresponding S4 method: see the section on S4 method documentation.
The calls \code{\var{pkg}::\var{topic}} and
\code{\var{pkg}:::\var{topic}} are treated specially, and look for
help on \code{topic} in package \pkg{\var{pkg}}.

\item[\code{type}] the special type of documentation to use for this topic;
for example, if the type is \code{class}, documentation is
provided for the class with name \code{topic}.
See the Section `S4 Method Documentation' for the uses of
\code{type} to get help on formal methods, including
\code{methods?\var{function}} and \code{method?\var{call}}.

\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This is a shortcut to \code{\LinkA{help}{help}} and uses its default type of help.

Some topics need to be quoted (by \LinkA{backtick}{backtick}s) or given as a
character string.  There include those which cannot syntactically
appear on their own such as unary and binary operators,
\code{function} and control-flow \LinkA{reserved}{reserved} words (including
\code{if}, \code{else} \code{for}, \code{in}, \code{repeat},
\code{while}, \code{break} and \code{next}.  The other \code{reserved}
words can be used as if they were names, for example \code{TRUE},
\code{NA} and \code{Inf}.
\end{Details}
%
\begin{Section}{S4 Method Documentation}
Authors of formal (`S4') methods can provide documentation
on specific methods, as well as overall documentation on the methods
of a particular function.  The \code{"?"} operator allows access to
this documentation in three ways.

The expression \code{methods?\var{f}} will look for the overall
documentation methods for the function \code{\var{f}}.  Currently,
this means the documentation file containing the alias
\code{\var{f}-methods}.

There are two different ways to look for documentation on a
particular method.  The first is to supply the \code{topic} argument
in the form of a function call, omitting the \code{type} argument.
The effect is to look for documentation on the method that would be
used if this function call were actually evaluated. See the examples
below.  If the function is not a generic (no S4 methods are defined
for it), the help reverts to documentation on the function name.

The \code{"?"} operator can also be called with \code{doc\_type} supplied
as \code{method}; in this case also, the \code{topic} argument is
a function call, but the arguments are now interpreted as specifying
the class of the argument, not the actual expression that will
appear in a real call to the function.  See the examples below.

The first approach will be tedious if the actual call involves
complicated expressions, and may be slow if the arguments take a
long time to evaluate.  The second approach avoids these
issues, but you do have to know what the classes of the actual
arguments will be when they are evaluated.

Both approaches make use of any inherited methods; the signature of
the method to be looked up is found by using \code{selectMethod}
(see the documentation for \code{\LinkA{getMethod}{getMethod}}).
\end{Section}
%
\begin{References}\relax
Becker, R. A., Chambers, J. M. and Wilks, A. R. (1988)
\emph{The New S Language}.
Wadsworth \& Brooks/Cole.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{help}{help}}

\code{\LinkA{??}{??}} for finding help pages on a vague topic.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
?lapply

?"for"                  # but quotes/backticks are needed
?`+`

?women                  # information about data set "women"

## Not run: 
require(methods)
## define a S4 generic function and some methods
combo <- function(x, y) c(x, y)
setGeneric("combo")
setMethod("combo", c("numeric", "numeric"), function(x, y) x+y)

## assume we have written some documentation
## for combo, and its methods ....

?combo  # produces the function documentation

methods?combo  # looks for the overall methods documentation

method?combo("numeric", "numeric")  # documentation for the method above

?combo(1:10, rnorm(10))  # ... the same method, selected according to
                         # the arguments (one integer, the other numeric)

?combo(1:10, letters)    # documentation for the default method

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{rcompgen}{A Completion Generator for R}{rcompgen}
\aliasA{.DollarNames}{rcompgen}{.DollarNames}
\methaliasA{.DollarNames.default}{rcompgen}{.DollarNames.default}
\methaliasA{.DollarNames.environment}{rcompgen}{.DollarNames.environment}
\methaliasA{.DollarNames.list}{rcompgen}{.DollarNames.list}
\aliasA{rc.getOption}{rcompgen}{rc.getOption}
\aliasA{rc.options}{rcompgen}{rc.options}
\aliasA{rc.settings}{rcompgen}{rc.settings}
\aliasA{rc.status}{rcompgen}{rc.status}
\keyword{IO}{rcompgen}
\keyword{utilities}{rcompgen}
%
\begin{Description}\relax
This package provides a mechanism to generate relevant completions
from a partially completed command line.  It is not intended to be
useful by itself, but rather in conjunction with other mechanisms that
use it as a backend.  The functions listed in the usage section
provide a simple control and query mechanism.  The actual interface
consists of a few unexported functions described further down.
\end{Description}
%
\begin{Usage}
\begin{verbatim}

rc.settings(ops, ns, args, func, ipck, S3, data, help,
            argdb, files)

rc.status()
rc.getOption(name)
rc.options(...)

.DollarNames(x, pattern)

## Default S3 method:
.DollarNames(x, pattern = "")
## S3 method for class 'list':
.DollarNames(x, pattern = "")
## S3 method for class 'environment':
.DollarNames(x, pattern = "")

\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{ops, ns, args, func, ipck, S3, data, help, argdb, files}] 

logical, turning some optional completion features on and off.

\begin{description}

\item[\code{ops}:] activates completion after the \code{\$} and
\code{@} operators

\item[\code{ns}:] controls name space related completions

\item[\code{args}:] enables completion of function arguments

\item[\code{func}:] enables detection of functions.  If enabled,
a customizable extension (\code{"("} by default) is appended to
function names.  The process of determining whether a potential
completion is a function requires evaluation, including for lazy
loaded symbols.  This is extremely undesirable for large
objects, because of potentially wasteful use of memory in
addition to the time overhead associated with loading.  For this
reason, this feature is disabled by default. 

\item[\code{S3}:]  when \code{args=TRUE}, activates completion on
arguments of all S3 methods (otherwise just the generic, which
usually has very few arguments) 

\item[\code{ipck}:]  enables completion of installed package names
inside \code{\LinkA{library}{library}} and \code{\LinkA{require}{require}} 

\item[\code{data}:]  enables completion of data sets (including
those already visible) inside \code{\LinkA{data}{data}} 

\item[\code{help}:]  enables completion of help requests starting
with a question mark, by looking inside help index files 

\item[\code{argdb}:]  when \code{args=TRUE}, completion is
attempted on function arguments.  Generally, the list of valid
arguments is determined by dynamic calls to \code{\LinkA{args}{args}}.
While this gives results that are technically correct, the use
of the \code{...} argument often hides some useful arguments.
To give more flexibility in this regard, an optional table of
valid arguments names for specific functions is retained
internally.  Setting \code{argdb=TRUE} enables preferential
lookup in this internal data base for functions with an entry in
it.  Of course, this is useful only when the data base contains
information about the function of interest.  Some functions are
included in the package (the maintainer is happy to add more
upon request), and more can be added by the user through the
unexported function \code{.addFunctionInfo} (see below).


\item[\code{files}:]  enables filename completion in R code.  This
is initially set to \code{FALSE}, in which case the underlying
completion front-end can take over (and hopefully do a better
job than we would have done).  For systems where no such
facilities exist, this can be set to \code{TRUE} if file name
completion is desired.  This is currently experimental and may
not work very well.  


\end{description}


All settings are turned on by default except \code{ipck},
\code{func} and \code{files}.  Turn more off if your CPU cycles are
valuable; you will still retain basic completion on names of objects
in the search list.  See below for additional details.

\item[\code{name, ...}]  user-settable options.  Currently valid names are
\begin{description}

\item[\code{function.suffix}:]  default \code{"("} 
\item[\code{funarg.suffix}:]  default \code{" = "} 
\item[\code{package.suffix}]  default \code{"::"} 

\end{description}

See \code{\LinkA{options}{options}} for detailed usage description


\item[\code{x}]  An R object for which valid names after \code{"\$"}
are computed and returned.


\item[\code{pattern}]  A regular expression.  Only matching names are
returned.

\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
There are several types of completion, some of which can be disabled
using \code{rc.settings}.  The most basic level, which can not be
turned off once the package is loaded, provides completion on names
visible on the search path, along with a few special keywords
(e.g. \code{TRUE}).  This type of completion is not attempted if the
partial `word' (a.k.a. token) being completed is empty (since
there would be too many completions).  The more advanced types of
completion are described below.

\begin{description}


\item[\bold{Completion after extractors \code{\$} and \code{@}}:] 
When the \code{ops} setting is turned on, completion after
\code{\$} and \code{@} is attempted.  This requires the prefix to
be evaluated, which is attempted unless it involves an explicit
function call (implicit function calls involving the use of
\code{[}, \code{\$}, etc \emph{do not} inhibit evaluation).

Valid completions after the \code{\$} extractor are determined by
the generic function \code{.DollarNames}.  Some basic methods are
provided, and more can be written for custom classes.


\item[\bold{Completion inside name spaces}:] 
When the \code{ns} setting is turned on, completion inside
name spaces is attempted when a token is preceded by the \code{::}
or \code{:::} operators.  Additionally, the basic completion
mechanism is extended to include attached name spaces, or more
precisely, \code{foopkg::} becomes a valid completion of
\code{foo} if the return value of \code{\LinkA{search}{search}()} includes
the string \code{"package:foopkg"}.

The completion of package name spaces applies only to attached
packages, i.e. if \code{MASS} is not attached (whether or not it
is loaded), \code{MAS} will not complete to \code{MASS::}.
However, attempted completion \emph{inside} an apparent name space
will attempt to load the name space if it is not already loaded,
e.g. trying to complete on \code{MASS::fr} will load \code{MASS}
(but not necessarily attach it) even if it is not already loaded.


\item[\bold{Completion of function arguments}:] 
When the \code{args} setting is turned on, completion on function
arguments is attempted whenever deemed appropriate.  The mechanism
used will currently fail if the relevant function (at the point
where completion is requested) was entered on a previous prompt
(which implies in particular that the current line is being typed
in response to a continuation prompt, usually \code{+}).  Note
that separation by newlines is fine.

The list of possible argument completions that is generated can be
misleading.  There is no problem for non-generic functions (except
that \code{...} is listed as a completion; this is intentional
as it signals the fact that the function can accept further
arguments).  However, for generic functions, it is practically
impossible to give a reliable argument list without evaluating
arguments (and not even then, in some cases), which is risky (in
addition to being difficult to code, which is the real reason it
hasn't even been tried), especially when that argument is itself
an inline function call.  Our compromise is to consider arguments
of \emph{all} currently available methods of that generic.  This
has two drawbacks.  First, not all listed completions may be
appropriate in the call currently being constructed.  Second, for
generics with many methods (like \code{print} and \code{plot}),
many matches will need to be considered, which may take a
noticeable amount of time.  Despite these drawbacks, we believe
this behaviour to be more useful than the only other practical
alternative, which is to list arguments of the generic only.

Only S3 methods are currently supported in this fashion, and that
can be turned off using the \code{S3} setting.

Since arguments can be unnamed in \R{} function calls, other types
of completion are also appropriate whenever argument completion
is.  Since there are usually many many more visible objects than
formal arguments of any particular function, possible argument
completions are often buried in a bunch of other possibilities.
However, recall that basic completion is suppressed for blank
tokens.  This can be useful to list possible arguments of a
function.  For example, trying to complete \code{seq([TAB]} and
\code{seq(from = 1, [TAB])} will both list only the arguments of
\code{seq} (or any of its methods), whereas trying to complete
\code{seq(length[TAB]} will list both the \code{length.out}
argument and the \code{length(} function as possible completions.
Note that no attempt is made to remove arguments already supplied,
as that would incur a further speed penalty.


\item[\bold{Special functions}:] 
For a few special functions (\code{\LinkA{library}{library}},
\code{\LinkA{data}{data}}, etc), the first argument is treated specially,
in the sense that normal completion is suppressed, and some
function specific completions are enabled if so requested by the
settings.  The \code{ipck} setting, which controls whether
\code{\LinkA{library}{library}} and \code{\LinkA{require}{require}} will complete on
\emph{installed packages}, is disabled by default because the
first call to \code{\LinkA{installed.packages}{installed.packages}} is potentially time
consuming (e.g. when packages are installed on a remote network
file server).  Note, however, that the results of a call to
\code{\LinkA{installed.packages}{installed.packages}} is cached, so subsequent calls
are usually fast, so turning this option on is not particularly
onerous even in such situations.




\end{description}

\end{Details}
%
\begin{Value}
\code{rc.status} returns, as a list, the contents of an internal
(unexported) environment that is used to record the results of the
last completion attempt.  This can be useful for debugging.  For such
use, one must resist the temptation to use completion when typing the
call to \code{rc.status} itself, as that then becomes the last attempt
by the time the call is executed.

The items of primary interest in the returned list are:

\begin{ldescription}
\item[\code{comps}]  the possible completions generated by the last
call to \code{.completeToken}, as a character vector 
\item[\code{token}]  the token that was (or, is to be) completed, as
set by the last call to \code{.assignToken} (possibly inside a call
to \code{.guessTokenFromLine}) 
\item[\code{linebuffer}]  the full line, as set by the last call to
\code{.assignLinebuffer}
\item[\code{start}]  the start position of the token in the line
buffer, as set by the last call to \code{.assignStart} 
\item[\code{end}]  the end position of the token in the line
buffer, as set by the last call to \code{.assignEnd} 
\item[\code{fileName}]  logical, indicating whether the cursor is
currently inside quotes.  If so, no completion is attempted.  A
reasonable default behaviour for the backend in that case is to fall
back to filename completion.  
\item[\code{fguess}]  the name of the function \code{rcompgen} thinks
the cursor is currently inside 
\item[\code{isFirstArg}]  logical.  If cursor is inside a function, is it the
first argument? 

\end{ldescription}
In addition, the components \code{settings} and \code{options} give
the current values of settings and options respectively.

\code{rc.getOption} and \code{rc.options} behave much like
\code{\LinkA{getOption}{getOption}} and \code{\LinkA{options}{options}} respectively.
\end{Value}
%
\begin{Section}{Unexported API}

There are several unexported functions in the package.  Of these, 
a few are special because they provide the API through which other
mechanisms can make use of the facilities provided by this package
(they are unexported because they are not meant to be called directly
by users).  The usage of these functions are:

\begin{alltt}
    .assignToken(text)
    .assignLinebuffer(line)
    .assignStart(start)
    .assignEnd(end)

    .completeToken()
    .retrieveCompletions()
    .getFileComp()

    .guessTokenFromLine()
    .win32consoleCompletion(linebuffer, cursorPosition,
                            check.repeat = TRUE, 
                            minlength = -1)

    .addFunctionInfo(...)
  \end{alltt}


The first four functions set up a completion attempt by specifying the
token to be completed (\code{text}), and indicating where
(\code{start} and \code{end}, which should be integers) the token is
placed within the complete line typed so far (\code{line}).

Potential completions of the token are generated by
\code{.completeToken}, and the completions can be retrieved as an \R{}
character vector using \code{.retrieveCompletions}.

If the cursor is inside quotes, no completion is attempted.  The
function \code{.getFileComp} can be used after a call to
\code{.completeToken} to determine if this is the case (returns
\code{TRUE}), and alternative completions generated as deemed useful.
In most cases, filename completion is a reasonable fallback.

The \code{.guessTokenFromLine} function is provided for use with
backends that do not already break a line into tokens.  It requires
the linebuffer and endpoint (cursor position) to be already set, and
itself sets the token and the start position.  It returns the token as
a character string.  (This is used by the ESS completion hook example
given in the \code{examples/altesscomp.el} file.)

The \code{.win32consoleCompletion} is similar in spirit, but is more
geared towards the Windows GUI (or rather, any front-end that has no
completion facilities of its own).  It requires the linebuffer
and cursor position as arguments, and returns a list with three
components, \code{addition}, \code{possible} and \code{comps}.  If
there is an unambiguous extension at the current position,
\code{addition} contains the additional text that should be inserted
at the cursor.  If there is more than one possibility, these are
available either as a character vector of preformatted strings in
\code{possible}, or as a single string in \code{comps}.
\code{possible} consists of lines formatted using the current
\code{width} option, so that printing them on the console one line at
a time will be a reasonable way to list them.  \code{comps} is a space
separated (collapsed) list of the same completions, in case the
front-end wishes to display it in some other fashion.

The \code{minlength} argument can be used to suppress completion when
the token is too short (which can be useful if the front-end is set up
to try completion on every keypress).  If \code{check.repeat} is
\code{TRUE}, it is detected if the same completion is being requested
more than once in a row, and ambiguous completions are returned only
in that case.  This is an attempt to emulate GNU Readline behaviour,
where a single TAB completes up to any unambiguous part, and multiple
possibilities are reported only on two consecutive TABs.

As the various front-end interfaces evolve, the details of these
functions are likely to change as well.

The function \code{.addFunctionInfo} can be used to add information
about the permitted argument names for specific functions.  Multiple
named arguments are allowed in calls to it, where the tags are names
of functions and values are character vectors representing valid
arguments.  When the \code{argdb} setting is \code{TRUE}, these are
used as a source of valid argument names for the relevant functions.
\end{Section}
%
\begin{Note}\relax
If you are uncomfortable with unsolicited evaluation of pieces of
code, you should set \code{ops = FALSE}.  Otherwise, trying to
complete \code{foo@ba} will evaluate \code{foo}, trying to complete
\code{foo[i,1:10]\$ba} will evaluate \code{foo[i,1:10]}, etc.  This
should not be too bad, as explicit function calls (involving
parentheses) are not evaluated in this manner.  However, this
\emph{will} affect lazy loaded symbols (and presumably other promise
type thingies).
\end{Note}
%
\begin{Author}\relax
 Deepayan Sarkar, \email{deepayan.sarkar@r-project.org} 
\end{Author}
\inputencoding{latin1}
\HeaderA{read.DIF}{Data Input from Spreadsheet}{read.DIF}
\keyword{file}{read.DIF}
\keyword{connection}{read.DIF}
%
\begin{Description}\relax
Reads a file in Data Interchange Format (DIF) and creates a data frame
from it.  DIF is a format for data matrices such as single spreadsheets.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
read.DIF(file, header = FALSE,
           dec = ".", row.names, col.names,
           as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE,
           blank.lines.skip = TRUE,
           stringsAsFactors = default.stringsAsFactors(),
           transpose = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] the name of the file which the data are to be read from,
or a connection, or a complete URL.


\item[\code{header}] a logical value indicating whether the spreadsheet contains the
names of the variables as its first line.  If missing, the value is
determined from the file format: \code{header} is set to \code{TRUE}
if and only if the first row contains only character values and
the top left cell is empty.

\item[\code{dec}] the character used in the file for decimal points.

\item[\code{row.names}] a vector of row names.  This can be a vector giving
the actual row names, or a single number giving the column of the
table which contains the row names, or character string giving the
name of the table column containing the row names.

If there is a header and the first row contains one fewer field than
the number of columns, the first column in the input is used for the
row names.  Otherwise if \code{row.names} is missing, the rows are
numbered.

Using \code{row.names = NULL} forces row numbering.


\item[\code{col.names}] a vector of optional names for the variables.
The default is to use \code{"V"} followed by the column number.

\item[\code{as.is}] the default behavior of \code{read.DIF} is to convert
character variables (which are not converted to logical, numeric or
complex) to factors.  The variable \code{as.is} controls the
conversion of columns not otherwise specified by \code{colClasses}.
Its value is either a vector of logicals (values are recycled if
necessary), or a vector of numeric or character indices which
specify which columns should not be converted to factors.

Note: to suppress all conversions including those of numeric
columns, set \code{colClasses = "character"}.

Note that \code{as.is} is specified per column (not per
variable) and so includes the column of row names (if any) and any
columns to be skipped.


\item[\code{na.strings}] a character vector of strings which are to be
interpreted as \code{\LinkA{NA}{NA}} values.  Blank fields are also
considered to be missing values in logical, integer, numeric and
complex fields.

\item[\code{colClasses}] character.  A vector of classes to be assumed for
the columns.  Recycled as necessary, or if the character vector is
named, unspecified values are taken to be \code{NA}.

Possible values are \code{NA} (when \code{\LinkA{type.convert}{type.convert}} is
used), \code{"NULL"} (when the column is skipped), one of the atomic
vector classes (logical, integer, numeric, complex, character, raw),
or \code{"factor"}, \code{"Date"} or \code{"POSIXct"}.  Otherwise
there needs to be an \code{as} method (from package \pkg{methods})
for conversion from \code{"character"} to the specified formal
class.

Note that \code{colClasses} is specified per column (not per
variable) and so includes the column of row names (if any).


\item[\code{nrows}] the maximum number of rows to read in.  Negative values
are ignored.

\item[\code{skip}] the number of lines of the data file to skip before
beginning to read data.

\item[\code{check.names}] logical.  If \code{TRUE} then the names of the
variables in the data frame are checked to ensure that they are
syntactically valid variable names.  If necessary they are adjusted
(by \code{\LinkA{make.names}{make.names}}) so that they are, and also to ensure
that there are no duplicates.

\item[\code{blank.lines.skip}] logical: if \code{TRUE} blank lines in the
input are ignored.

\item[\code{stringsAsFactors}] logical: should character vectors be converted
to factors?

\item[\code{transpose}] logical, indicating if the row and column
interpretation should be transposed.  Microsoft's Excel has been
known to produce (non-standard conforming) DIF files which would
need \code{transpose = TRUE} to be read correctly.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
A data frame (\code{\LinkA{data.frame}{data.frame}}) containing a representation of
the data in the file.  Empty input is an error unless \code{col.names}
is specified, when a 0-row data frame is returned: similarly giving
just a header line if \code{header = TRUE} results in a 0-row data frame.
\end{Value}
%
\begin{Note}\relax
The columns referred to in \code{as.is} and \code{colClasses} include
the column of row names (if any).

Less memory will be used if \code{colClasses} is specified as one of
the six atomic vector classes.
\end{Note}
%
\begin{Author}\relax
R Core; \code{transpose} option by
Christoph Buser, ETH Zurich
\end{Author}
%
\begin{References}\relax
The DIF format specification can be found by searching on
\url{http://www.wotsit.org/}; the optional header fields are ignored.
See also
\url{http://en.wikipedia.org/wiki/Data_Interchange_Format}.

The term is likely to lead to confusion: Windows will have a
`Windows Data Interchange Format (DIF) data format' as part of
its WinFX system, which may or may not be compatible.
\end{References}
%
\begin{SeeAlso}\relax
The \emph{R Data Import/Export} manual.

\code{\LinkA{scan}{scan}}, \code{\LinkA{type.convert}{type.convert}},
\code{\LinkA{read.fwf}{read.fwf}} for reading \emph{f}ixed \emph{w}idth
\emph{f}ormatted input;
\code{\LinkA{read.table}{read.table}};
\code{\LinkA{data.frame}{data.frame}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## read.DIF() needs transpose=TRUE for file exported from Excel
udir <- system.file("misc", package="utils")
dd <- read.DIF(file.path(udir, "exDIF.dif"), header= TRUE, transpose=TRUE)
dc <- read.csv(file.path(udir, "exDIF.csv"), header= TRUE)
stopifnot(identical(dd,dc), dim(dd) == c(4,2))
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{read.fortran}{Read fixed-format data}{read.fortran}
\keyword{file}{read.fortran}
\keyword{connection}{read.fortran}
%
\begin{Description}\relax
Read fixed-format data files using Fortran-style format specifications.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
read.fortran(file, format, ..., as.is = TRUE, colClasses = NA)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] File or connection to read from
\item[\code{format}] Character vector or list of vectors.  See
`Details' below.
\item[\code{...}] Other arguments for \code{read.table}
\item[\code{as.is}] Keep characters as characters?
\item[\code{colClasses}] Variable classes to override defaults. See
\code{\LinkA{read.table}{read.table}} for details.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The format for a field is of one of the following forms: \code{rFl.d},
\code{rDl.d}, \code{rXl}, \code{rAl}, \code{rIl}, where \code{l} is
the number of columns, \code{d} is the number of decimal places, and
\code{r} is the number of repeats. \code{F} and \code{D} are numeric
formats, \code{A} is character, \code{I} is integer, and \code{X}
indicates columns to be skipped. The repeat code \code{r} and decimal
place code \code{d} are always optional. The length code \code{l} is
required except for \code{X} formats when \code{r} is present.

For a single-line record, \code{format} should be a character
vector. For a multiline record it should be a list with a character
vector for each line.

Skipped (\code{X}) columns are not passed to \code{read.table}, so
\code{colClasses}, \code{col.names}, and similar arguments passed to
\code{read.table} should not reference these columns.
\end{Details}
%
\begin{Value}
A data frame
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{read.fwf}{read.fwf}}, \code{\LinkA{read.csv}{read.csv}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
ff <- tempfile()
cat(file=ff, "123456", "987654", sep="\n")
read.fortran(ff, c("F2.1","F2.0","I2"))
read.fortran(ff, c("2F1.0","2X","2A1"))
unlink(ff)
cat(file=ff, "123456AB", "987654CD", sep="\n")
read.fortran(ff, list(c("2F3.1","A2"), c("3I2","2X")))
unlink(ff)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{read.fwf}{Read Fixed Width Format Files}{read.fwf}
\keyword{file}{read.fwf}
\keyword{connection}{read.fwf}
%
\begin{Description}\relax
Read a table of \bold{f}ixed \bold{w}idth \bold{f}ormatted
data into a \code{\LinkA{data.frame}{data.frame}}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
read.fwf(file, widths, header = FALSE, sep = "\t",
         skip = 0, row.names, col.names, n = -1,
         buffersize = 2000, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] 
the name of the file which the data are to be read from.

Alternatively, \code{file} can be a \code{\LinkA{connection}{connection}}, which
will be opened if necessary, and if so closed at the end of the
function call.

\item[\code{widths}] integer vector, giving the widths of the fixed-width
fields (of one line), or list of integer vectors giving widths for
multiline records.
\item[\code{header}] a logical value indicating whether the file contains the
names of the variables as its first line.  If present, the names
must be delimited by \code{sep}.
\item[\code{sep}] character; the separator used internally; should be a
character that does not occur in the file (except in the header).
\item[\code{skip}] number of initial lines to skip; see
\code{\LinkA{read.table}{read.table}}.
\item[\code{row.names}] see \code{\LinkA{read.table}{read.table}}.
\item[\code{col.names}] see \code{\LinkA{read.table}{read.table}}.
\item[\code{n}] the maximum number of records (lines) to be read, defaulting
to no limit.
\item[\code{buffersize}] Maximum number of lines to read at one time
\item[\code{...}] further arguments to be passed to
\code{\LinkA{read.table}{read.table}}.  Useful further arguments include
\code{as.is}, \code{na.strings}, \code{colClasses} and \code{strip.white}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Multiline records are concatenated to a single line before processing.
Fields that are of zero-width or are wholly beyond the end of the line
in \code{file} are replaced by \code{NA}.

Negative-width fields are used to indicate columns to be skipped, eg
\code{-5} to skip 5 columns.  These fields are not seen by
\code{read.table} and so should not be included in a \code{col.names}
or \code{colClasses} argument (nor in the header line, if present).

Reducing the \code{buffersize} argument may reduce memory use when
reading large files with long lines.  Increasing \code{buffersize} may
result in faster processing when enough memory is available.
\end{Details}
%
\begin{Value}
A \code{\LinkA{data.frame}{data.frame}} as produced by \code{\LinkA{read.table}{read.table}}
which is called internally.
\end{Value}
%
\begin{Author}\relax
Brian Ripley for \R{} version: original \code{Perl} by Kurt Hornik.
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{scan}{scan}} and \code{\LinkA{read.table}{read.table}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
ff <- tempfile()
cat(file=ff, "123456", "987654", sep="\n")
read.fwf(ff, widths=c(1,2,3))    #> 1 23 456 \ 9 87 654
read.fwf(ff, widths=c(1,-2,3))   #> 1 456 \ 9 654
unlink(ff)
cat(file=ff, "123", "987654", sep="\n")
read.fwf(ff, widths=c(1,0, 2,3))    #> 1 NA 23 NA \ 9 NA 87 654
unlink(ff)
cat(file=ff, "123456", "987654", sep="\n")
read.fwf(ff, widths=list(c(1,0, 2,3), c(2,2,2))) #> 1 NA 23 456 98 76 54
unlink(ff)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{read.socket}{Read from or Write to a Socket}{read.socket}
\aliasA{write.socket}{read.socket}{write.socket}
\keyword{misc}{read.socket}
%
\begin{Description}\relax
\code{read.socket} reads a string from the specified socket,
\code{write.socket} writes to the specified socket.  There is very
little error checking done by either.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
read.socket(socket, maxlen = 256, loop = FALSE)
write.socket(socket, string)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{socket}] a socket object
\item[\code{maxlen}] maximum length of string to read
\item[\code{loop}] wait for ever if there is nothing to read?
\item[\code{string}] string to write to socket
\end{ldescription}
\end{Arguments}
%
\begin{Value}
\code{read.socket} returns the string read.
\end{Value}
%
\begin{Author}\relax
Thomas Lumley
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{close.socket}{close.socket}}, \code{\LinkA{make.socket}{make.socket}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
finger <- function(user, host = "localhost", port = 79, print = TRUE)
{
    if (!is.character(user))
        stop("user name must be a string")
    user <- paste(user,"\r\n")
    socket <- make.socket(host, port)
    on.exit(close.socket(socket))
    write.socket(socket, user)
    output <- character(0)
    repeat{
        ss <- read.socket(socket)
        if (ss == "") break
        output <- paste(output, ss)
    }
    close.socket(socket)
    if (print) cat(output)
    invisible(output)
}
## Not run: 
finger("root")  ## only works if your site provides a finger daemon
## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{read.table}{Data Input}{read.table}
\aliasA{read.csv}{read.table}{read.csv}
\aliasA{read.csv2}{read.table}{read.csv2}
\aliasA{read.delim}{read.table}{read.delim}
\aliasA{read.delim2}{read.table}{read.delim2}
\keyword{file}{read.table}
\keyword{connection}{read.table}
%
\begin{Description}\relax
Reads a file in table format and creates a data frame from it, with
cases corresponding to lines and variables to fields in the file.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", row.names, col.names,
           as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = default.stringsAsFactors(),
           fileEncoding = "", encoding = "unknown")

read.csv(file, header = TRUE, sep = ",", quote="\"", dec=".",
         fill = TRUE, comment.char="", ...)

read.csv2(file, header = TRUE, sep = ";", quote="\"", dec=",",
          fill = TRUE, comment.char="", ...)

read.delim(file, header = TRUE, sep = "\t", quote="\"", dec=".",
           fill = TRUE, comment.char="", ...)

read.delim2(file, header = TRUE, sep = "\t", quote="\"", dec=",",
            fill = TRUE, comment.char="", ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] the name of the file which the data are to be read from.
Each row of the table appears as one line of the file.  If it does
not contain an \emph{absolute} path, the file name is
\emph{relative} to the current working directory,
\code{\LinkA{getwd}{getwd}()}. Tilde-expansion is performed where supported.
As from \R{} 2.10.0 this can be a compressed file (see \code{\LinkA{file}{file}}).

Alternatively, \code{file} can be a readable text-mode
\code{\LinkA{connection}{connection}} (which will be opened for reading if
necessary, and if so \code{\LinkA{close}{close}}d (and hence destroyed) at
the end of the function call).  (If \code{\LinkA{stdin}{stdin}()} is used,
the prompts for lines may be somewhat confusing.  Terminate input
with a blank line or an EOF signal, \code{Ctrl-D} on Unix and
\code{Ctrl-Z} on Windows.  Any pushback on \code{stdin()} will be
cleared before return.)

\code{file} can also be a complete URL.


\item[\code{header}] a logical value indicating whether the file contains the
names of the variables as its first line.  If missing, the value is
determined from the file format: \code{header} is set to \code{TRUE}
if and only if the first row contains one fewer field than the
number of columns.

\item[\code{sep}] the field separator character.  Values on each line of the
file are separated by this character.  If \code{sep = ""} (the
default for \code{read.table}) the separator is `white space',
that is one or more spaces, tabs, newlines or carriage returns.

\item[\code{quote}] the set of quoting characters. To disable quoting
altogether, use \code{quote = ""}.  See \code{\LinkA{scan}{scan}} for the
behaviour on quotes embedded in quotes.  Quoting is only considered
for columns read as character, which is all of them unless
\code{colClasses} is specified.

\item[\code{dec}] the character used in the file for decimal points.

\item[\code{row.names}] a vector of row names.  This can be a vector giving
the actual row names, or a single number giving the column of the
table which contains the row names, or character string giving the
name of the table column containing the row names.

If there is a header and the first row contains one fewer field than
the number of columns, the first column in the input is used for the
row names.  Otherwise if \code{row.names} is missing, the rows are
numbered.

Using \code{row.names = NULL} forces row numbering. Missing or
\code{NULL} \code{row.names} generate row names that are considered
to be `automatic' (and not preserved by \code{\LinkA{as.matrix}{as.matrix}}).


\item[\code{col.names}] a vector of optional names for the variables.
The default is to use \code{"V"} followed by the column number.

\item[\code{as.is}] the default behavior of \code{read.table} is to convert
character variables (which are not converted to logical, numeric or
complex) to factors.  The variable \code{as.is} controls the
conversion of columns not otherwise specified by \code{colClasses}.
Its value is either a vector of logicals (values are recycled if
necessary), or a vector of numeric or character indices which
specify which columns should not be converted to factors.

Note: to suppress all conversions including those of numeric
columns, set \code{colClasses = "character"}.

Note that \code{as.is} is specified per column (not per
variable) and so includes the column of row names (if any) and any
columns to be skipped.


\item[\code{na.strings}] a character vector of strings which are to be
interpreted as \code{\LinkA{NA}{NA}} values.  Blank fields are also
considered to be missing values in logical, integer, numeric and
complex fields.

\item[\code{colClasses}] character.  A vector of classes to be assumed for
the columns.  Recycled as necessary, or if the character vector is
named, unspecified values are taken to be \code{NA}.

Possible values are \code{NA} (when \code{\LinkA{type.convert}{type.convert}} is
used), \code{"NULL"} (when the column is skipped), one of the atomic
vector classes (logical, integer, numeric, complex, character, raw),
or \code{"factor"}, \code{"Date"} or \code{"POSIXct"}.  Otherwise
there needs to be an \code{as} method (from package \pkg{methods})
for conversion from \code{"character"} to the specified formal
class.

Note that \code{colClasses} is specified per column (not per
variable) and so includes the column of row names (if any).


\item[\code{nrows}] integer: the maximum number of rows to read in.  Negative
and other invalid values are ignored.

\item[\code{skip}] integer: the number of lines of the data file to skip before
beginning to read data.

\item[\code{check.names}] logical.  If \code{TRUE} then the names of the
variables in the data frame are checked to ensure that they are
syntactically valid variable names.  If necessary they are adjusted
(by \code{\LinkA{make.names}{make.names}}) so that they are, and also to ensure
that there are no duplicates.

\item[\code{fill}] logical. If \code{TRUE} then in case the rows have unequal
length, blank fields are implicitly added.  See `Details'.

\item[\code{strip.white}] logical. Used only when \code{sep} has
been specified, and allows the stripping of leading and trailing
white space from \code{character} fields (\code{numeric} fields
are always stripped).  See \code{\LinkA{scan}{scan}} for further details,
remembering that the columns may include the row names.

\item[\code{blank.lines.skip}] logical: if \code{TRUE} blank lines in the
input are ignored.

\item[\code{comment.char}] character: a character vector of length one
containing a single character or an empty string.  Use \code{""} to
turn off the interpretation of comments altogether.

\item[\code{allowEscapes}] logical.  Should C-style escapes such as
\samp{\bsl{}n} be processed or read verbatim (the default)?   Note that if
not within quotes these could be interpreted as a delimiter (but not
as a comment character).  For more details see \code{\LinkA{scan}{scan}}.

\item[\code{flush}] logical: if \code{TRUE}, \code{scan} will flush to the
end of the line after reading the last of the fields requested.
This allows putting comments after the last field.

\item[\code{stringsAsFactors}] logical: should character vectors be converted
to factors?  Note that this is overridden bu \code{as.is} and
\code{colClasses}, both of which allow finer control.

\item[\code{fileEncoding}] character string: if non-empty declares the
encoding used on a file (not a connection) so the character data can
be re-encoded.  See \code{\LinkA{file}{file}}.


\item[\code{encoding}] encoding to be assumed for input strings.  It is
used to mark character strings as known to be in
Latin-1 or UTF-8: it is not used to re-encode the input, but allows
\R{} to handle encoded strings in their native encoding (if one of
those two).  See `Value'.


\item[\code{...}] Further arguments to be passed to \code{read.table}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This function is the principal means of reading tabular data into \R{}.

A field or line is `blank' if it contains nothing (except
whitespace if no separator is specified) before a comment character or
the end of the field or line.

If \code{row.names} is not specified and the header line has one less
entry than the number of columns, the first column is taken to be the
row names.  This allows data frames to be read in from the format in
which they are printed.  If \code{row.names} is specified and does
not refer to the first column, that column is discarded from such files.

The number of data columns is determined by looking at the first five lines
of input (or the whole file if it has less than five lines), or from
the length of \code{col.names} if it is specified and
is longer.  This could conceivably be wrong if \code{fill} or
\code{blank.lines.skip} are true, so specify \code{col.names} if necessary.

\code{read.csv} and \code{read.csv2} are identical to
\code{read.table} except for the defaults.  They are intended for
reading `comma separated value' files (\file{.csv}) or
(\code{read.csv2}) the variant
used in countries that use a comma as decimal point and a semicolon
as field separator.  Similarly, \code{read.delim} and
\code{read.delim2} are for reading delimited files, defaulting to the
TAB character for the delimiter.  Notice that \code{header = TRUE} and
\code{fill = TRUE} in these variants, and that the comment character
is disabled.

The rest of the line after a comment character is skipped; quotes
are not processed in comments.  Complete comment lines are allowed
provided \code{blank.lines.skip = TRUE}; however, comment lines prior
to the header must have the comment character in the first non-blank
column.

Quoted fields with embedded newlines are supported except after a
comment character.

Note that unless \code{colClasses} is specified, all columns are read
as character columns and then converted.  This means that quotes are
interpreted in all fields and that a column of values like \code{"42"}
will result in an integer column.
\end{Details}
%
\begin{Value}
A data frame (\code{\LinkA{data.frame}{data.frame}}) containing a representation of
the data in the file.

Empty input is an error unless \code{col.names} is specified, when a
0-row data frame is returned: similarly giving just a header line if
\code{header = TRUE} results in a 0-row data frame.  Note that in
either case the columns will be logical unless \code{colClasses} was
supplied.

Character strings in the result (including factor levels) will have a
declared encoding if \code{encoding} is \code{"latin1"} or
\code{"UTF-8"}.
\end{Value}
%
\begin{Note}\relax
The columns referred to in \code{as.is} and \code{colClasses} include
the column of row names (if any).

Less memory will be used if \code{colClasses} is specified as one of
the six \LinkA{atomic}{atomic} vector classes.  This can be particularly so when
reading a column that takes many distinct numeric values, as storing
each distinct value as a character string can take up to 14 times as
much memory as storing it as an integer.

Using \code{nrows}, even as a mild over-estimate, will help memory
usage.

Using \code{comment.char = ""} will be appreciably faster than the
\code{read.table} default.

\code{read.table} is not the right tool for reading large matrices,
especially those with many columns: it is designed to read
\emph{data frames} which may have columns of very different classes.
Use \code{\LinkA{scan}{scan}} instead.
\end{Note}
%
\begin{References}\relax
Chambers, J. M. (1992)
\emph{Data for models.}
Chapter 3 of \emph{Statistical Models in S}
eds J. M. Chambers and T. J. Hastie, Wadsworth \& Brooks/Cole.
\end{References}
%
\begin{SeeAlso}\relax
The \emph{R Data Import/Export} manual.

\code{\LinkA{scan}{scan}}, \code{\LinkA{type.convert}{type.convert}},
\code{\LinkA{read.fwf}{read.fwf}} for reading \emph{f}ixed \emph{w}idth
\emph{f}ormatted input;
\code{\LinkA{write.table}{write.table}};
\code{\LinkA{data.frame}{data.frame}}.

\code{\LinkA{count.fields}{count.fields}} can be useful to determine problems with
reading files which result in reports of incorrect record lengths.

\url{http://tools.ietf.org/html/rfc4180} for the IANA definition of
CSV files (which requires comma as separator and CRLF line endings).
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{recover}{Browsing after an Error}{recover}
\aliasA{limitedLabels}{recover}{limitedLabels}
\keyword{programming}{recover}
\keyword{debugging}{recover}
%
\begin{Description}\relax
This function allows the user to browse directly on any of the
currently active function calls, and is suitable as an error option.
The expression \code{options(error=recover)} will make this
the error option.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
recover()
\end{verbatim}
\end{Usage}
%
\begin{Details}\relax
When called, \code{recover} prints the list of current calls, and
prompts the user to select one of them.  The standard \R{}
\code{\LinkA{browser}{browser}} is then invoked from the corresponding
environment; the user can type ordinary S language expressions to be
evaluated in that environment.

When finished browsing in this call, type \code{c} to return to
\code{recover} from the browser.  Type another frame number to browse
some more, or type \code{0} to exit \code{recover}.

The use of \code{recover} largely supersedes \code{\LinkA{dump.frames}{dump.frames}}
as an error option, unless you really want to wait to look at the
error.  If \code{recover} is called in non-interactive mode, it
behaves like \code{dump.frames}.  For computations involving large
amounts of data, \code{recover} has the advantage that it does not
need to copy out all the environments in order to browse in them.  If
you do decide to quit interactive debugging, call
\code{\LinkA{dump.frames}{dump.frames}} directly while browsing in any frame (see
the examples).
\end{Details}
%
\begin{Value}
Nothing useful is returned.  However, you \emph{can} invoke
\code{recover} directly from a function, rather than through the error
option shown in the examples.  In this case, execution continues
after you type \code{0} to exit \code{recover}.
\end{Value}
%
\begin{Section}{Compatibility Note}
The \R{} \code{recover} function can be used in the same way as the
S function of the same name; therefore, the error option shown is
a compatible way to specify the error action.  However, the actual
functions are essentially unrelated and interact quite differently
with the user.  The navigating commands \code{up} and \code{down} do
not exist in the \R{} version; instead, exit the browser and select
another frame.
\end{Section}
%
\begin{References}\relax
John M. Chambers (1998).
\emph{Programming with Data}; Springer. \\{}
See the compatibility note above, however.
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{browser}{browser}} for details about the interactive computations;
\code{\LinkA{options}{options}} for setting the error option;
\code{\LinkA{dump.frames}{dump.frames}} to save the current environments for later
debugging.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 

options(error = recover) # setting the error option

### Example of interaction

> myFit <- lm(y ~ x, data = xy, weights = w)
Error in lm.wfit(x, y, w, offset = offset, ...) :
        missing or negative weights not allowed

Enter a frame number, or 0 to exit
1:lm(y ~ x, data = xy, weights = w)
2:lm.wfit(x, y, w, offset = offset, ...)
Selection: 2
Called from: eval(expr, envir, enclos)
Browse[1]> objects() # all the objects in this frame
[1] "method" "n"      "ny"     "offset" "tol"    "w"
[7] "x"      "y"
Browse[1]> w
[1] -0.5013844  1.3112515  0.2939348 -0.8983705 -0.1538642
[6] -0.9772989  0.7888790 -0.1919154 -0.3026882
Browse[1]> dump.frames() # save for offline debugging
Browse[1]> c # exit the browser

Enter a frame number, or 0 to exit
1:lm(y ~ x, data = xy, weights = w)
2:lm.wfit(x, y, w, offset = offset, ...)
Selection: 0 # exit recover
>


## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{relist}{Allow Re-Listing an unlist()ed Object}{relist}
\aliasA{as.relistable}{relist}{as.relistable}
\aliasA{is.relistable}{relist}{is.relistable}
\methaliasA{relist.default}{relist}{relist.default}
\methaliasA{relist.factor}{relist}{relist.factor}
\methaliasA{relist.list}{relist}{relist.list}
\methaliasA{relist.matrix}{relist}{relist.matrix}
\aliasA{unlist.relistable}{relist}{unlist.relistable}
\keyword{list}{relist}
\keyword{manip}{relist}
%
\begin{Description}\relax
\code{relist()} is an S3 generic function with a few methods in order
to allow easy inversion of \code{\LinkA{unlist}{unlist}(obj)} when that is used
with an object \code{obj} of (S3) class \code{"relistable"}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
relist(flesh, skeleton)
## Default S3 method:
relist(flesh, skeleton = attr(flesh, "skeleton"))
## S3 method for class 'factor':
relist(flesh, skeleton = attr(flesh, "skeleton"))
## S3 method for class 'list':
relist(flesh, skeleton = attr(flesh, "skeleton"))
## S3 method for class 'matrix':
relist(flesh, skeleton = attr(flesh, "skeleton"))

as.relistable(x)
is.relistable(x)

## S3 method for class 'relistable':
unlist(x, recursive = TRUE, use.names = TRUE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{flesh}] a vector to be relisted
\item[\code{skeleton}] a list, the structure of which determines the structure
of the result
\item[\code{x}] an \R{} object, typically a list (or vector).
\item[\code{recursive}] logical.  Should unlisting be applied to list
components of \code{x}?
\item[\code{use.names}] logical.  Should names be preserved?
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Some functions need many parameters, which are most easily represented in
complex structures, e.g., nested lists.  Unfortunately, many
mathematical functions in \R{}, including \code{\LinkA{optim}{optim}} and
\code{\LinkA{nlm}{nlm}} can only operate on functions whose domain is
a vector.  \R{} has \code{\LinkA{unlist}{unlist}()} to convert nested list
objects into a vector representation.  \code{relist()}, it's methods and
the functionality mentioned here provide the inverse operation to convert
vectors back to the convenient structural representation.
This allows structured functions (such as \code{optim()}) to have simple
mathematical interfaces.

For example, a likelihood function for a multivariate normal model needs a
variance-covariance matrix and a mean vector.  It would be most convenient to
represent it as a list containing a vector and a matrix.  A typical parameter
might look like
\begin{alltt}
      list(mean=c(0, 1), vcov=cbind(c(1, 1), c(1, 0))).
  \end{alltt}

However, \code{\LinkA{optim}{optim}} cannot operate on functions that take
lists as input; it only likes numeric vectors.  The solution is
conversion. Given a function \code{mvdnorm(x, mean, vcov, log=FALSE)}
which computes the required probability density, then
\begin{alltt}
        ipar <- list(mean=c(0, 1), vcov=cbind(c(1, 1), c(1, 0)))
        initial.param <- as.relistable(ipar)

        ll <- function(param.vector)
        \{
           param <- relist(param.vector, skeleton=ipar))
           -sum(mvdnorm(x, mean = param$mean, vcov = param$vcov,
                        log = TRUE))
        \}

        optim(unlist(initial.param), ll)
  \end{alltt}

\code{relist} takes two parameters: skeleton and flesh.  Skeleton is a sample
object that has the right \code{shape} but the wrong content.  \code{flesh}
is a vector with the right content but the wrong shape.  Invoking
\begin{alltt}
    relist(flesh, skeleton)
  \end{alltt}

will put the content of flesh on the skeleton.  You don't need to specify
skeleton explicitly if the skeleton is stored as an attribute inside flesh.
In particular, if flesh was created from some object obj with
\code{unlist(as.relistable(obj))}
then the skeleton attribute is automatically set.  (Note that this
does not apply to the example here, as \code{\LinkA{optim}{optim}} is creating
a new vector to pass to \code{ll} and not its \code{par} argument.)

As long as \code{skeleton} has the right shape, it should be a precise inverse
of \code{\LinkA{unlist}{unlist}}.  These equalities hold:
\begin{alltt}
   relist(unlist(x), x) == x
   unlist(relist(y, skeleton)) == y

   x <- as.relistable(x)
   relist(unlist(x)) == x
  \end{alltt}

\end{Details}
%
\begin{Value}
an object of (S3) class \code{"relistable"} (and \code{"\LinkA{list}{list}"}).
\end{Value}
%
\begin{Author}\relax
R Core, based on a code proposal by Andrew Clausen.
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{unlist}{unlist}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
 ipar <- list(mean=c(0, 1), vcov=cbind(c(1, 1), c(1, 0)))
 initial.param <- as.relistable(ipar)
 ul <- unlist(initial.param)
 relist(ul)
 stopifnot(identical(relist(ul), initial.param))
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{REMOVE}{Remove Add-on Packages}{REMOVE}
\keyword{utilities}{REMOVE}
%
\begin{Description}\relax
Utility for removing add-on packages.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
R CMD REMOVE [options] [-l lib] pkgs
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{pkgs}] a space-separated list with the names of the bundles or
packages to be removed.
\item[\code{lib}] the path name of the \R{} library tree to remove from.  May
be absolute or relative.  Also accepted in the form \samp{--library=lib}.
\item[\code{options}] further options for help or version.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If the name of a bundle is given, the whole bundle will be removed.

If used as \command{R CMD REMOVE pkgs} without explicitly specifying
\code{lib}, packages are removed from the library tree rooted at the
first directory in the library path which would be used by \R{} run in
the current environment.

To remove from the library tree \code{\var{lib}} instead of the default
one, use \command{R CMD REMOVE -l lib \var{pkgs}}.

Use \command{R CMD REMOVE --help} for more usage information.  
\end{Details}
%
\begin{Note}\relax
Some binary distributions of \R{} have \code{REMOVE} in a separate
bundle, e.g. an \code{R-devel} RPM.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{INSTALL}{INSTALL}}, \code{\LinkA{remove.packages}{remove.packages}}
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{remove.packages}{Remove Installed Packages}{remove.packages}
\keyword{utilities}{remove.packages}
%
\begin{Description}\relax
Removes installed packages/bundles and updates index information
as necessary.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
remove.packages(pkgs, lib)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{pkgs}] a character vector with the names of the package(s) or
bundle(s) to be removed.
\item[\code{lib}] a character vector giving the library directories to remove the
packages from.  If missing, defaults to the first element in
\code{\LinkA{.libPaths}{.libPaths}()}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If an element of \code{pkgs} matches a bundle name, all the packages
in the bundle will be removed.  This takes precedence over matching a
package name.
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{REMOVE}{REMOVE}} for a command line version;
\code{\LinkA{install.packages}{install.packages}} for installing packages.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{RHOME}{R Home Directory}{RHOME}
\keyword{utilities}{RHOME}
%
\begin{Description}\relax
Returns the location of the \R{} home directory, which is the root of
the installed \R{} tree.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
R RHOME
\end{verbatim}
\end{Usage}
\inputencoding{latin1}
\HeaderA{roman}{Roman Numerals}{roman}
\aliasA{as.roman}{roman}{as.roman}
\keyword{arith}{roman}
%
\begin{Description}\relax
Manipulate integers as roman numerals.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
as.roman(x)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] a numeric vector, or a character vector of arabic or roman
numerals.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\code{as.roman} creates objects of class \code{"roman"} which are
internally represented as integers, and have suitable methods for
printing, formatting, subsetting, and coercion to \code{character}.

Only numbers between 1 and 3899 have a unique representation as roman
numbers.
\end{Details}
%
\begin{References}\relax
Wikipedia contributors (2006). Roman numerals.
Wikipedia, The Free Encyclopedia.
\url{http://en.wikipedia.org/w/index.php?title=Roman_numerals&oldid=78252134}.
Accessed September 29, 2006.
\end{References}
%
\begin{Examples}
\begin{ExampleCode}
## First five roman 'numbers'.
(y <- as.roman(1 : 5))
## Middle one.
y[3]
## Current year as a roman number.
(y <- as.roman(format(Sys.Date(), "%Y")))
## 10 years ago ...
y - 10
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{Rprof}{Enable Profiling of R's Execution}{Rprof}
\keyword{utilities}{Rprof}
%
\begin{Description}\relax
Enable or disable profiling of the execution of \R{} expressions.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
Rprof(filename = "Rprof.out", append = FALSE, interval = 0.02,
       memory.profiling=FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{filename}] 
The file to be used for recording the profiling results.
Set to \code{NULL} or \code{""} to disable profiling.

\item[\code{append}] 
logical: should the file be over-written or appended to?

\item[\code{interval}] 
real: time interval between samples.

\item[\code{memory.profiling}] logical: write memory use information to the file?
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Enabling profiling automatically disables any existing profiling to
another or the same file.

Profiling works by writing out the call stack every \code{interval}
seconds, to the file specified.  Either the \code{\LinkA{summaryRprof}{summaryRprof}}
function or the Perl script \command{R CMD Rprof} can be used to process
the output file to produce a summary of the
usage; use \command{R CMD Rprof --help} for usage information.

How time is measured varies by platform: on a Unix-alike it is the CPU
time of the \R{} process, so for example excludes time when \R{} is waiting
for input or for processes run by \code{\LinkA{system}{system}} to return.

Note that the timing interval cannot usefully be too small: once the
timer goes off, the information is not recorded until the next timing
click (probably in the range 1--10msecs).

Functions will only be recorded in the profile log if they put a
context on the call stack (see \code{\LinkA{sys.calls}{sys.calls}}).  Some
\LinkA{primitive}{primitive} functions do not do so: specifically those which are
of \LinkA{type}{type} \code{"special"} (see the `R Internals' manual
for more details).
\end{Details}
%
\begin{Note}\relax
Profiling is not available on all platforms.  By default, it is
attempted to compile support for profiling.  Configure \R{} with
\option{--disable-R-profiling} to change this.

As \R{} profiling uses the same mechanisms as C profiling, the two
cannot be used together, so do not use \code{Rprof} in an executable
built for profiling.
\end{Note}
%
\begin{SeeAlso}\relax
The chapter on ``Tidying and profiling R code'' in
``Writing \R{} Extensions'' (see the \file{doc/manual} subdirectory
of the \R{} source tree).

\code{\LinkA{summaryRprof}{summaryRprof}}

\code{\LinkA{tracemem}{tracemem}}, \code{\LinkA{Rprofmem}{Rprofmem}} for other ways to track
memory use.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: Rprof()
## some code to be profiled
Rprof(NULL)
## some code NOT to be profiled
Rprof(append=TRUE)
## some code to be profiled
Rprof(NULL)
...
## Now post-process the output as described in Details

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{Rprofmem}{Enable Profiling of R's Memory Use}{Rprofmem}
\keyword{utilities}{Rprofmem}
%
\begin{Description}\relax
Enable or disable reporting of memory allocation in R.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
Rprofmem(filename = "Rprofmem.out", append = FALSE, threshold = 0)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{filename}] The file to be used for recording the memory
allocations. Set to \code{NULL} or \code{""} to disable reporting. 
\item[\code{append}] logical: should the file be over-written or appended to? 
\item[\code{threshold}] numeric: allocations on R's "large vector" heap
larger than this number of bytes will be reported.

\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Enabling profiling automatically disables any existing profiling to
another or the same file.

Profiling writes the call stack to the specified file every time
\code{malloc} is called to allocate a large vector object or to
allocate a page of memory for small objects. The size of a page of
memory and the size above which \code{malloc} is used for vectors are
compile-time constants, by default 2000 and 128 bytes respectively.

The profiler tracks allocations, some of which will be to previously
used memory and will not increase the total memory use of R.
\end{Details}
%
\begin{Value}
None  
\end{Value}
%
\begin{Note}\relax
The memory profiler slows down R even when not in use, and so is a
compile-time option.
The memory profiler can be used at the same time as other \R{} and C profilers.

\end{Note}
%
\begin{SeeAlso}\relax
The R sampling profiler, \code{\LinkA{Rprof}{Rprof}} also collects
memory information.

\code{\LinkA{tracemem}{tracemem}} traces duplications of specific objects.

The "Writing R Extensions" manual section on "Tidying and profiling R code"
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
## not supported unless R is compiled to support it.
Rprofmem("Rprofmem.out", threshold=1000)
example(glm)
Rprofmem(NULL)
noquote(readLines("Rprofmem.out", n=5))

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{Rscript}{Scripting Front-End for R}{Rscript}
\keyword{utilities}{Rscript}
%
\begin{Description}\relax
This is an alternative front end for use in \samp{\#!} scripts and
other scripting applications.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
Rscript [options] [-e expression] file [args]
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{options}] A list of options beginning with \samp{--}.  These can
be any of the options of the standard \R{} front-end, and also those
described in the details.
\item[\code{expression}] a \R{} expression.
\item[\code{file}] The name of a file containing \R{} commands.  \samp{-}
indicates \file{stdin}.
\item[\code{args}] Arguments to be passed to the script in \code{file}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\command{Rscript --help} gives details of usage, and
\command{Rscript --version} gives the version of \command{Rscript}.

Other invocations invoke the \R{} front-end with selected options.  This
front-end is convenient for writing \samp{\#!} scripts since it is an
executable and takes \code{file} directly as an argument.  Options
\option{--slave --no-restore} are always supplied: these imply
\option{--no-save}.

\emph{Either} one or more \option{-e} options or \code{file} should
be supplied.  When using \option{-e} options be aware of the quoting
rules in the shell used: see the examples.

Additional options accepted (before \code{file} or \code{args}) are
\begin{description}

\item[\option{--verbose}] gives details of what \command{Rscript} is
doing.  Also passed on to \R{}.
\item[\option{--default-packages=list}] where \code{list} is a
comma-separated list of package names or \code{NULL}.  Sets the
environment variable \env{R\_DEFAULT\_PACKAGES} which determines the
packages loaded on startup.  The default for \command{Rscript}
omits \pkg{methods} as it takes about 60\% of the startup time.


\end{description}


Normally the version of \R{} is determined at installation, but this can
be overridden by setting the environment variable \env{RHOME}.

\code{\LinkA{stdin}{stdin}()} refers to the input file, and
\code{\LinkA{file}{file}("stdin")} to the \code{stdin} file stream of the
process.
\end{Details}
%
\begin{Note}\relax
\command{Rscript} is only supported on systems with the \code{execv}
system call.
\end{Note}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
Rscript -e 'date()' -e 'format(Sys.time(), "%a %b %d %X %Y")'

## example #! script for a Unix-alike

#! /path/to/Rscript --vanilla --default-packages=utils
args <- commandArgs(TRUE)
res <- try(install.packages(args))
if(inherits(res, "try-error")) q(status=1) else q()


## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{RShowDoc}{Show R Manuals and Other Documentation}{RShowDoc}
\keyword{documentation}{RShowDoc}
%
\begin{Description}\relax
Utility function to find and display \R{} documentation.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
RShowDoc(what, type = c("pdf", "html", "txt"), package)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{what}] a character string: see `Details'.
\item[\code{type}] an optional character string giving the preferred format.
\item[\code{package}] an optional character string specifying the name of a
package within which to look for documentation.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\code{what} can specify one of several different sources of documentation,
including the \R{} manuals (\code{R-admin}, \code{R-data}, \code{R-exts},
\code{R-intro}, \code{R-ints}, \code{R-lang}), \code{NEWS},
\code{COPYING} (the GPL licence), \code{FAQ} (also available as
\code{R-FAQ}), and the files in \file{\var{\LinkA{R\_HOME}{R.Rul.HOME}}/doc}.

If \code{package} is supplied, documentation is looked for in the
\file{doc} and top-level directories of an installed package of that name.

If \code{what} is missing a brief usage message is printed.

The documentation types are tried in turn starting with the first
specified in \code{type} (or \code{"pdf"} if none is specified).
\end{Details}
%
\begin{Value}
A invisible character string given the path to the file found.
\end{Value}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
RShowDoc("R-lang")
RShowDoc("FAQ", type="html")
RShowDoc("frame", package="grid")
RShowDoc("changes.txt", package="grid")
RShowDoc("NEWS", package="MASS")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{RSiteSearch}{Search for Key Words or Phrases in the R-help Mailing List Archives
 or Documentation}{RSiteSearch}
\keyword{utilities}{RSiteSearch}
\keyword{documentation}{RSiteSearch}
%
\begin{Description}\relax
Search for key words or phrases in the R-help mailing list
archives, help pages, vignettes or task views, using the search engine
at \url{http://search.r-project.org} and view them in a web browser.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
RSiteSearch(string,
            restrict = c("Rhelp08", "functions", "views"),
            format = c("normal", "short"),
            sortby = c("score", "date:late", "date:early",
                       "subject", "subject:descending",
                       "from", "from:descending",
                       "size", "size:descending"),
            matchesPerPage = 20)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{string}] word(s) or phrase to search.  If the words are to be
searched as one entity, enclose all words in braces (see example).
\item[\code{restrict}] a character vector, typically of length larger than one:
What areas to search in:
\code{Rhelp08} for R-help mailing list archive from January 2008
\code{Rhelp02} for mailing list archive 2002-2007
\code{Rhelp01} for mailing list archive before 2002
\code{R-devel} for R-devel mailing list
\code{R-sig-mix} for R-devel mailing list    
\code{functions} for help pages
\code{views} for task views
\code{vignettes} for package vignettes
Use \code{c()} to specify more than one.
\item[\code{format}] \code{normal} or \code{short} (no excerpts); can be
abbreviated.
\item[\code{sortby}] character string (can be abbreviated) indicating how to
sort the search results:\\{}
(\code{score},
\code{date:late} for sorting by date with latest results first,
\code{date:early} for earliest first,
\code{subject} for subject in alphabetical order,
\code{subject:descending} for reverse alphabetical order,
\code{from} or \code{from:descending} for sender (when applicable),
\code{size} or \code{size:descending} for size.)
\item[\code{matchesPerPage}] How many items to show per page.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This function is designed to work with the search site at
\url{http://search.r-project.org}, and depends on that site
continuing to be made available (thanks to Jonathan Baron and the
School of Arts and Sciences of the University of Pennsylvania).

Unique partial matches will work for all arguments.  Each new
browser window will stay open unless you close it.
\end{Details}
%
\begin{Value}
(Invisibly) the complete URL passed to the browser,
including the query string.
\end{Value}
%
\begin{Author}\relax
Andy Liaw and Jonathan Baron
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{help.search}{help.search}}, \code{\LinkA{help.start}{help.start}} for local searches.

\code{\LinkA{browseURL}{browseURL}} for how the help file is displayed.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
 # need Internet connection
RSiteSearch("{logistic regression}") # matches exact phrase
Sys.sleep(5) # allow browser to open, take a quick look
RSiteSearch("Baron Liaw", restrict = "Rhelp02")
## Search in R-devel archive and recent mail (and store the query-string):
Sys.sleep(5)
fullquery <- RSiteSearch("S4", restrict = c("R-dev", "Rhelp08"))
fullquery # a string of ~ 116 characters
## the latest purported bug reports, responses ...

Sys.sleep(5)
RSiteSearch("bug", restrict = "R-devel", sortby = "date:late")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{rtags}{An etags-like a tagging utility for R}{rtags}
\keyword{programming}{rtags}
\keyword{utilities}{rtags}
%
\begin{Description}\relax
\code{rtags} provides etags-like indexing capabilities for R code,
using R's own parser.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
rtags(path = ".", pattern = "\\.[RrSs]$",
      recursive = FALSE,
      src = list.files(path = path, pattern = pattern,
                       full.names = TRUE,
                       recursive = recursive),
      keep.re = NULL,
      ofile = "", append = FALSE,
      verbose = getOption("verbose"))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{path, pattern, recursive}] 
Arguments passed on to \code{\LinkA{list.files}{list.files}} to determine the
files to be tagged.  By default, these are all files with extension
\code{.R}, \code{.r}, \code{.S}, and \code{.s} in the current
directory.  These arguments are ignored if \code{src} is specified.

\item[\code{src}] 
A vector of file names to be indexed.

\item[\code{keep.re}]  A regular expression further restricting \code{src}
(the files to be indexed).  For example, specifying
\code{keep.re="/R/[\textasciicircum{}/]*\bsl{}\bsl{}.R\$"} will only retain files with
extension \code{.R} inside a directory named \code{R}.

\item[\code{ofile}]  Passed on to \code{\LinkA{cat}{cat}} as the \code{file}
argument; typically the output file where the tags will be written
(\code{"TAGS"} by convention).  By default, the output is written to
the R console (unless redirected).

\item[\code{append}]  Logical, indicating whether the output should overwrite
an existing file, or append to it.

\item[\code{verbose}]  Logical.  If \code{TRUE}, file names are echoed to the
R console as they are processed.

\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax

Many text editors allow definitions of functions and other language
objects to be quickly and easily located in source files through a
tagging utility.  This functionality requires the relevant source
files to be preprocessed, producing an index (or tag) file containing
the names and their corresponding locations.  There are multiple tag
file formats, the most popular being the vi-style ctags format and the
and emacs-style etags format.  Tag files in these formats are usually
generated by the \code{ctags} and \code{etags} utilities respectively.
Unfortunately, these programs do not recognize R code syntax.  They do
allow tagging of arbitrary language files through regular expressions,
but this too is insufficient.

The \code{rtags} function is intended to be a tagging utility for R
code.  It parses R code files (using R's parser) and produces tags in
Emacs' etags format.  Support for vi-style tags is currently absent,
but should not be difficult to add.
\end{Details}
%
\begin{Author}\relax
Deepayan Sarkar
\end{Author}
%
\begin{References}\relax
\url{http://en.wikipedia.org/wiki/Ctags},
\url{http://www.gnu.org/software/emacs/emacs-lisp-intro/html_node/emacs.html#Tags}
\end{References}
%
\begin{SeeAlso}\relax
 \code{\LinkA{list.files}{list.files}}, \code{\LinkA{cat}{cat}} 
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}

## Not run: 
rtags("/path/to/src/repository",
      pattern = "[.]*\\.[RrSs]$",
      keep.re = "/R/",
      verbose = TRUE,
      ofile = "TAGS",
      append = FALSE,
      recursive = TRUE)

## End(Not run)

\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{Rtangle}{R Driver for Stangle}{Rtangle}
\aliasA{RtangleSetup}{Rtangle}{RtangleSetup}
\keyword{utilities}{Rtangle}
%
\begin{Description}\relax
A driver for \code{\LinkA{Stangle}{Stangle}} that extracts R code chunks.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
Rtangle()
RtangleSetup(file, syntax, output = NULL, annotate = TRUE,
             split = FALSE, prefix = TRUE, quiet = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] Name of Sweave source file.
\item[\code{syntax}] An object of class \code{SweaveSyntax}.
\item[\code{output}] Name of output file, default is to remove extension \file{.nw},
\file{.Rnw} or \file{.Snw} and to add extension \file{.R}. Any
directory names in \code{file} are also removed such that the output
is created in the current working directory.
\item[\code{annotate}] By default, code chunks are separated by comment
lines specifying the names and numbers of the code chunks. If
\code{FALSE}, only the code chunks without any decorating comments
are extracted.
\item[\code{split}] Split output in single files per code chunk?
\item[\code{prefix}] If \code{split = TRUE}, prefix the chunk labels by the
basename of the input file to get output file names?
\item[\code{quiet}] If \code{TRUE} all progress messages are suppressed.
\end{ldescription}
\end{Arguments}
%
\begin{Author}\relax
Friedrich Leisch
\end{Author}
%
\begin{References}\relax
Friedrich Leisch: Sweave User Manual, 2008\\{}
\url{http://www.stat.uni-muenchen.de/~leisch/Sweave}
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{Sweave}{Sweave}}, \code{\LinkA{RweaveLatex}{RweaveLatex}}
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{RweaveLatex}{R/LaTeX Driver for Sweave}{RweaveLatex}
\aliasA{RweaveLatexSetup}{RweaveLatex}{RweaveLatexSetup}
\keyword{utilities}{RweaveLatex}
%
\begin{Description}\relax
A driver for \code{\LinkA{Sweave}{Sweave}} that translates R code chunks in
LaTeX files.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
RweaveLatex()

RweaveLatexSetup(file, syntax, output = NULL, quiet = FALSE,
                 debug = FALSE, stylepath, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] Name of Sweave source file.
\item[\code{syntax}] An object of class \code{SweaveSyntax}.
\item[\code{output}] Name of output file, default is to remove extension
\file{.nw}, \file{.Rnw} or \file{.Snw} and to add extension
\file{.tex}. Any
directory names in \code{file} are also removed such that the output
is created in the current working directory.
\item[\code{quiet}] If \code{TRUE} all progress messages are suppressed.
\item[\code{debug}] If \code{TRUE}, input and output of all code
chunks is copied to the console.
\item[\code{stylepath}] See `Details'.
\item[\code{...}] named values for the options listed in `Supported
Options'.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The LaTeX file generated needs to contain
\samp{\bsl{}usepackage\{Sweave\}}, and if this is not present in the
Sweave source file, it is inserted by the \code{RweaveLatex} driver.
If \code{stylepath = TRUE}, a hard-coded path to the file
\file{Sweave.sty} in the \R{} installation is set in place of \code{Sweave}.
The hard-coded path makes the TeX file less portable, but avoids the
problem of installing the current version of \file{Sweave.sty} to some
place in your TeX input path.  However, TeX may not be able to
process the hard-coded path if it contains spaces (as it often will
under Windows) or TeX special characters.

The default in \R{} prior to 2.7.0 was \code{stylepath = TRUE}.  It is
now taken from the environment variable
\env{SWEAVE\_STYLEPATH\_DEFAULT}, or is \code{FALSE} it that is unset or
empty.  If set, it should be exactly \code{TRUE} or \code{FALSE}: any
other values are taken as \code{FALSE}.

By default, \file{Sweave.sty} sets the width of all included graphics to:\\{}
\samp{\bsl{}setkeys\{Gin\}\{width=0.8\bsl{}textwidth\}}.

This setting affects the width size option passed to the
\samp{\bsl{}includegraphics\{\}} directive for each plot file and in turn
impacts the scaling of your plot files as they will appear in your
final document.

Thus, for example, you may set \code{width=3} in your figure chunk and
the runtime generated EPS and PDF files will be set to 3 inches in
width. However, the width of your graphic in your final document, will
be set to \samp{0.8\bsl{}textwidth} and the height dimension will be
scaled accordingly. Fonts and symbols will be similarly scaled in the
final document. 

You can adjust the default value by including the
\samp{\bsl{}setkeys\{Gin\}\{width=...\}} directive in your .Rnw file after the
\samp{\bsl{}begin\{document\}} directive and changing the \code{width}
option value as you prefer, using standard LaTeX measurement values.

If you wish to override this default behavior entirely, you can add a
\samp{\bsl{}usepackage[nogin]\{Sweave\}} directive in your preamble. In this
case, no size/scaling options will be passed to the
\samp{\bsl{}includegraphics\{\}} directive and the \code{height} and
\code{width} options will determine both the runtime generated graphic
file sizes and the size of the graphics in your final document.

\file{Sweave.sty} also supports the \samp{[noae]} option, which
suppresses the use of the \samp{ae} package, the use of which may
interfere with certain encoding and typeface selections. If you have
problems in the rendering of certain character sets, try this option.
\end{Details}
%
\begin{Section}{Supported Options}
\code{RweaveLatex} supports the following options for code chunks (the values
in parentheses show the default values):
\begin{description}

\item[echo:] logical (\code{TRUE}). Include S code in the
output file?
\item[keep.source:] logical (\code{FALSE}).  When echoing, if
\code{keep.source == TRUE} the original source is copied to the file.
Otherwise, deparsed source is echoed.
\item[eval:] logical (\code{TRUE}). If \code{FALSE}, the code chunk is not
evaluated, and hence no text or graphical output produced.
\item[results:] character string (\code{verbatim}).
If \code{verbatim}, the output of S commands is
included in the verbatim-like Soutput environment. If
\code{tex}, the output is taken to be already proper latex markup
and included as is. If \code{hide} then all output is
completely suppressed (but the code executed during the weave).
\item[print:] logical (\code{FALSE})
If \code{TRUE}, each expression in the
code chunk is wrapped into a \code{print()} statement before evaluation,
such that the values of all expressions become visible.
\item[term:] logical (\code{TRUE}). If \code{TRUE}, visibility of values
emulates an interactive R session: values of assignments are not
printed, values of single objects are printed. If \code{FALSE},
output comes only from explicit \code{\LinkA{print}{print}} or
\code{\LinkA{cat}{cat}} statements.
\item[split:] logical (\code{FALSE}). If \code{TRUE}, text output is
written to separate files
for each code chunk.
\item[strip.white:] character string (\code{false}). If \code{true}, blank
lines at the beginning and end of output are removed. If
\code{all}, then all blank lines are removed from the output.
\item[prefix:] logical (\code{TRUE}). If \code{TRUE} generated filenames of
figures and output have a common prefix.
\item[prefix.string:] a character string, default is the name of the
\file{.Snw} source file.
\item[include:] logical (\code{TRUE}), indicating whether input
statements for text output and includegraphics statements
for figures should be auto-generated. Use \code{include = FALSE} if
the output should appear in a different place than the code chunk
(by placing the input line manually).
\item[fig:] logical (\code{FALSE}), indicating whether the code
chunk produces
graphical output. Note that only one figure per code chunk can be
processed this way.
\item[eps:] logical (\code{TRUE}), indicating whether EPS figures should be
generated. Ignored if \code{fig = FALSE}.
\item[pdf:] logical (\code{TRUE}), indicating whether PDF figures should be
generated. Ignored if \code{fig = FALSE}.
\item[pdf.version, pdf.encoding:] passed to \code{\LinkA{pdf}{pdf}} to set
the version and encoding.  Defaults taken from \code{pdf.options()}.
\item[width:] numeric (6), width of figures in inches. See `Details'.
\item[height:] numeric (6), height of figures in inches. See `Details'.
\item[expand:] logical (\code{TRUE}).  Expand references to other chunks
so that only R code appears in the output file.  If \code{FALSE}, the
chunk reference (e.g. \code{<{}<chunkname>{}>}) will appear.  The
\code{expand=FALSE} option requires \code{keep.source = TRUE} or it
will have no effect.
\item[concordance:] logical (\code{FALSE}).  Write a concordance file
to link the input line numbers to the output line numbers.  This is
an experimental feature; see the source code for the output format,
which is subject to change in future releases.

\end{description}

\end{Section}
%
\begin{Author}\relax
Friedrich Leisch
\end{Author}
%
\begin{References}\relax
Friedrich Leisch: Sweave User Manual, 2008\\{}
\url{http://www.stat.uni-muenchen.de/~leisch/Sweave}
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{Sweave}{Sweave}}, \code{\LinkA{Rtangle}{Rtangle}}
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{savehistory}{Load or Save or Display the Commands History}{savehistory}
\aliasA{history}{savehistory}{history}
\aliasA{loadhistory}{savehistory}{loadhistory}
\aliasA{timestamp}{savehistory}{timestamp}
\keyword{utilities}{savehistory}
%
\begin{Description}\relax
Load or save or display the commands history.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
loadhistory(file = ".Rhistory")
savehistory(file = ".Rhistory")

history(max.show = 25, reverse = FALSE, pattern, ...)

timestamp(stamp = date(),
          prefix = "##------ ", suffix = " ------##",
          quiet = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] The name of the file in which to save the history, or
from which to load it. The path is relative to the current
working directory.
\item[\code{max.show}] The maximum number of lines to show. \code{Inf} will
give all of the currently available history.
\item[\code{reverse}] logical. If true, the lines are shown in reverse
order. Note: this is not useful when there are continuation lines.
\item[\code{pattern}] A character string to be matched against the lines of
the history
\item[\code{...}] Arguments to be passed to \code{\LinkA{grep}{grep}} when doing
the matching.
\item[\code{stamp}] A value or vector of values to be written into the history.
\item[\code{prefix}] A prefix to apply to each line.
\item[\code{suffix}] A suffix to apply to each line.
\item[\code{quiet}] If \code{TRUE}, suppress printing timestamp to the console.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
There are several history mechanisms available for the different \R{}
consoles, which work in similar but not identical ways.  Other uses of
\R{}, in particular embedded uses, may have no history.
This works under the \code{readline} command-line interface and the
\file{R.app} Mac OS X GUI, but not otherwise (for example, in
batch use or in an embedded application).

The \code{readline} history mechanism is controlled by two environment
variables: \env{R\_HISTSIZE} controls the number of lines that are
saved (default 512), and \env{R\_HISTFILE} sets the filename used for
the loading/saving of history if requested at the beginning/end of a
session (but not the default for these functions).  There is no limit
on the number of lines of history retained during a session, so
setting \env{R\_HISTSIZE} to a large value has no penalty unless a
large file is actually generated.

These variables are read at the time of saving, so can be altered
within a session by the use of \code{\LinkA{Sys.setenv}{Sys.setenv}}.

Note that \code{readline} history library save files with permission
\code{0600}, that is with read/write permission for the user and not
even read permission for any other account.

\code{history} shows only unique matching lines if \code{pattern} is
supplied.

The \code{timestamp} function writes a timestamp (or other message)
into the history and echos it to the console.  On platforms that do not
support a history mechanism (where the mechanism does not support
timestamps) only the console message is printed.
\end{Details}
%
\begin{Note}\relax
If you want to save the history at the end of (almost) every
interactive session (even those in which you do not save the
workspace), you can put a call to \code{savehistory()} in
\code{\LinkA{.Last}{.Last}}.  See the examples.
\end{Note}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
.Last <- function()
    if(interactive()) try(savehistory("~/.Rhistory"))

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{select.list}{Select Items from a List}{select.list}
\keyword{utilities}{select.list}
%
\begin{Description}\relax
Select item(s) from a character vector.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
select.list(list, preselect = NULL, multiple = FALSE, title = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{list}] a character vector of items.
\item[\code{preselect}] a character vector, or \code{NULL}.  If non-null and
if the string(s) appear in the list, the item(s) are selected
initially.
\item[\code{multiple}] logical: can more than one item be selected?
\item[\code{title}] optional character string for window title, or
\code{NULL} for no title.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Under the Mac OS X GUI this brings up a modal dialog box
with a (scrollable) list of items, which can be selected by the mouse.

Otherwise it displays a text list from which the user can choose by
number(s). The \code{multiple = FALSE} case uses \code{\LinkA{menu}{menu}}.
Preselection is only supported for \code{multiple = TRUE}, where it is
indicated by a \code{"+"} preceding the item.

It is an error to use \code{select.list} in a non-interactive session. 
\end{Details}
%
\begin{Value}
A character vector of selected items.  If \code{multiple} is false and
no item was selected (or \code{Cancel} was used), \code{""} is
returned.   If \code{multiple} is true and no item was selected (or
\code{Cancel} was used) then a character vector of length 0 is returned.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{menu}{menu}}, \code{\LinkA{tk\_select.list}{tk.Rul.select.list}} for a graphical
version using Tcl/Tk.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
select.list(sort(.packages(all.available = TRUE)))

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{sessionInfo}{Collect Information About the Current R Session}{sessionInfo}
\aliasA{print.sessionInfo}{sessionInfo}{print.sessionInfo}
\aliasA{toLatex.sessionInfo}{sessionInfo}{toLatex.sessionInfo}
\keyword{misc}{sessionInfo}
%
\begin{Description}\relax
Print version information about \R{} and attached or loaded packages.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
sessionInfo(package=NULL)
## S3 method for class 'sessionInfo':
print(x, locale=TRUE, ...)
## S3 method for class 'sessionInfo':
toLatex(object, locale=TRUE, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{package}] a character vector naming installed packages.  By
default all attached packages are used.
\item[\code{x}] an object of class \code{"sessionInfo"}.
\item[\code{object}] an object of class \code{"sessionInfo"}.
\item[\code{locale}] show locale information?
\item[\code{...}] currently not used.
\end{ldescription}
\end{Arguments}
%
\begin{SeeAlso}\relax
\code{\LinkA{R.version}{R.version}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
sessionInfo()
toLatex(sessionInfo(), locale=FALSE)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{setRepositories}{Select Package Repositories}{setRepositories}
\keyword{utilities}{setRepositories}
%
\begin{Description}\relax
Interact with the user to choose the package repositories to be used.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
setRepositories(graphics = getOption("menu.graphics"), ind = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{graphics}] Logical.
If true and \pkg{tcltk} and an X server are available, use a Tk
widget, or if under the AQUA interface use a Mac OS X widget,
otherwise use a text list in the console.
\item[\code{ind}] \code{NULL} or a vector of integer indices, which have the
same effect as if they were entered at the prompt for
\code{graphics=FALSE}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The default list of known repositories is stored in the file
\file{\var{\LinkA{R\_HOME}{R.Rul.HOME}}/etc/repositories}.
That file can be edited for a site, or a user can have a personal copy
in \file{\var{HOME}/.R/repositories} which will take precedence.

A Bioconductor mirror can be selected by setting
\code{options("BioC\_mirror")}: the default value is
\samp{"http://www.bioconductor.org"}.

The items that are preselected are those that are currently in
\code{options("repos")} plus those marked as default in the
list of known repositories.

The list of repositories offered depends on the setting of option
\code{"pkgType"} as some repositories only offer a subset of types
(e.g. only source packages or not Mac OS X binary packages).

This function requires the \R{} session to be interactive unless
\code{ind} is supplied.
\end{Details}
%
\begin{Value}
This function is invoked mainly for its side effect of updating
\code{options("repos")}.  It returns (invisibly) the previous
\code{repos} options setting (as a \code{\LinkA{list}{list}} with component
\code{repos}) or \code{\LinkA{NULL}{NULL}} if no changes were applied.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{chooseCRANmirror}{chooseCRANmirror}}, \code{\LinkA{install.packages}{install.packages}}.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{SHLIB}{Build Shared Object/DLL for Dynamic Loading}{SHLIB}
\keyword{utilities}{SHLIB}
%
\begin{Description}\relax
Compile the given source files and then link all specified object
files into a shared library aka DLL which can be loaded into \R{} using
\code{dyn.load} or \code{library.dynam}.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
R CMD SHLIB [options] [-o dllname] files
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{files}] a list specifying the object files to be included in the
shared object/DLL.  You can also include the name of source files (for
which the object files are automagically made from their sources)
and library linking commands.

\item[\code{dllname}] the full name of the shared object/DLL to be built,
including the extension (typically \file{.so} on Unix systems, and
\file{.dll} on Windows).  If not given, the basename of the object
is taken from the basename of the first file.
\item[\code{options}] Further options to control the processing.  Use
\command{R CMD SHLIB --help} for a current list.

\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\command{R CMD SHLIB} is the mechanism used by \code{\LinkA{INSTALL}{INSTALL}} to
compile source code in packages.  It will generate suitable
compilation commands for C, C++, ObjC(++) and Fortran sources: Fortran
90/95 sources can also be used but it may not be possible to mix these
with other languages (on most platforms it is possible to mix with C,
but mixing with C++ rarely works).

Please consult section `Creating shared objects' in the manual
`Writing R Extensions' for how to customize it (for example to
add \code{cpp} flags and to add libraries to the link step) and for
details of some of its quirks.

Items in \code{files} with extensions \file{.c}, \file{.cpp},
\file{.cc}, \file{.C}, \file{.f}, \file{.f90}, \file{.f95}, \file{.m}
(ObjC), \file{.M} and \file{.mm} (ObjC++) are regarded as source
files, and those with extension \file{.o} as object files.  All other
items are passed to the linker.

Option \option{-n} (also known as \option{--dry-run}) will show the
commands that would be run without actually executing them.
\end{Details}
%
\begin{Note}\relax
Some binary distributions of \R{} have \code{SHLIB} in a separate
bundle, e.g., an \code{R-devel} RPM.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{COMPILE}{COMPILE}},
\code{\LinkA{dyn.load}{dyn.load}},
\code{\LinkA{library.dynam}{library.dynam}}.

The section on ``Customizing compilation'' in the ``R
Administration and Installation'' manual (see the \file{doc/manual}
subdirectory of the \R{} source tree).

The `R Installation and Administration' and `Writing R 
Extensions' manuals.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
R CMD SHLIB -o mylib.so a.f b.f -L/opt/acml3.5.0/gnu64/lib -lacml

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{stack}{Stack or Unstack Vectors from a Data Frame or List}{stack}
\methaliasA{stack.data.frame}{stack}{stack.data.frame}
\methaliasA{stack.default}{stack}{stack.default}
\aliasA{unstack}{stack}{unstack}
\methaliasA{unstack.data.frame}{stack}{unstack.data.frame}
\methaliasA{unstack.default}{stack}{unstack.default}
\keyword{manip}{stack}
%
\begin{Description}\relax
Stacking vectors concatenates multiple vectors into a single vector
along with a factor indicating where each observation originated.
Unstacking reverses this operation.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
stack(x, ...)
## Default S3 method:
stack(x, ...)
## S3 method for class 'data.frame':
stack(x, select, ...)

unstack(x, ...)
## Default S3 method:
unstack(x, form, ...)
## S3 method for class 'data.frame':
unstack(x, form, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] object to be stacked or unstacked
\item[\code{select}] expression, indicating variables to select from a
data frame
\item[\code{form}] a two-sided formula whose left side evaluates to the
vector to be unstacked and whose right side evaluates to the
indicator of the groups to create.  Defaults to \code{formula(x)}
in \code{unstack.data.frame}.
\item[\code{...}] further arguments passed to or from other methods.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
The \code{stack} function is used to transform data available as
separate columns in a data frame or list into a single column that can
be used in an analysis of variance model or other linear model.  The
\code{unstack} function reverses this operation. 
\end{Details}
%
\begin{Value}
\code{unstack} produces a list of columns according to the formula
\code{form}.  If all the columns have the same length, the resulting
list is coerced to a data frame.

\code{stack} produces a data frame with two columns
\begin{ldescription}
\item[\code{values}] the result of concatenating the selected vectors in
\code{x}
\item[\code{ind}] a factor indicating from which vector in \code{x} the
observation originated
\end{ldescription}
\end{Value}
%
\begin{Author}\relax
Douglas Bates
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{lm}{lm}}, \code{\LinkA{reshape}{reshape}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(stats)
formula(PlantGrowth)         # check the default formula
pg <- unstack(PlantGrowth)   # unstack according to this formula
pg
stack(pg)                    # now put it back together
stack(pg, select = -ctrl)    # omitting one vector
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{str}{Compactly Display the Structure of an Arbitrary R Object}{str}
\methaliasA{str.data.frame}{str}{str.data.frame}
\methaliasA{str.default}{str}{str.default}
\aliasA{strOptions}{str}{strOptions}
\keyword{print}{str}
\keyword{documentation}{str}
\keyword{utilities}{str}
%
\begin{Description}\relax
Compactly display the internal \bold{str}ucture of an \R{} object, a
diagnostic function and an alternative to \code{\LinkA{summary}{summary}}
(and to some extent, \code{\LinkA{dput}{dput}}).  Ideally, only one line for
each `basic' structure is displayed.  It is especially well suited
to compactly display the (abbreviated) contents of (possibly nested)
lists.  The idea is to give reasonable output for \bold{any} \R{}
object.  It calls \code{\LinkA{args}{args}} for (non-primitive) function objects.

\code{strOptions()} is a convenience function for setting
\code{\LinkA{options}{options}(str = .)}, see the examples.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
str(object, ...)

## S3 method for class 'data.frame':
str(object, ...)

## Default S3 method:
str(object, max.level = NA,
    vec.len  = strO$vec.len, digits.d = strO$digits.d,
    nchar.max = 128, give.attr = TRUE,
    give.head = TRUE, give.length = give.head,
    width = getOption("width"), nest.lev = 0,
    indent.str = paste(rep.int(" ", max(0, nest.lev + 1)),
                       collapse = ".."),
    comp.str="$ ", no.list = FALSE, envir = baseenv(),
    strict.width = strO$strict.width,
    formatNum = strO$formatNum, ...)

strOptions(strict.width = "no", digits.d = 3, vec.len = 4,
           formatNum = function(x, ...)
                       format(x, trim=TRUE, drop0trailing=TRUE, ...))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{object}] any \R{} object about which you want to have some
information.
\item[\code{max.level}] maximal level of nesting which is applied for
displaying nested structures, e.g., a list containing sub lists.
Default NA: Display all nesting levels.
\item[\code{vec.len}] numeric (>= 0) indicating how many `first few' elements
are displayed of each vector.  The number is multiplied by different
factors (from .5 to 3) depending on the kind of vector.  Defaults to
the \code{vec.len} component of option \code{"str"} (see
\code{\LinkA{options}{options}}) which defaults to 4.
\item[\code{digits.d}] number of digits for numerical components (as for
\code{\LinkA{print}{print}}).  Defaults to the \code{digits.d} component of
option \code{"str"} which defaults to 3.
\item[\code{nchar.max}] maximal number of characters to show for
\code{\LinkA{character}{character}} strings.  Longer strings are truncated, see
\code{longch} example below.
\item[\code{give.attr}] logical; if \code{TRUE} (default), show attributes
as sub structures.
\item[\code{give.length}] logical; if \code{TRUE} (default), indicate
length (as \code{[1:...]}).
\item[\code{give.head}] logical; if \code{TRUE} (default), give (possibly
abbreviated) mode/class and length (as \code{<type>[1:...]}).
\item[\code{width}] the page width to be used.  The default is the currently
active \code{\LinkA{options}{options}("width")}; note that this has only a
weak effect, unless \code{strict.width} is not \code{"no"}.
\item[\code{nest.lev}] current nesting level in the recursive calls to
\code{str}.
\item[\code{indent.str}] the indentation string to use.
\item[\code{comp.str}] string to be used for separating list components.
\item[\code{no.list}] logical; if true, no `list of \dots' nor the
class are printed.
\item[\code{envir}] the environment to be used for \emph{promise} (see
\code{\LinkA{delayedAssign}{delayedAssign}}) objects only.
\item[\code{strict.width}] string indicating if the \code{width} argument's
specification should be followed strictly, one of the values
\code{c("no", "cut", "wrap")}.
Defaults to the \code{strict.width} component of option \code{"str"}
(see \code{\LinkA{options}{options}}) which defaults to \code{"no"} for back
compatibility reasons; \code{"wrap"} uses
\code{\LinkA{strwrap}{strwrap}(*, width=width)} whereas \code{"cut"} cuts
directly to \code{width}.  Note that a small \code{vec.length}
may be better than setting \code{strict.width = "wrap"}.
\item[\code{formatNum}] a function such as \code{\LinkA{format}{format}} for
formatting numeric vectors.  It defaults to the \code{formatNum}
component of option \code{"str"}, see ``Usage'' of
\code{strOptions()} above, which is almost back compatible to \R{}
version \eqn{\le}{} 2.7.x, however, using \code{\LinkA{formatC}{formatC}}
may be slightly better.
\item[\code{...}] potential further arguments (required for Method/Generic
reasons).
\end{ldescription}
\end{Arguments}
%
\begin{Value}
\code{str} does not return anything, for efficiency reasons.
The obvious side effect is output to the terminal.
\end{Value}
%
\begin{Author}\relax
Martin Maechler \email{maechler@stat.math.ethz.ch} since 1990.
\end{Author}
%
\begin{SeeAlso}\relax
\code{\LinkA{ls.str}{ls.str}} for \emph{listing} objects with their structure;
\code{\LinkA{summary}{summary}}, \code{\LinkA{args}{args}}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
require(stats); require(grDevices); require(graphics)
## The following examples show some of 'str' capabilities
str(1:12)
str(ls)
str(args) #- more useful than  args(args) !
str(freeny)
str(str)
str(.Machine, digits.d = 20)
str( lsfit(1:9,1:9))
str( lsfit(1:9,1:9), max.level = 1)
str( lsfit(1:9,1:9), width = 60, strict.width = "cut")
str( lsfit(1:9,1:9), width = 60, strict.width = "wrap")
op <- options(); str(op) # save first;
                         # otherwise internal options() is used.
need.dev <-
  !exists(".Device") || is.null(.Device) || .Device == "null device"
{ if(need.dev) postscript()
  str(par())
  if(need.dev) graphics.off()
}
ch <- letters[1:12]; is.na(ch) <- 3:5
str(ch) # character NA's

nchar(longch <- paste(rep(letters,100), collapse=""))
str(longch)
str(longch, nchar.max = 52)

str(longch, strict.width = "wrap")

## Settings for narrow transcript :
op <- options(width = 60,
              str = strOptions(strict.width = "wrap"))
str(lsfit(1:9,1:9))
str(options())
## reset to previous:
options(op)



str(quote( { A+B; list(C,D) } ))



## S4 classes :
require(stats4)
x <- 0:10; y <- c(26, 17, 13, 12, 20, 5, 9, 8, 5, 4, 8)
ll <- function(ymax=15, xh=6)
      -sum(dpois(y, lambda=ymax/(1+x/xh), log=TRUE))
fit <- mle(ll)
str(fit)

\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{summaryRprof}{Summarise Output of R Sampling Profiler}{summaryRprof}
\keyword{utilities}{summaryRprof}
%
\begin{Description}\relax
Summarise the output of the \code{\LinkA{Rprof}{Rprof}} function to show the
amount of time used by different \R{} functions.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
summaryRprof(filename = "Rprof.out", chunksize = 5000,
              memory=c("none","both","tseries","stats"),
              index=2, diff=TRUE, exclude=NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{filename}] Name of a file produced by \code{Rprof()}
\item[\code{chunksize}] Number of lines to read at a time
\item[\code{memory}] Summaries for memory information.  See `Details' below
\item[\code{index}] How to summarize the stack trace for memory
information.  See `Details' below.
\item[\code{diff}] If \code{TRUE} memory summaries use change in memory
rather than current memory
\item[\code{exclude}] Functions to exclude when summarizing the stack trace
for memory summaries

\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This function is an alternative to \command{R CMD Rprof}. It provides
the convenience of an all-\R{} implementation but will be slower for large
files.

As the profiling output file could be larger than available memory, it
is read in blocks of \code{chunksize} lines. Increasing \code{chunksize}
will make the function run faster if sufficient memory is available.

When called with \code{memory.profiling=TRUE}, the profiler writes
information on three aspects of memory use: vector memory in small
blocks on the R heap, vector memory in large blocks (from
\code{malloc}), memory in nodes on the R heap.  It also records the number of
calls to the internal function \code{duplicate} in the time
interval. \code{duplicate} is called by C code when arguments need to be
copied. Note that the profiler does not track which function actually
allocated the memory.

With \code{memory="both"} the change in total memory (truncated at zero)
is reported in addition to timing data. 

With \code{memory="tseries"} or \code{memory="stats"} the \code{index}
argument specifies how to summarize the stack trace. A positive number
specifies that many calls from the bottom of the stack; a negative
number specifies the number of calls from the top of the stack.  With
\code{memory="tseries"} the index is used to construct labels and may be
a vector to give multiple sets of labels. With \code{memory="stats"} the
index must be a single number and specifies how to aggregate the data to
the maximum and average of the memory statistics.  With both
\code{memory="tseries"} and \code{memory="stats"} the argument
\code{diff=TRUE} asks for summaries of the increase in memory use over
the sampling interval and \code{diff=FALSE} asks for the memory use at
the end of the interval.

\end{Details}
%
\begin{Value}
If \code{memory="none"},a  list with components
\begin{ldescription}
\item[\code{by.self}] Timings sorted by `self' time
\item[\code{by.total}] Timings sorted by `total' time
\item[\code{sampling.time}] Total length of profiling run

\end{ldescription}
If \code{memory="both"} the same list but with memory consumption in Mb
in addition to the timings.  

If \code{memory="tseries"} a data frame giving memory statistics over
time

If \code{memory="stats"} a \code{by} object giving memory statistics by function.
\end{Value}
%
\begin{SeeAlso}\relax
The chapter on ``Tidying and profiling R code'' in
``Writing \R{} Extensions'' (see the \file{doc/manual} subdirectory
of the \R{} source tree).

\code{\LinkA{Rprof}{Rprof}}

\code{\LinkA{tracemem}{tracemem}} traces copying of an object via the C function
\code{duplicate}.

\code{\LinkA{Rprofmem}{Rprofmem}} is a non-sampling memory use profiler.

\url{http://developer.r-project.org/memory-profiling.html}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
## Rprof() is not available on all platforms
Rprof(tmp <- tempfile())
example(glm)
Rprof()
summaryRprof(tmp)
unlink(tmp)

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{Sweave}{Automatic Generation of Reports}{Sweave}
\aliasA{Stangle}{Sweave}{Stangle}
\aliasA{SweaveSyntaxLatex}{Sweave}{SweaveSyntaxLatex}
\aliasA{SweaveSyntaxNoweb}{Sweave}{SweaveSyntaxNoweb}
\keyword{utilities}{Sweave}
%
\begin{Description}\relax
\code{Sweave} provides a flexible framework for mixing text and S code
for automatic report generation.  The basic idea is to replace the S
code with its output, such that the final document only contains the
text and the output of the statistical analysis. 
\end{Description}
%
\begin{Usage}
\begin{verbatim}
Sweave(file, driver = RweaveLatex(),
       syntax = getOption("SweaveSyntax"), ...)

Stangle(file, driver = Rtangle(),
        syntax = getOption("SweaveSyntax"), ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] Name of Sweave source file.
\item[\code{driver}] The actual workhorse, see details below.
\item[\code{syntax}] An object of class \code{SweaveSyntax} or a character
string with its name. The default installation provides
\code{SweaveSyntaxNoweb} and \code{SweaveSyntaxLatex}.
\item[\code{...}] Further arguments passed to the driver's setup function.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Automatic generation of reports by mixing word processing markup (like
latex) and S code.  The S code gets replaced by its output (text or
graphs) in the final markup file.  This allows a report to be re-generated
if the input data change and documents the code to reproduce the
analysis in the same file that also produces the report.

\code{Sweave} combines the documentation and code chunks together
(or their output) into a single document.  \code{Stangle} extracts only
the code from the Sweave file creating a valid S source file (that can
be run using \code{\LinkA{source}{source}}).  Code inside \code{\bsl{}Sexpr\{\}}
statements is ignored by \code{Stangle}.

\code{Stangle} is just a frontend to \code{Sweave} using a simple
driver by default, which discards the documentation and concatenates
all code chunks the current S engine understands.
\end{Details}
%
\begin{Section}{Hook Functions}
Before each code chunk is evaluated, a number of hook functions can be
executed.  If \code{getOption("SweaveHooks")} is set, it is taken to be
a collection of hook functions. For each logical option of a code chunk
(\code{echo}, \code{print}, \ldots) a hook can be specified, which is
executed if and only if the respective option is \code{TRUE}. Hooks must
be named elements of the list returned by
\code{getOption("SweaveHooks")} and be functions taking no
arguments.  E.g., if option \code{"SweaveHooks"} is defined as
\code{list(fig = foo)}, and \code{foo} is a function, then it would be
executed before the code in each figure chunk. This is especially useful
to set defaults for the graphical parameters in a series of figure
chunks.

Note that the user is free to define new Sweave options and associate
arbitrary hooks with them.  E.g., one could define a hook function for
option \code{clean} that removes all objects in the global
environment.  Then all code chunks with \code{clean = TRUE} would start
operating on an empty workspace.
\end{Section}
%
\begin{Section}{Syntax Definition}
Sweave allows a very flexible syntax framework for marking
documentation and text chunks. The default is a noweb-style syntax, as
alternative a latex-style syntax can be used. See the user manual for
details.
\end{Section}
%
\begin{Author}\relax
Friedrich Leisch
\end{Author}
%
\begin{References}\relax
Friedrich Leisch: Dynamic generation of statistical reports using
literate data analysis. In W. Hrdle and
B. Rnz, editors, Compstat 2002 - Proceedings in Computational
Statistics, pages 575--580. Physika Verlag, Heidelberg, Germany,
2002. ISBN 3-7908-1517-9.

Friedrich Leisch: Sweave User Manual, 2008\\{}
\url{http://www.stat.uni-muenchen.de/~leisch/Sweave}
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{RweaveLatex}{RweaveLatex}}, \code{\LinkA{Rtangle}{Rtangle}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
testfile <- system.file("Sweave", "Sweave-test-1.Rnw", package = "utils")

## enforce par(ask=FALSE)
options(device.ask.default=FALSE)

## create a LaTeX file
Sweave(testfile)

## This can be compiled to PDF by
## Not run: tools::texi2dvi("Sweave-test-1.tex", pdf=TRUE)
## or outside R by
## R CMD texi2dvi Sweave-test-1.tex
## which sets the appropriate TEXINPUTS path.

## create an S source file from the code chunks
Stangle(testfile)
## which can be sourced, e.g.
source("Sweave-test-1.R")


\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{SweaveSyntConv}{Convert Sweave Syntax}{SweaveSyntConv}
\keyword{utilities}{SweaveSyntConv}
%
\begin{Description}\relax
This function converts the syntax of files in \code{\LinkA{Sweave}{Sweave}}
format to another Sweave syntax definition. 
\end{Description}
%
\begin{Usage}
\begin{verbatim}
SweaveSyntConv(file, syntax, output = NULL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] Name of Sweave source file.
\item[\code{syntax}] An object of class \code{SweaveSyntax} or a character
string with its name giving the target syntax to which the file is
converted.
\item[\code{output}] Name of output file, default is to remove the extension
from the input file and to add the default extension of the target
syntax. Any directory names in \code{file} are also removed such
that the output is created in the current working directory.
\end{ldescription}
\end{Arguments}
%
\begin{Author}\relax
Friedrich Leisch
\end{Author}
%
\begin{References}\relax
Friedrich Leisch: Sweave User Manual, 2008\\{}
\url{http://www.stat.uni-muenchen.de/~leisch/Sweave}
\end{References}
%
\begin{SeeAlso}\relax
\code{\LinkA{RweaveLatex}{RweaveLatex}}, \code{\LinkA{Rtangle}{Rtangle}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
testfile <- system.file("Sweave", "Sweave-test-1.Rnw", package = "utils")

## convert the file to latex syntax
SweaveSyntConv(testfile, SweaveSyntaxLatex)

## and run it through Sweave
Sweave("Sweave-test-1.Stex")


\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{tar}{Create a Tar Archive}{tar}
\keyword{file}{tar}
\keyword{utilities}{tar}
%
\begin{Description}\relax
Create a tar archive.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
tar(tarfile, files = NULL,
    compression = c("none", "gzip", "bzip2", "xz"),
    compression_level = 6, tar = Sys.getenv("tar"))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{tarfile}] The pathname of the tar file: tilde expansion (see
\code{\LinkA{path.expand}{path.expand}}) will be performed.  Alternatively, a
connection that can be used for binary writes.

\item[\code{files}] A character vector of filepaths to be archived:
the default is to archive all files under the current directory.

\item[\code{compression}] logical or character.  The type of compression to
be used.  Can be abbreviated.

\item[\code{compression\_level}] integer: the level of compression.  Only used
for the internal method.

\item[\code{tar}] character string: the path to the command to be used.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This is either a wrapper for a \command{tar} command or uses an
internal implementation in \R{}.  The latter is used if \code{tarfile}
is a connection or if the argument \code{tar} is \code{"internal"} or
\code{""}.

Beware of portability considerations: the `tar' format no
longer has an agreed standard (`Unix Standard Tar' was part of
POSIX 1003.1:1998 but has been removed in favour of \command{pax}),
and in any case many common implementations diverged from the former
standard.  Known problems arise from
\begin{itemize}

\item The handling of file names of more than 100 bytes.  These were
unsupported in early versions of \command{tar}, and supported in one
way by POSIX \command{tar} and in another by GNU \command{tar}.  The
internal implementation uses the POSIX way which supports up to 255
bytes (depending on the path), and warns on paths of more than 100
bytes.

\item (File) links.  \command{tar} was developed on an OS that used
hard links, and physical files that were referred to more than one
in the list of files to be included were included only once, the
remaining instance being added as links.  Later a means to include
symbolic links was added.  The internal implementation supports
symbolic links (on OSes that support them), only.  Of course, the
question arises as to how links should be unpacked on OSes that do
not support them: for files at least file copies can be used.

\item Header fields, in particular the padding to be used when
fields are not full or not used.  POSIX did define the correct
behaviour but commonly used implementations did (and still do)
not comply.

\end{itemize}

For portability, avoid file paths of more than 100 bytes, and links
(or at least, hard links and symbolic links to directories).

The internal implementation writes only the blocks of 512 bytes
required, unlike GNU \command{tar} which by default pads with
\samp{nul} to a multiple of 20 blocks (10KB).  Implementations differ
to whether the block padding should occur before or after compression
(or both).

The internal implementation currently skips empty directories.
\end{Details}
%
\begin{Value}
The return code from \code{\LinkA{system}{system}}, invisibly.
\end{Value}
%
\begin{SeeAlso}\relax
\url{http://en.wikipedia.org/wiki/Tar_(file_format)},
\url{http://www.opengroup.org/onlinepubs/009695399/utilities/pax.html#tag_04_100_13_06}
for the way the POSIX utility \command{pax} handles \command{tar} formats.

\code{\LinkA{untar}{untar}}.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{toLatex}{Converting R Objects to BibTeX or LaTeX}{toLatex}
\aliasA{print.Bibtex}{toLatex}{print.Bibtex}
\aliasA{print.Latex}{toLatex}{print.Latex}
\aliasA{toBibtex}{toLatex}{toBibtex}
\keyword{misc}{toLatex}
%
\begin{Description}\relax
These methods convert R objects to character vectors with
BibTeX or LaTeX markup.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
toBibtex(object, ...)
toLatex(object, ...)
## S3 method for class 'Bibtex':
print(x, prefix="", ...)
## S3 method for class 'Latex':
print(x, prefix="", ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{object}] object of a class for which a \code{toBibtex} or
\code{toLatex} method exists.
\item[\code{x}] object of class \code{"Bibtex"} or
\code{"Latex"}.
\item[\code{prefix}] a character string which is printed at the beginning of
each line, mostly used to insert whitespace for indentation.
\item[\code{...}] currently not used in the print methods.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Objects of class \code{"Bibtex"} or \code{"Latex"} are simply
character vectors where each element holds one line of the
corresponding BibTeX or LaTeX file.
\end{Details}
%
\begin{SeeAlso}\relax
\code{\LinkA{citEntry}{citEntry}} and \code{\LinkA{sessionInfo}{sessionInfo}} for examples
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{txtProgressBar}{Text Progress Bar}{txtProgressBar}
\aliasA{close.txtProgressBar}{txtProgressBar}{close.txtProgressBar}
\aliasA{getTxtProgressBar}{txtProgressBar}{getTxtProgressBar}
\aliasA{setTxtProgressBar}{txtProgressBar}{setTxtProgressBar}
\keyword{utilities}{txtProgressBar}
%
\begin{Description}\relax
Text progress bar in the \R{} console.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
txtProgressBar(min = 0, max = 1, initial = 0, char = "=",
               width = NA, title, label, style = 1)

getTxtProgressBar(pb)
setTxtProgressBar(pb, value, title = NULL, label = NULL)
## S3 method for class 'txtProgressBar':
close(con, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{min, max}] (finite) numeric values for the extremes of the
progress bar.
\item[\code{initial, value}] initial or new value for the progress bar.
\item[\code{char}] the character (or character string) to form the progress bar.
\item[\code{width}] the width of the progress bar, as a multiple of the width
of \code{char}.  If \code{NA}, the default, the number of characters
is that which fits into \code{getOption("width")}.
\item[\code{style}] the `style' of the bar -- see `Details'.
\item[\code{pb, con}] an object of class \code{"txtProgressBar"}.
\item[\code{title, label}] ignored, for compatibility with other progress bars.
\item[\code{...}] for consistency with the generic.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
\code{txtProgressBar} will display a progress bar on the \R{} console
via a text representation.

\code{setTxtProgessBar} will update the value.  Missing
(\code{\LinkA{NA}{NA}}) and out-of-range values of \code{value} will be
(silently) ignored.

The progress bar should be \code{close}d when finished with: this
outputs the final newline character.

\code{style = 1} and \code{style = 2} just shows a line of
\code{char}. They differ in that \code{style = 2} redraws the line
each time, which is useful if other code might be writing to the \R{}
console.  \code{style = 3} marks the end of the range by \code{|} and
gives a percentage to the right of the bar.
\end{Details}
%
\begin{Value}
For \code{txtProgressBar} an object of class \code{"txtProgressBar"}.

For \code{getTxtProgressBar} and \code{setTxtProgressBar}, a
length-one numeric vector giving the previous value (invisibly for
\code{setTxtProgressBar}).
\end{Value}
%
\begin{Note}\relax
Using \code{style} 2 or 3 or reducing the value with \code{style = 1}
uses \samp{\bsl{}r} to return to the left margin -- the interpretation of
carriage return is up to the terminal or console in which \R{} is running.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{tkProgressBar}{tkProgressBar}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
 # slow
testit <- function(x = sort(runif(20)), ...)
{
    pb <- txtProgressBar(...)
    for(i in c(0, x, 1)) {Sys.sleep(0.5); setTxtProgressBar(pb, i)}
    Sys.sleep(1)
    close(pb)
}
testit()
testit(runif(10))
testit(style=3)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{type.convert}{Type Conversion on Character Variables}{type.convert}
\keyword{manip}{type.convert}
%
\begin{Description}\relax
Convert a character vector to logical, integer, numeric, complex or
factor as appropriate.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
type.convert(x, na.strings = "NA", as.is = FALSE, dec = ".")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] a character vector.

\item[\code{na.strings}] a vector of strings which are to be interpreted as
\code{\LinkA{NA}{NA}} values.  Blank fields are also considered to be
missing values in logical, integer, numeric or complex vectors.

\item[\code{as.is}] logical.  See `Details'.

\item[\code{dec}] the character to be assumed for decimal points.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This is principally a helper function for \code{\LinkA{read.table}{read.table}}.
Given a character vector, it attempts to convert it to logical,
integer, numeric or complex, and failing that converts it to factor
unless \code{as.is = TRUE}.  The first type that can accept all the
non-missing values is chosen.

Vectors which are entirely missing values are converted to logical,
since \code{NA} is primarily logical.

Since this is a helper function, the caller should always pass an
appropriate value of \code{as.is}.
\end{Details}
%
\begin{Value}
A vector of the selected class, or a factor.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{read.table}{read.table}}
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{untar}{Extract or List Tar Archives}{untar}
\keyword{file}{untar}
\keyword{utilities}{untar}
%
\begin{Description}\relax
Extract files from or list a tar archive.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
untar(tarfile, files = NULL, list = FALSE, exdir = ".",
      compressed = NA, extras = NULL, verbose = FALSE,
      tar = Sys.getenv("TAR"))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{tarfile}] The pathname of the tar file: tilde expansion (see
\code{\LinkA{path.expand}{path.expand}}) will be performed.  Alternatively, a
connection that can be used for binary reads.

\item[\code{files}] A character vector of recorded filepaths to be extracted:
the default is to extract all files.

\item[\code{list}] If \code{TRUE}, just list the files.  The equivalent of
\command{tar -tf}.  Otherwise extract the files (the equivalent of
\command{tar -xf}).

\item[\code{exdir}] The directory to extract files to (the equivalent of
\command{tar -C}).  It will be created if necessary.

\item[\code{compressed}] logical or character.  Values \code{"gzip"},
\code{"bzip2"} and \code{"xz"} select that form of compression (and
may be abbreviated to the first letter).  \code{TRUE} indicates gzip
compression, \code{FALSE} no known compression (but the
\command{tar} command may detect compression automagically), and
\code{NA} (the default) that the type is inferred from the file
header.

\item[\code{extras}] \code{NULL} or a character string: further command-line
flags such as \option{-p} to be passed to the \command{tar} program.

\item[\code{verbose}] logical: if true echo the command used.

\item[\code{tar}] character string: the path to the command to be used.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
This is either a wrapper for a \command{tar} command or for an
internal implementation written in \R{}.  The latter is used if
\code{tarfile} is a connection or if the argument \code{tar} is
\code{"internal"} or \code{""} (except on Windows, when
\command{tar.exe} is tried first).

What options are supported will depend on the \command{tar} used.
Modern GNU flavours of \command{tar} will support compressed archives,
and since 1.15 are able to detect the type of compression
automatically: version 1.20 added support for \command{lzma} and
version 1.22 for \command{xz} compression using LZMA2.  For other
flavours of \command{tar}, environment variable \env{R\_GZIPCMD} gives
the command to decompress \command{gzip} and \command{compress} files,
and \command{R\_BZIPCMD} for its files.  (There is a \command{bsdtar}
command from the \samp{libarchive} project used by Mac OS 10.6
(`Snow Leopard') which can also detect \command{gzip} and
\command{bzip2} compression automatically, as can the \command{tar}
from the `Heirloom Toolchest' project.)

Arguments \code{compressed}, \code{extras} and \code{verbose} are only
used when an external \command{tar} is used.

The internal implementation restores links (hard and soft) as symbolic
links on a Unix-alike, and as file copies on Windows (which works only
for files, not for directories).  Since it uses \code{\LinkA{gzfile}{gzfile}}
to read a file it can handle files compressed by any of the methods
that function can handle: at least \command{compress}, \command{gzip},
\command{bzip2} and \command{xz} compression, and some types of
\command{lzma} compression.  It does not guard against restoring
absolute file paths, as some \command{tar} implementations do.  It
will create the parent directories for directories or files in the
archive if necessary. It handles both the standard (USTAR/POSIX) and
GNU ways of handling file paths of more than 100 bytes.

The standards only support ASCII filenames (indeed, only alphanumeric
plus period, underscore and hyphen).  \code{untar} makes no attempt to map
filenames to those acceptable on the current system, and treats the
filenames in the archive as applicable without any re-encoding in the
current locale.
\end{Details}
%
\begin{Value}
If \code{list = TRUE}, a character vector of (relative or absolute)
paths of files contained in the tar archive.

Otherwise the return code from \code{\LinkA{system}{system}}, invisibly.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{tar}{tar}}, \code{\LinkA{unzip}{unzip}}.  
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{unzip}{Extract or List Zip Archives}{unzip}
\keyword{file}{unzip}
\keyword{utilities}{unzip}
%
\begin{Description}\relax
Extract files from or list a zip archive.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
unzip(zipfile, files = NULL, list = FALSE, overwrite = TRUE,
       junkpaths = FALSE, exdir = ".")
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{zipfile}] The pathname of the zip file: tilde expansion (see
\code{\LinkA{path.expand}{path.expand}}) will be performed.

\item[\code{files}] A character vector of recorded filepaths to be extracted:
the default is to extract all files.

\item[\code{list}] If \code{TRUE}, list the files and extract none.  The
equivalent of \command{unzip -l}.

\item[\code{overwrite}] If \code{TRUE}, overwrite existing files, otherwise
ignore such files.  The equivalent of \command{unzip -o}.

\item[\code{junkpaths}] If \code{TRUE}, use only the basename of the stored
filepath when extracting.  The equivalent of \command{unzip -j}.

\item[\code{exdir}] The directory to extract files to (the equivalent of
\code{unzip -d}).  It will be created if necessary.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
If \code{list = TRUE}, a data frame with columns \code{Name},
\code{Length} (the size of the uncompressed file) and \code{Date} (of
class \code{"\LinkA{POSIXct}{POSIXct}"}).

Otherwise, a character vector of the filepaths extracted to, invisibly.
\end{Value}
%
\begin{Source}\relax
The C code uses \code{zlib} and is in particular based on the
contributed \samp{minizip} application in the \code{zlib} sources
(from \url{zlib.net}) by Gilles Vollant.
\end{Source}
%
\begin{SeeAlso}\relax
\code{\LinkA{unz}{unz}} and \code{\LinkA{zip.file.extract}{zip.file.extract}} to read a single
component from a zip file.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{update.packages}{Download Packages from CRAN-like repositories}{update.packages}
\aliasA{available.packages}{update.packages}{available.packages}
\aliasA{contrib.url}{update.packages}{contrib.url}
\aliasA{download.packages}{update.packages}{download.packages}
\aliasA{install.packages}{update.packages}{install.packages}
\aliasA{new.packages}{update.packages}{new.packages}
\aliasA{old.packages}{update.packages}{old.packages}
\keyword{utilities}{update.packages}
%
\begin{Description}\relax
These functions can be used to automatically compare the version
numbers of installed packages with the newest available version on
the repositories and update outdated packages on the fly.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
update.packages(lib.loc = NULL, repos = getOption("repos"),
                contriburl = contrib.url(repos, type),
                method, instlib = NULL,
                ask = TRUE, available = NULL,
                oldPkgs = NULL, ..., checkBuilt = FALSE,
                type = getOption("pkgType"))

available.packages(contriburl = contrib.url(getOption("repos"), type),
                   method, fields = NULL, type = getOption("pkgType"),
                   filters = NULL)

old.packages(lib.loc = NULL, repos = getOption("repos"),
             contriburl = contrib.url(repos, type),
             instPkgs = installed.packages(lib.loc = lib.loc),
             method, available = NULL, checkBuilt = FALSE,
             type = getOption("pkgType"))

new.packages(lib.loc = NULL, repos = getOption("repos"),
             contriburl = contrib.url(repos, type),
             instPkgs = installed.packages(lib.loc = lib.loc),
             method, available = NULL, ask = FALSE, ...,
             type = getOption("pkgType"))

download.packages(pkgs, destdir, available = NULL,
                  repos = getOption("repos"),
                  contriburl = contrib.url(repos, type),
                  method, type = getOption("pkgType"), ...)

install.packages(pkgs, lib, repos = getOption("repos"),
                 contriburl = contrib.url(repos, type),
                 method, available = NULL, destdir = NULL,
                 dependencies = NA, type = getOption("pkgType"),
                 configure.args = getOption("configure.args"),
                 configure.vars = getOption("configure.vars"),
                 clean = FALSE, Ncpus = getOption("Ncpus"), ...)

contrib.url(repos, type = getOption("pkgType"))
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{lib.loc}] character vector describing the location of R
library trees to search through (and update packages therein), or
\code{NULL} for all known trees (see \code{\LinkA{.libPaths}{.libPaths}}).
\item[\code{repos}] character vector, the base URL(s) of the repositories
to use, i.e., the URL of the CRAN master such as
\code{"http://cran.r-project.org"} or its Statlib mirror,
\code{"http://lib.stat.cmu.edu/R/CRAN"}.
Can be \code{NULL} to install from local files
(\file{.tar.gz} for source packages).

\item[\code{contriburl}] URL(s) of the contrib sections of the
repositories.  Use this argument only if your repository mirror is
incomplete, e.g., because you burned only the \file{contrib} section on a
CD.  Overrides argument \code{repos}.
As \code{repos}, can also be \code{NULL} to install from local files.

\item[\code{method}] Download method, see \code{\LinkA{download.file}{download.file}}.
\item[\code{pkgs}] character vector of the short names of packages/bundles whose
current versions should be downloaded from the repositories.
If \code{repos = NULL}, a character vector of file paths of
\file{.tar.gz} files.  These can be source archives or binary
package/bundle archive files (as created by
\command{R CMD build --binary}).
Tilde-expansion will be done on the file paths.
If this is a zero-length character vector, a listbox of available
packages (including those contained in bundles) is presented where
possible.

\item[\code{destdir}] directory where downloaded packages are stored.
\item[\code{available}] an object listing packages available at the repositories
as returned by \code{available.packages}.
\item[\code{lib}] character vector giving the library directories where to
install the packages.  Recycled as needed.  If missing, defaults to
\code{\LinkA{.libPaths}{.libPaths}()[1]}.
\item[\code{ask}] logical indicating whether to ask user before packages
are actually downloaded and installed, or the character string
\code{"graphics"}, which brings up a widget to allow the user to
(de-)select from the list of packages which could be updated.  The
latter only works on systems with a GUI version of
\code{\LinkA{select.list}{select.list}}, and is otherwise equivalent to \code{ask
      = TRUE}.

\item[\code{checkBuilt}] If \code{TRUE}, a package built under an earlier
minor version of \R{} is considered to be `old'.
\item[\code{instlib}] character string giving the library directory where to
install the packages.
\item[\code{dependencies}] logical indicating to also install uninstalled
packages on which these packages depend/suggest/import
(and so on recursively).  Not used if \code{repos = NULL}.
Can also be a character vector, a subset of
\code{c("Depends", "Imports", "LinkingTo", "Suggests", "Enhances")}.

Only supported if \code{lib} is of length one (or missing),
so it is unambiguous where to install the dependent packages.

The default, \code{NA}, means
\code{c("Depends", "Imports", "LinkingTo")}
if \code{lib} is unambiguous, and \code{FALSE} otherwise.

\item[\code{configure.args}] (not Windows) a character vector or a named list.
If a character vector with no names is supplied, the elements are
concatenated into a single string (separated by a space) and used
as the value for the \option{--configure-args}
flag in the call to \command{R CMD INSTALL}.
If the character vector has names these are assumed to identify
values for \option{--configure-args} for individual packages.
This allows one to specify settings for an entire collection of packages
which will be used if any of those packages are to be installed.
(These settings can therefore be reused and act as default settings.)

A named list can be used also to the same effect, and that
allows multi-element character strings for each package
which are concatenated to a single string to be used as the
value for \option{--configure-args}.

\item[\code{configure.vars}] (not Windows) similar, for
\option{--configure-vars}, which is used to set environment variables
for the \command{configure} run.
\item[\code{oldPkgs}] 
if specified as non-NULL, \code{update.packages()} only considers
these packages for updating. 

\item[\code{instPkgs}] 
by default all installed packages,
\code{\LinkA{installed.packages}{installed.packages}(lib.loc=lib.loc)}.  A subset can be 
specified; currently this must be in the same (character matrix)
format as returned by \code{installed.packages()}.

\item[\code{...}] 
(for \code{update.packages}).  Arguments such as \code{destdir},
\code{installWithVers} and \code{dependencies} to be passed to
\code{install.packages}.

(for \code{new.packages}).  Arguments such as \code{destdir}
and \code{dependencies} to be passed to \code{install.packages}.

(for \code{install.packages} and \code{download.packages}) arguments
to be passed to \code{\LinkA{download.file}{download.file}}.

\item[\code{type}] character, indicating the type of package to download and
install.

Possible values except on Windowsare \code{"source"} (the default
except under the CRAN Mac OS X build), \code{"mac.binary"} and
\code{"win.binary"} (which can be downloaded but not installed).

Possible values on Windows are \code{"win.binary"} (the default) and
\code{"source"} (for which suitable tools may need to be installed:
see the `Details').  Value \code{"mac.binary"} can be used to
explore and download Mac OS X binaries.

\item[\code{clean}] a logical value indicating whether to specify
to add the \option{--clean} flag to the call to
\command{R CMD INSTALL}.
This is sometimes used to perform additional operations at the end
of the package installation in addition to removing intermediate files.

\item[\code{Ncpus}] The number of parallel processes to use for a parallel
install of source packages.  Values greater than one are supported
only if GNU \command{make} is in use (more precisely, if
\command{make -j \var{Ncpus}} works).  Defaults to \code{1} is the
option is unset.

\item[\code{fields}] a character vector giving the fields to extract from
the \code{PACKAGES} file(s) in addition to the default ones, or
\code{NULL} (default).  Unavailable fields result in \code{NA}
values.

\item[\code{filters}] a character vector or list specifying the filter
operations to be performed on the packages found in the
repositories, or \code{NULL} (default).  Specified filters can be
one of the strings \code{"R\_version"}, \code{"OS\_type"},
\code{"duplicates"}, or \code{"license/FOSS"}, indicating built-in
filters which retain only packages whose version and OS type
requirements are met by the running version of \R{}, the latest
versions of packages, and packages for which installation can
proceed solely based on packages which can be verified as Free or
Open Source Software (FOSS, e.g.,
\url{http://en.wikipedia.org/wiki/FOSS}) employing the available
license specifications, respectively.  Filters can also be
user-defined functions which subscript the rows of objects returned
by \code{available.packages}.  The default corresponds to the
specification \code{c("R\_version", "OS\_type", "duplicates")}, and
may be changed by setting option \code{available\_packages\_filters}
to something non-\code{NULL}.  If the filters specification used is
a list with an \code{add = TRUE} element, the other elements are
used for filtering in addition to the default filters.

\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
All of these functions work with the names of a package or bundle (and
not the component packages of a bundle, except for
\code{install.packages} if the repository provides the necessary
information).

\code{available.packages} returns a matrix of details corresponding to
packages/bundles currently available at one or more repositories. The
current list of packages is downloaded over the internet (or copied
from a local mirror).  By default, it returns only packages whose
version and OS type requirements are met by the running version of \R{},
and only information on the latest versions or packages.

\code{old.packages} compares the information from
\code{available.packages} with that from \code{instPkgs} (computed by
\code{\LinkA{installed.packages}{installed.packages}} by default) and reports installed
packages/bundles that have newer versions on the repositories or, if
\code{checkBuilt = TRUE}, that were built under an earlier minor
version of \R{} (for example built under 2.8.x when running \R{} 2.9.0).

\code{new.packages} does the same comparison but reports uninstalled
packages/bundles that are available at the repositories.  It will
give warnings about incompletely installed bundles (provided the
information is available) and bundles whose contents has changed.
If \code{ask != FALSE} it asks which packages should be installed
in the first element of \code{lib.loc}.

\code{download.packages} takes a list of package/bundle names and a
destination directory, downloads the newest versions and saves them in
\code{destdir}.  If the list of available packages is not given as
argument, it is obtained from repositories.  If a repository is local,
i.e. the URL starts with \code{"file:"}, then the packages are not
downloaded but used directly.  Both \code{"file:"} and
\code{"file:///"} are allowed as prefixes to a file path.  Use the
latter only for URLs: see \code{\LinkA{url}{url}} for their interpretation.
(Other forms of \samp{file://} URLs are not supported.)

The main function of the set is \code{update.packages}.  First a list
of all packages/bundles found in \code{lib.loc} is created and
compared with those available at the repositories.  If
\code{ask = TRUE} (the default) packages/bundles with a newer version
are reported and for each one the user can specify if it should be
updated.
If so, the package sources are downloaded from the repositories and
installed in the respective library path (or \code{instlib}
if specified) using the \R{} \code{\LinkA{INSTALL}{INSTALL}} mechanism.

\code{install.packages} can be used to install new
packages/bundles. It takes a vector of names and a destination
library, downloads the packages from the repositories and installs
them.  (If the library is omitted it defaults to the first directory
in \code{.libPaths()}, with a warning if there is more than one.)  If
\code{lib} is omitted or is of length one and is not a (group)
writeable directory, the code offers to create a personal library tree
(the first element of \code{Sys.getenv("R\_LIBS\_USER")}) and install
there.

If a repository is used (rather than local \file{.tar.gz} files),
an attempt is made to install the packages in an order that respects
their dependencies.  This does assume that all the entries in
\code{lib} are on the default library path for installs (set by
\env{R\_LIBS}). 

\code{contrib.url} adds the appropriate type-specific path within a
repository to each URL in \code{repos}.

For \code{install.packages}, \code{destdir}
is the directory to which packages will be downloaded.  If it is
\code{NULL} (the default) a directory \code{downloaded\_packages} of
the session temporary directory will be used (and the files will be
deleted at the end of the session).

If \code{repos} or \code{contriburl} is a vector of length greater than
one, the newest version of the package compatible with this version of \R{}
is fetched from the first repository on the list within which it is found.
\end{Details}
%
\begin{Value}
For \code{available.packages}, a matrix with one row per
package/bundle, row names the package names and 
column names
\code{"Package"}, \code{"Version"}, \code{"Priority"},
\code{"Bundle"}, \code{"Contains"},
\code{"Depends"}, \code{"Imports"}, \code{"LinkingTo"},
\code{"Suggests"}, \code{"Enhances"},
\code{"OS\_type"}, \code{"License"},
\code{"File"} and \code{"Repository"}.
Additional columns can be specified using the \code{fields} argument.

For \code{old.packages}, \code{NULL} or a matrix with one row per
package/bundle, row names the package names and column names
\code{"Package"}, \code{"LibPath"}, \code{"Installed"} (the version),
\code{"Built"} (the version built under), \code{"ReposVer"} and
\code{"Repository"}.

For \code{new.packages} a character vector of package/bundle names,
\emph{after} any have been installed.

For \code{download.packages}, a two-column matrix of names and
destination file names, for those packages/bundles successfully
downloaded.  If packages are not available or there is a problem with
the download, suitable warnings are given.

\code{install.packages} and \code{update.packages} have no return value.
\end{Value}
%
\begin{Section}{Warning}
Take care when using \code{dependencies} with \code{update.packages},
for it is unclear where new dependencies should be installed.  The
current implementation will only allow it if all the packages to be
updated are in a single library, when that library will be used.

You are advised to run \code{update.packages} before
\code{install.packages} to ensure that any installed dependencies have
their latest versions.
\end{Section}
%
\begin{Note}\relax
Some binary distributions of \R{} have \code{INSTALL} in a separate
bundle, e.g. an \code{R-devel} RPM.  \code{install.packages} will
give an error if called with \code{type = "source"} on such a system.
\end{Note}
%
\begin{SeeAlso}\relax
\code{\LinkA{installed.packages}{installed.packages}}, \code{\LinkA{remove.packages}{remove.packages}}

See \code{\LinkA{download.file}{download.file}} for how to handle proxies and
other options to monitor file transfers.

\code{\LinkA{INSTALL}{INSTALL}}, \code{\LinkA{REMOVE}{REMOVE}},
\code{\LinkA{library}{library}}, \code{\LinkA{.packages}{.packages}}, \code{\LinkA{read.dcf}{read.dcf}}

The `R Installation and Administration' manual for how to 
set up a repository.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}

## Not run: 
install.packages(
    c("XML_0.99-5.tar.gz",
      "../../Interfaces/Perl/RSPerl_0.8-0.tar.gz"),
    repos = NULL,
    configure.args = c(XML = '--with-xml-config=xml-config',
                       RSPerl = "--with-modules='IO Fcntl'"))

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{url.show}{Display a text URL}{url.show}
\keyword{file}{url.show}
\keyword{misc}{url.show}
%
\begin{Description}\relax
Extension of \code{\LinkA{file.show}{file.show}} to display text files from a remote 
server.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
url.show(url, title = url, file = tempfile(),
         delete.file = TRUE, method, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{url}] The URL to read from.
\item[\code{title}] Title for the browser.
\item[\code{file}] File to copy to.
\item[\code{delete.file}] Delete the file afterwards?
\item[\code{method}] File transfer method: see \code{\LinkA{download.file}{download.file}}
\item[\code{...}] Arguments to pass to \code{\LinkA{file.show}{file.show}}.
\end{ldescription}
\end{Arguments}
%
\begin{SeeAlso}\relax
\code{\LinkA{url}{url}}, \code{\LinkA{file.show}{file.show}}, \code{\LinkA{download.file}{download.file}}
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: url.show("http://lib.stat.cmu.edu/datasets/csb/ch3a.txt")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{URLencode}{Encode or Decode a (partial) URL}{URLencode}
\aliasA{URLdecode}{URLencode}{URLdecode}
\keyword{utilities}{URLencode}
%
\begin{Description}\relax
Functions to encode or decode characters in URLs.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
URLencode(URL, reserved = FALSE)
URLdecode(URL)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{URL}] A character string.
\item[\code{reserved}] should reserved characters be encoded?  See
`Details'.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Characters in a URL other than the English alphanumeric characters and
\samp{\$ - \_ . + ! * ' ( ) ,} should be encoded as \code{\%}
plus a two-digit hexadecimal representation, and any single-byte
character can be so encoded. (Multi-byte characters are encoded as
byte-by-byte.)

In addition, \samp{; / ? : @ = \&} are reserved characters, and should
be encoded unless used in their reserved sense, which is scheme
specific.  The default in \code{URLencode} is to leave them alone, which
is appropriate for \samp{file://} URLs, but probably not for
\samp{http://} ones.
\end{Details}
%
\begin{Value}
A character string.
\end{Value}
%
\begin{References}\relax
RFC1738, \url{http://www.rfc-editor.org/rfc/rfc1738.txt}
\end{References}
%
\begin{Examples}
\begin{ExampleCode}
(y <- URLencode("a url with spaces and / and @"))
URLdecode(y)
(y <- URLencode("a url with spaces and / and @", reserved=TRUE))
URLdecode(y)
URLdecode("ab%20cd")
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{utils-deprecated}{Deprecated Functions in Package utils}{utils.Rdash.deprecated}
\aliasA{CRAN.packages}{utils-deprecated}{CRAN.packages}
\keyword{misc}{utils-deprecated}
%
\begin{Description}\relax
These functions are provided for compatibility with older versions of
\R{} only, and may be defunct as soon as of the next release.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
CRAN.packages(CRAN = getOption("repos"), method,
              contriburl = contrib.url(CRAN))

\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{CRAN}] character, an earlier way to specify a repository.
\item[\code{method}] Download method, see \code{\LinkA{download.file}{download.file}}.
\item[\code{contriburl}] URL(s) of the contrib section of the
repositories. Use this argument only if your CRAN mirror is
incomplete, e.g., because you burned only the \file{contrib} section on a
CD.  Overrides argument \code{repos}.
 
\end{ldescription}
\end{Arguments}
%
\begin{SeeAlso}\relax
\code{\LinkA{Deprecated}{Deprecated}}, \code{\LinkA{Defunct}{Defunct}}
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{View}{Invoke a Data Viewer}{View}
\keyword{utilities}{View}
%
\begin{Description}\relax
Invoke a spreadsheet-style data viewer on a matrix-like \R{} object.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
View(x, title)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] an \R{} object which can be coerced to a data frame with
non-zero numbers of rows and columns.
\item[\code{title}] title for viewer window.  Defaults to name of \code{x}
prefixed by \code{Data:}.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Object \code{x} is coerced (if possible) to a data frame, and all
non-numeric columns are then coerced to character.   The object is
then viewed in a spreadsheet-like data viewer, a read-only version of
\code{\LinkA{data.entry}{data.entry}}.

If there are row names on the data frame that are not \code{1:nrow},
they are displayed in a separate first column called \code{row.names}.

Objects with zero columns or zero rows are not accepted.

The array of cells can be navigated by the cursor keys and Home, End,
Page Up and Page Down (where supported by X11) as well as Enter
and Tab.
\end{Details}
%
\begin{Value}
Invisible \code{NULL}.  The functions puts up a window and returns
immediately: the window can be closed via its controls or menus.
\end{Value}
%
\begin{SeeAlso}\relax
\code{\LinkA{edit.data.frame}{edit.data.frame}},
\code{\LinkA{data.entry}{data.entry}}.
\end{SeeAlso}
\inputencoding{latin1}
\HeaderA{vignette}{View or List Vignettes}{vignette}
\aliasA{edit.vignette}{vignette}{edit.vignette}
\aliasA{print.vignette}{vignette}{print.vignette}
\aliasA{vignettes}{vignette}{vignettes}
\keyword{documentation}{vignette}
%
\begin{Description}\relax
View a specified vignette, or list the available ones.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
vignette(topic, package = NULL, lib.loc = NULL, all = TRUE)

## S3 method for class 'vignette':
print(x, ...)
## S3 method for class 'vignette':
edit(name, ...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{topic}] a character string giving the (base) name of the vignette
to view. If omitted, all vignettes from all installed packages are listed.
\item[\code{package}] a character vector with the names of packages to
search through, or \code{NULL} in which "all" packages (as defined
by argument \code{all}) are searched.
\item[\code{lib.loc}] a character vector of directory names of \R{} libraries,
or \code{NULL}.  The default value of \code{NULL} corresponds to all
libraries currently known.
\item[\code{all}] logical; if \code{TRUE} search
all available packages in the library trees specified by \code{lib.loc}, 
and if \code{FALSE}, search only attached packages.
\item[\code{x, name}] Object of class \code{vignette}.
\item[\code{...}] Ignored by the \code{print} method, passed on to
\code{\LinkA{file.edit}{file.edit}} by the \code{edit} method.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Function \code{vignette} returns an object of the same class, the
print method opens a viewer for it.
Currently, only PDF versions of vignettes can be viewed.
The program specified by the \code{pdfviewer} option is used for this.
If several vignettes have PDF versions with base name identical to
\code{topic}, the first one found is used.

If no topics are given, all available vignettes are listed.  The
corresponding information is returned in an object of class
\code{"packageIQR"}.

The \code{edit} method
extracts the \R{} code from the vignette to a temporary file and
opens the file in an editor (see \code{\LinkA{edit}{edit}}). This makes it
very easy to execute the commands line by line, modify them in any way
you want to help you test variants, etc.. An alternative way of
extracting the \R{} code from the vignette is to run
\code{\LinkA{Stangle}{Stangle}} on the source code of the vignette,
see the examples below.
\end{Details}
%
\begin{Examples}
\begin{ExampleCode}
## List vignettes from all *attached* packages
vignette(all = FALSE)

## List vignettes from all *installed* packages (can take a long time!):
vignette(all = TRUE)

## Not run: 
## Open the grid intro vignette
vignette("grid")

## The same
v1 <- vignette("grid")
print(v1)

## Now let us have a closer look at the code
edit(v1)

## An alternative way of extracting the code,
## R file is written to current working directory
Stangle(v1$file)

## A package can have more than one vignette (package grid has several):
vignette(package="grid")
vignette("rotated")
## The same, but without searching for it:
vignette("rotated", package="grid")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{write.table}{Data Output}{write.table}
\aliasA{write.csv}{write.table}{write.csv}
\aliasA{write.csv2}{write.table}{write.csv2}
\keyword{print}{write.table}
\keyword{file}{write.table}
%
\begin{Description}\relax
\code{write.table} prints its required argument \code{x} (after
converting it to a data frame if it is not one nor a matrix) to
a file or connection.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
write.table(x, file = "", append = FALSE, quote = TRUE, sep = " ",
            eol = "\n", na = "NA", dec = ".", row.names = TRUE,
            col.names = TRUE, qmethod = c("escape", "double"))

write.csv(...)
write.csv2(...)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{x}] the object to be written, preferably a matrix or data frame.
If not, it is attempted to coerce \code{x} to a data frame.
\item[\code{file}] either a character string naming a file or a connection
open for writing.  \code{""} indicates output to the console.
\item[\code{append}] logical. Only relevant if \code{file} is a character
string.  If \code{TRUE}, the output is appended to the
file.  If \code{FALSE}, any existing file of the name is destroyed.
\item[\code{quote}] a logical value (\code{TRUE} or \code{FALSE}) or a
numeric vector.  If \code{TRUE}, any character or factor columns
will be surrounded by double quotes.  If a numeric vector, its
elements are taken as the indices of columns to quote.  In both
cases, row and column names are quoted if they are written.  If
\code{FALSE}, nothing is quoted.
\item[\code{sep}] the field separator string.  Values within each row of
\code{x} are separated by this string.
\item[\code{eol}] the character(s) to print at the end of each line (row).
For example, \code{eol="\bsl{}r\bsl{}n"} will produce Windows' line endings on
a Unix-alike OS, and \code{eol="\bsl{}r"} will produce files as expected by
Mac OS Excel 2004.
\item[\code{na}] the string to use for missing values in the data.
\item[\code{dec}] the string to use for decimal points in numeric or complex
columns: must be a single character.
\item[\code{row.names}] either a logical value indicating whether the row
names of \code{x} are to be written along with \code{x}, or a
character vector of row names to be written.
\item[\code{col.names}] either a logical value indicating whether the column
names of \code{x} are to be written along with \code{x}, or a
character vector of column names to be written.  See the section on
`CSV files' for the meaning of \code{col.names = NA}.
\item[\code{qmethod}] a character string specifying how to deal with embedded
double quote characters when quoting strings.  Must be one of
\code{"escape"} (default), in which case the quote character is
escaped in C style by a backslash, or \code{"double"}, in which case
it is doubled.  You can specify just the initial letter.

\item[\code{...}] arguments to \code{write.table}: \code{col.names},
\code{sep}, \code{dec} and \code{qmethod} cannot be altered.

\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
If the table has no columns the rownames will be written only if
\code{row.names=TRUE}, and \emph{vice versa}.

Real and complex numbers are written to the maximal possible precision.

If a data frame has matrix-like columns these will be converted to
multiple columns in the result (\emph{via} \code{\LinkA{as.matrix}{as.matrix}})
and so a character \code{col.names} or a numeric \code{quote} should
refer to the columns in the result, not the input.  Such matrix-like
columns are unquoted by default.

Any columns in a data frame which are lists or have a class
(e.g. dates) will be converted by the appropriate \code{as.character}
method: such columns are unquoted by default.  On the other hand,
any class information for a matrix is discarded and non-atomic
(e.g. list) matrices are coerced to character.

Only columns which have been converted to character will be quoted if
specified by \code{quote}.

The \code{dec} argument only applies to columns that are not subject
to conversion to character because they have a class or are part of a
matrix-like column (or matrix), in particular to columns protected by
\code{\LinkA{I}{I}()}.  Use \code{\LinkA{options}{options}("OutDec")} to control
such conversions.

In almost all cases the conversion of numeric quantities is governed
by the option \code{"scipen"} (see \code{\LinkA{options}{options}}), but with
the internal equivalent of \code{digits=15}.  For finer control, use
\code{\LinkA{format}{format}} to make a character matrix/data frame, and call
\code{write.table} on that.

These functions check for a user interrupt every 1000 lines of output.

If \code{file} is not open for writing, an attempt is made to open it
and then close it after use.
\end{Details}
%
\begin{Section}{CSV files}
By default there is no column name for a column of row names.  If
\code{col.names = NA} and \code{row.names = TRUE} a blank column name
is added, which is the convention used for CSV files to be read by
spreadsheets.

\code{write.csv} and \code{write.csv2} provide convenience wrappers
for writing CSV files.  They set \code{sep}, \code{dec} and
\code{qmethod}, and \code{col.names} to \code{NA} if \code{row.names =
    TRUE} and \code{TRUE} otherwise.

\code{write.csv} uses \code{"."} for the decimal point and a comma for
the separator.

\code{write.csv2} uses a comma for the decimal point and a semicolon for
the separator, the Excel convention for CSV files in some Western
European locales.

These wrappers are deliberately inflexible: they are designed to
ensure that the correct conventions are used to write a valid file.
Attempts to change \code{col.names}, \code{sep}, \code{dec} or
\code{qmethod} are ignored, with a warning.
\end{Section}
%
\begin{Note}\relax
\code{write.table} can be slow for data frames with large numbers
(hundreds or more) of columns: this is inevitable as each column could
be of a different class and so must be handled separately.  If they
are all of the same class, consider using a matrix instead.
\end{Note}
%
\begin{SeeAlso}\relax
The `R Data Import/Export' manual.

\code{\LinkA{read.table}{read.table}}, \code{\LinkA{write}{write}}.

\code{\LinkA{write.matrix}{write.matrix}} in package \pkg{MASS}.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
## Not run: 
## To write a CSV file for input to Excel one might use
x <- data.frame(a = I("a \" quote"), b = pi)
write.table(x, file = "foo.csv", sep = ",", col.names = NA,
            qmethod = "double")
## and to read this file back into R one needs
read.table("foo.csv", header = TRUE, sep = ",", row.names = 1)
## NB: you do need to specify a separator if qmethod = "double".

### Alternatively
write.csv(x, file = "foo.csv")
read.csv("foo.csv", row.names = 1)
## or without row names
write.csv(x, file = "foo.csv", row.names = FALSE)
read.csv("foo.csv")

## End(Not run)
\end{ExampleCode}
\end{Examples}
\inputencoding{latin1}
\HeaderA{zip.file.extract}{Extract File from a Zip Archive}{zip.file.extract}
\keyword{file}{zip.file.extract}
%
\begin{Description}\relax
This will extract the file named \code{file} from the zip archive,
if possible, and write it in a temporary location.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
zip.file.extract(file, zipname = "R.zip",
                 unzip = getOption("unzip"), dir = tempdir())
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{file}] file name. (If a path is given, see `Note'.)
\item[\code{zipname}] The file name (not path) of a \code{zip} archive,
including the \code{".zip"} extension if required.
\item[\code{unzip}] character string: the method to be used, an empty string
indicates \code{"internal"}.
\item[\code{dir}] directory (``folder'') name into which the extraction
happens.  Must be writable to the caller.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
All platforms support an \code{"internal"} unzip: this is the default
under Windows and the fall-back under Unix if no \code{unzip} program
was found during configuration and \env{R\_UNZIPCMD} is not set.

The file will be extracted if it is in the archive and any required
\code{unzip} utility is available.  It will be extracted to the
directory given by \code{dir}, overwriting any existing file of
that name.
\end{Details}
%
\begin{Value}
The name of the original or extracted file.  Success is indicated by
returning a different name.
\end{Value}
%
\begin{Note}\relax
The \code{"internal"} method is very simple, and will not set file dates.

This is a helper function for \code{\LinkA{help}{help}},
\code{\LinkA{example}{example}} and \code{\LinkA{data}{data}}.  As such, it handles
file paths in an unusual way.  Any path component of \code{zipname} is
ignored, and the path to \code{file} is used only to determine the
directory within which to find \code{zipname}.
\end{Note}
%
\begin{Source}\relax
The C code uses \code{zlib} and is in particular based on the
contributed \samp{minizip} application in the \code{zlib} sources by
Gilles Vollant.
\end{Source}
%
\begin{SeeAlso}\relax
\code{\LinkA{unzip}{unzip}}
\end{SeeAlso}
\clearpage
